### 2014-04-04 stat246/hw1.tex
poop

\documentclass{article}
\usepackage{geometry}
\usepackage[namelimits,sumlimits]{amsmath}
\usepackage{amssymb,amsfonts}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage[cm]{fullpage}
\newcommand{\tab}{\hspace*{5em}}
\newcommand{\conj}{\overline}
\newcommand{\dd}{\partial}
\newcommand{\ep}{\epsilon}
\newcommand{\openm}{\begin{pmatrix}}
\newcommand{\closem}{\end{pmatrix}}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\Null}{null}
\newcommand{\nc}{\newcommand}
\newcommand{\rn}{\mathbb{R}}
\nc{\cn}{\mathbb{C}}
\nc{\ssn}[1]{\subsubsection*{#1}}
\nc{\inner}[2]{\langle #1,#2\rangle}
\nc{\h}[1]{\widehat{#1}}
\nc{\tl}[1]{\widetilde{#1}}
\nc{\norm}[1]{\left\|{#1}\right\|}
\begin{document}

Name: Hall Liu

Date: \today 
\vspace{1.5cm}
\subsection*{1}
Over such a thin shell of thickness $\ep$, we can assume the density function to be roughly constant. Then, evaluating the density function at any point $x$ with norm $r$ gives
\[\frac{1}{(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-1}{2}x^T\sigma^{-2}Ix\right)=\frac{1}{(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-r^2}{2\sigma^2}\right)\]
The hypervolume of the surface of the sphere is $S_dr^{d-1}$ (since it's a $d-1$-dimensional set), so the volume of the shell is approximately $\ep S_dr^{d-1}$. The integral of the density over the shell is approximately the volume of the shell times the value of the density, or 
\[\ep\frac{S_dr^{d-1}}{(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-r^2}{2\sigma^2}\right)\]
To find the maximum of $f$, differentiate it by $r$ to obtain
\[\frac{(d-1)S_dr^{d-2}}{(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-r^2}{2\sigma^2}\right)-\frac{S_dr^d}{\sigma^2(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-r^2}{2\sigma^2}\right)\]
Setting this equal to $0$ and canceling gives
\[0=(d-1)-\frac{r^2}{\sigma^2}\implies r=\pm\sigma\sqrt{d-1}\]
Now, to check that this is a maximum, taking the second derivative gives us a common, positive multiplier times the following:
\[(d-1)(d-2)-\frac{r^2}{\sigma^2}(2d-1)+\frac{r^4}{\sigma^4}\]
Plugging in the previously obtained value for the stationary point, we have
\[d^2-3d+2-(d-1)(2d-1)+(d-1)^2=-2d+2\]
which becomes negative for large $d$. Thus, the point is a maximum, and we can estimate it by $\sigma\sqrt{d}$ for large $d$.
%TODO: defn of p?
\subsection{2}
By the AM-GM inequality, we have $\frac{a+b}{2}\leq\sqrt{ab}$. Since the minimum of $a$ and $b$ is bounded above by the arithmetic mean, we thus also have $\min(a,b)\leq\sqrt{ab}$.

Now, we have that $P(\text{error})=P(\text{error}|Y=1)P(Y=1)+P(\text{error}|Y=2)P(Y=2)$. If we let $\Sigma_i$ be the set which the classifier classifies as $\h{Y}=i$, then the above is equal to 
\[\int_{\Sigma_2}P(x|Y=1)P(Y=1)dx+\int_{\Sigma_1}P(x|Y=2)P(Y=2)dx\]
Over $\Sigma_1$, we have that $P(x|Y=1)P(Y=1)\leq P(x|Y=2)P(Y=2)$ by the form of the Bayes classifier, and vice versa for $\Sigma_2$. Thus, the integrals above are bounded above by
\begin{align*}
&\frac{1}{2}\int_{\Sigma_2}P(x|Y=1)P(Y=1)+P(x|Y=2)P(Y=2)dx+ \frac{1}{2}\int_{\Sigma_1}P(x|Y=1)P(Y=1)+P(x|Y=2)P(Y=2)dx\\
    &=\frac{1}{2}\int_{\rn^n}P(x|Y=1)P(Y=1)+P(x|Y=2)P(Y=2)dx
\end{align*}
if we replace the integrands with the average of the two integrands. Now, using the AM-GM inequality, we have that this is bounded above by 
\begin{align*}
\int_{\rn^n}\sqrt{P(x|Y=1)P(Y=1)P(x|Y=2)P(Y=2)}dx&=\sqrt{P(Y=1)P(Y=2)}\int_{\rn^n}\sqrt{P(x|Y=1)P(x|Y=2)}dx\\
&\leq\frac{1}{2}\int_{\rn^n}\sqrt{P(x|Y=1)P(x|Y=2)}dx
\end{align*}
by another application of AM-GM to $\sqrt{P(Y=1)P(Y=2)}$, noting that the average of $P(Y=1)$ and $P(Y=2)$ is $\frac{1}{2}$.
\subsection*{3}
\ssn{a}
The decision boundary is the point at which $P(x|Y=1)P(Y=1)=P(x|Y=2)P(Y=2)$, or $\pi_1\exp\left((x-\mu_1)^2/(2\sigma^2)\right)=\pi_2\exp\left((x-\mu_2)^2/(2\sigma^2)\right)$. Solving for $x$, we have 
\[x^*=\frac{2\sigma\log\frac{\pi_1}{\pi_2}+\mu_2^2-\mu_1^2}{2(\mu_2-\mu_1)}\]
\ssn{b}
The error probability is
\[\int_{-\infty}^{x^*}P(x|Y=2)P(Y=2)dx+\int_{x^*}^\infty P(x|Y=1)P(Y=1)dx=\pi_2\Phi\left(\frac{x^*-\mu_2}{\sigma}\right)+\pi_1-\pi_1\Phi\left(\frac{x^*-\mu_1}{\sigma}\right)\]

Rewrite $x^*-\mu_2=\frac{2\sigma\log\frac{\pi_1}{\pi_2}}{2(\mu_2-\mu_1)}+\frac{\mu_2+\mu_1}{2}-\mu_2$. Then, dividing this by $\sigma$ produces a term constant in $\sigma$ and a term that goes to $-\infty$ as $\sigma\to0$. Thus, the $\pi_2$ term in the error vanishes. Similarly, $\frac{x^*-\mu_1}{\sigma}$ goes to $\infty$ as $\sigma\to0$, which means that the overall error term approaches $\pi_1-\pi_1=0$.
\ssn{c}
As $\pi_1\to0$, the log term in the decision boundary approaches $-\infty$, dragging the decision boundary with it, as all other things are fixed. In this case, always classifying things as class 2 would produce an error rate of $\pi_1$, which is known to be small.
\ssn{d}
{
\nc{\lot}{L_{1,2}}
\nc{\lto}{L_{2,1}}
Let $S_1$ be the set of points in class $1$ and $S_2$ the same for class 2. If we predict class 1, the expected loss is $E_X(\lto P(Y=2|X)$. If we predict class 2, then it is $E_X(\lot P(Y=1|X))$. If we minimize this pointwise, we should predict class 1 when $\lto P(Y=2|X)\leq\lot P(Y=1|X)$, and class 2 otherwise.

Rewriting the conditional probabilities, the prediction is then $\text{argmax}\left(\lto P(X|Y=2)P(Y=2), \lot P(X|Y=1)P(Y=1)\right)$. Setting these equal to find the boundary, we have
\[\lto \pi_2\exp\left((x-\mu_2)^2/(2\sigma^2)\right)=\lot\pi_1\exp\left((x-\mu_1)^2/(2\sigma^2)\right)\]
Solving for $x$, we obtain
\[x^*=\frac{2\sigma\log\frac{\lot\pi_1}{\lto\pi_2}+\mu_2^2-\mu_1^2}{2(\mu_2-\mu_1)}\]
Now, if we choose $\lot$ to be proportional to $\pi_1^{-1}$ and choose $\lto$ to be proportional to $\pi_2^{-1}$ (possibly with different constants), we will avoid the degeneracy problem.
}
\ssn{e}
Using these values, we have that $x^*=1$, so plugging into the expression for the error rate from (b) gives $0.5\Phi(-1)+0.5-0.5\Phi(1)=0.1587$
\end{document}

### 2014-04-05 stat246/hw1.tex
s1

@@
 \DeclareMathOperator{\cov}{cov}
+\DeclareMathOperator{\var}{var}
 \DeclareMathOperator{\rank}{rank}
@@
 Now, if we choose $\lot$ to be proportional to $\pi_1^{-1}$ and choose $\lto$ to be proportional to $\pi_2^{-1}$ (possibly with different constants), we will avoid the degeneracy problem.
-}
 \ssn{e}
 Using these values, we have that $x^*=1$, so plugging into the expression for the error rate from (b) gives $0.5\Phi(-1)+0.5-0.5\Phi(1)=0.1587$
+}
+\subsection*{4}
+The expected loss of some decision function $\h{H}$ is $E_{X, Y}(L(\h{H}(x), Y)$. Writing out the integral gives
+\[\int_{\rn^n}\sum_{k=1}^KL(\h{H}(x), k)p(x, k) dx=\int_{\rn^n}p(x)\sum_{k=1}^KL(\h{H}(x), k)P(k|x)dx\]
+
+%If we define $R_i$ as the subset of $\rn^n$ for which $\h{H}(x)=i$, we can separate the integral as follows
+%\[\sum_{i=1}^K\int_{R_i}p(x)\sum_{k=1}^KL(i, k)P(k|x)dx\]
+%If there is some subset $A$ of $R_i$ for which there exists a $j$ such that  for all $x\in A$, then moving $A$ from $R_i$ to $R_j$ will reduce the value of the expected loss by comparison
+
+Define $R_i$ as $\{x\in\rn^n|\h{H}(x)=i\}$.
+Then, suppose that there exists some subset of nonzero measure $A$ of $R_i$ such that $\sum_{k=1}^KL(i, k)P(k|x)\ge\sum_{k=1}^KL(j, k)P(k|x)$.
+Decompose the above integral as
+\[\int_{\rn^n-A}p(x)\sum_{k=1}^KL(\h{H}(x), k)P(k|x)dx+\int_Ap(x)\sum_{k=1}^KL(i, k)P(k|x)dx\]
+Then, changing the class of $A$ from $i$ to $j$ will result in a strict decrease in the expected loss, which means that the optimal decision rule must satisfy, for each $i$, $\sum_{k=1}^KL(i, k)P(k|x)\ge\sum_{k=1}^KL(j, k)P(k|x)$ for all $j\neq i$ and almost all $x\in R_i$.
+The rule given satisfies this condition, and since there are a finite number of classes, changing the rule on a set of measure zero from the region associated with each class will not result in a change in the expected loss.
+Thus, the rule given is optimal.
+\subsection*{5}
+\ssn{a}
+We have $\cov(X_2,X_2)=v_2$, so writing $X_2=\alpha X_1+Z$ gives
+\[\cov(\alpha X_1+Z, \alpha X_1+Z)=\cov(\alpha X_1, \alpha X_1)+0+\cov(Z, Z)=\alpha^2v_1+\var(Z)=v_2\]
+In addition, using $\cov(X_1, X_2)=a$ gives us
+\[a=\cov(X_1, \alpha X_1+Z)=\alpha v_1+0\implies \alpha=\frac{a}{v_1}\]
+Substituting into the first equation gives us $\var(Z)=v_2-\frac{a^2}{v_1}$
+
+Finally, since $E(X_1)=E(X_2)=0$, we must also have $E(Z)=0$.
+\ssn{b}
+We know that $Z$ is Gaussian because we can express $Z$ as the sum of two Gaussians, $X_2-\alpha X_1$.
+Thus, it is independent from $X_1$ because their covariance is zero.
+The variance was derived above.
+\ssn{c}
+We have $E(X_2|X_1=x)=E(\alpha x+Z)=\alpha x$ and $\var(X_2|X_1=x)=\var(\alpha x+Z)=\var(Z)$. 
+\ssn{d}
+Since the samples are all independent, the cross terms in the formula for the variance of a sum disappear and we get
+\[\var(\h{a})=\frac{1}{n^2}\sum_{i=1}^n\var(X_{i1}X_{i2})=\frac{1}{n}\var(X_1X_2)\]
+Then, we have $\var(X_1X_2)=E((X_1X_2)^2)-E(X_1X_2)^2=E((X_1X_2)^2)-a^2$, since the means are zero.
+
+To obtain the first term, note that it is a cross moment and can be obtained by applying $\frac{\partial^4}{\partial x_1^2\partial x_2^2}$ to the mgf of the distribution and evaluating at zero.
+The mgf itself is
+\[\exp\left(\frac{1}{2}\openm x_1&x_2\closem\openm v_1&a\\a&v_2\closem\openm x_1\\x_2\closem\right)=\exp\left(\frac{1}{2}(x_1^2v_1+2x_1x_2a+x_2^2v_2)\right)\]
+
+Differentiating this wrt $x_1$ twice and wrt $x_2$ once gives 
+\[ax_1v_1T+x_1v_1(x_1v_1+x_2a)(x_2v_2+x_1a)T+v_1(x_2v_2+x_1a)T\]
+where $T$ is the exponential term from above. Then, taking the derivatives and then evaluating gives $v_1v_2$, so the variance of $\h{a}$ is $\frac{1}{n}(v_1v_2-a^2)$.
+\subsection*{6}
 \end{document}

### 2014-04-05 stat246/hw1.tex
s1

@@
 which becomes negative for large $d$. Thus, the point is a maximum, and we can estimate it by $\sigma\sqrt{d}$ for large $d$.
-%TODO: defn of p?
+
+Plugging $r^*+\ep$ to $f$, we get 
+\[\frac{S_d(r^*+\ep)^{d-1}}{(2\pi\sigma^2)^{d/2}}\exp\left(-\frac{{r^*}^2+2r^*\ep+\ep^2}{2\sigma^2}\right)\]
+Ignoring the first-order term in $\ep$ inside the exponential, we're left with
+\[\frac{S_d(r^*+\ep)^{d-1}}{(2\pi\sigma^2)^{d/2}}\exp\left(-\frac{{r^*}^2}{2\sigma^2}\right)\exp\left(-\frac{\ep^2}{2\sigma^2}\right)\]
+As $\ep$ grows larger, the decay from the exponential term will dominate the at-most-order-$d-1$ contribution from $(r^*+\ep)^{d-1}$, so we can approximate this as 
+\[f(r^*)\exp\left(-\frac{\ep^2}{2\sigma^2}\right)\]
+Thus, the value of $f$ decays as the exponential of the square of the distance from the maximum point, which means that the mass of the density in shells away from that optimal radius is in fact very low.
 \subsection{2}

### 2014-04-05 stat246/hw1.tex
s1

@@
 \subsection*{6}
+\ssn{a}
+$\Sigma$ is positive definite because $A$ is nonsingular -- computing $x^T\Sigma x=x^TA^TAx=\|Ax\|_2^2$ will result in a positive value as long as $x$ is nonzero.
+The eigenvalues of $\Sigma$ are the squared eigenvalues of $A$, so the lowest eigenvalue of $\Sigma$ is $0.01$.
+\ssn{b}
+Sampling from a multivariate normal distribution with some specified covariance matrix can be done by first obtaining an iid sample of $n$ standard normal random variables, then transforming it via left-multiplication by $A^T$.
+Python code is attached.
 \end{document}

### 2014-04-05 stat246/hw1_files/p6.py

@@ -1,0 +1,29 @@
+#!/usr/bin/python2
+import numpy as np
+import scipy.stats as st
+
+'''
+Computes the e and m arrays as desired. e is returned
+as a d by d by 1000 array. A is the component of the 
+covariance matrix.
+'''
+def compute_results(d, A):
+    iters = 1000
+    n_samples = 500
+
+    e_data = np.zeros((iters, d, d))
+    m_data = np.zeros(iters)
+
+    for t in range(iters):
+        # Perform the random sample
+        samples = np.random.randn(d, n_samples)
+        # Transform it to have the correct distribution
+        samples = np.dot(A.T, samples)
+
+        # Compute the covariance estimation
+        cov_est = np.dot(samples, samples.T) / n_samples
+
+        # Compute the standardized errors
+        stderr_array = e_data[t, :, :]
+
+

### 2014-04-06 stat246/hw1.tex
s1 rev

@@
 
-To obtain the first term, note that it is a cross moment and can be obtained by applying $\frac{\partial^4}{\partial x_1^2\partial x_2^2}$ to the mgf of the distribution and evaluating at zero.
+%To obtain the first term, note that it is a cross moment and can be obtained by applying $\frac{\partial^4}{\partial x_1^2\partial x_2^2}$ to the mgf of the distribution and evaluating at zero.
 The mgf itself is
-\[\exp\left(\frac{1}{2}\openm x_1&x_2\closem\openm v_1&a\\a&v_2\closem\openm x_1\\x_2\closem\right)=\exp\left(\frac{1}{2}(x_1^2v_1+2x_1x_2a+x_2^2v_2)\right)\]
-
-Differentiating this wrt $x_1$ twice and wrt $x_2$ once gives 
-\[ax_1v_1T+x_1v_1(x_1v_1+x_2a)(x_2v_2+x_1a)T+v_1(x_2v_2+x_1a)T\]
-where $T$ is the exponential term from above. Then, taking the derivatives and then evaluating gives $v_1v_2$, so the variance of $\h{a}$ is $\frac{1}{n}(v_1v_2-a^2)$.
+%\[\exp\left(\frac{1}{2}\openm x_1&x_2\closem\openm v_1&a\\a&v_2\closem\openm x_1\\x_2\closem\right)=\exp\left(\frac{1}{2}(x_1^2v_1+2x_1x_2a+x_2^2v_2)\right)\]
+%
+%Differentiating this wrt $x_1$ twice and wrt $x_2$ once gives 
+%\[ax_1v_1T+x_1v_1(x_1v_1+x_2a)(x_2v_2+x_1a)T+v_1(x_2v_2+x_1a)T\]
+%where $T$ is the exponential term from above. Then, taking the derivatives and then evaluating gives $v_1v_2$, so the variance of $\h{a}$ is $\frac{1}{n}(v_1v_2-a^2)$.
+Write $X_1^2X_2^2=X_1^2(\alpha X_1+Z)^2$. Then, the expectation of this is
+\[\alpha^2E(X_1^4)+2\alpha E(Z)E(X_1^3)+E(X_1^2)E(Z^2)=\alpha^2E(X_1^4)+E(X_1^2)E(Z^2)\]
+By differentiating the mgf for $X_1$ four times and evaluating at zero, we obtain $E(X_1^4)=3v_1$. We have that $E(X_1^2)=v_1$ and $E(Z^2)=\var(Z)$, so the resulting expression for $E(X_1^2X_2^2)$ is
+\[3\frac{a^2}{v_1^2}v_1^2+v_1\left(v_2-\frac{a^2}{v_1}\right)=v_1v_2+2a^2\]
+which gives us $\var(\h{a})=\frac{1}{n}(v_1v_2+a^2)$
 \subsection*{6}

### 2014-04-06 stat246/hw1_files/helper_fns.pyx
p6 code

@@ -1,0 +1,13 @@
+import numpy as np
+cimport numpy as np
+
+def compute_stderrs(np.ndarray[double, ndim=2] cov, int n):
+    cdef int i, j
+    cdef int d = cov.shape[0]
+    cdef np.ndarray[double, ndim=2] cov_stderrs = np.zeros((d, d), dtype='float64')
+    for i in range(d):
+        for j in range(d):
+            cov_stderrs[i, j] = cov[i, i]*cov[j, j] + cov[i, j] * cov[i, j]
+    cov_stderrs /= n
+    return cov_stderrs
+

### 2014-04-06 stat246/hw1_files/p6.py

@@ -1,6 +1,9 @@
 #!/usr/bin/python2
 import numpy as np
 import scipy.stats as st
+import pyximport
+pyximport.install()
+from helper_fns import *
 
 '''
 Computes the e and m arrays as desired. e is returned
@@ -24,6 +27,8 @@ def compute_results(d, A):
         cov_est = np.dot(samples, samples.T) / n_samples
 
         # Compute the standardized errors
-        stderr_array = e_data[t, :, :]
+        stderrs = compute_stderrs(np.dot(A.T, A), n)
+        e_data[t, :, :] = (cov - cov_est) / stderrs
+        m_data[t] = np.min(np.linalg.eigvalsh(cov_est))
 
-
+    return (e_data, m_data)

### 2014-04-06 stat246/hw1_files/p6.py
O3 for cython

@@ -1,8 +1,7 @@
 #!/usr/bin/python2
 import numpy as np
 import scipy.stats as st
-import pyximport
-pyximport.install()
+import matplotlib.pyplot as plt
 from helper_fns import *
 
 '''
@@ -10,12 +9,12 @@ Computes the e and m arrays as desired. e is returned
 as a d by d by 1000 array. A is the component of the 
 covariance matrix.
 '''
-def compute_results(d, A):
+def compute_results(d, A, n_samples):
     iters = 1000
-    n_samples = 500
 
     e_data = np.zeros((iters, d, d))
     m_data = np.zeros(iters)
+    cov = np.dot(A.T, A)
 
     for t in range(iters):
         # Perform the random sample
@@ -27,8 +26,34 @@ def compute_results(d, A):
         cov_est = np.dot(samples, samples.T) / n_samples
 
         # Compute the standardized errors
-        stderrs = compute_stderrs(np.dot(A.T, A), n)
+        stderrs = compute_stderrs(cov, n_samples)
         e_data[t, :, :] = (cov - cov_est) / stderrs
         m_data[t] = np.min(np.linalg.eigvalsh(cov_est))
 
     return (e_data, m_data)
+
+def generate_results():
+    a100 = np.loadtxt('./a100.txt')
+    a10 = np.loadtxt('./a10.txt')
+
+    for ssize in [500, 5000]:
+        e10_data, m10_data = compute_results(10, a10, ssize)
+        e100_data, m100_data = compute_results(100, a100, ssize)
+        e10f = e10_data.flatten()
+        e100f = e100_data.flatten()
+
+        print('10 mean: {0}. 100 mean: {1}'.format(e10f.mean(), e100f.mean()))
+        print('10 std: {0}. 100 std: {1}'.format(e10f.std(), e100f.std()))
+
+        plt.figure()
+        hist10, bins10 = np.histogram(e10f, bins=50)
+        width10 = 0.6*(bins10[1] - bins10[0])
+        centers10 = (bins10[:-1] + bins10[1:]) / 2
+        plt.bar(centers10, hist10, align='center', width=width10)
+
+        plt.figure()
+        hist100, bins100 = np.histogram(e100f, bins=50)
+        width100 = 0.6*(bins100[1] - bins100[0])
+        centers100 = (bins100[:-1] + bins100[1:]) / 2
+        plt.bar(centers100, hist100, align='center', width=width100)
+        import ipdb;ipdb.set_trace()

### 2014-04-06 stat246/hw1_files/setup.py

@@ -1,0 +1,9 @@
+from distutils.core import setup
+from Cython.Distutils import build_ext, Extension
+import numpy as np
+
+setup(
+    cmdclass = {'build_ext': build_ext},
+    ext_modules = [Extension("helper_fns", ['helper_fns.pyx'], extra_compile_args=['-O3', '-std=gnu99'])],
+    include_dirs=[np.get_include()]
+)

### 2014-04-06 stat246/hw1_files/helper_fns.pyx
boundscheck

@@ -1,6 +1,8 @@
 import numpy as np
 cimport numpy as np
+cimport cython
 
+@cython.boundscheck(False)
 def compute_stderrs(np.ndarray[double, ndim=2] cov, int n):
     cdef int i, j
     cdef int d = cov.shape[0]

### 2014-04-06 stat246/hw1_files/p6.py
auto

@@ -50,10 +50,13 @@ def generate_results():
         width10 = 0.6*(bins10[1] - bins10[0])
         centers10 = (bins10[:-1] + bins10[1:]) / 2
         plt.bar(centers10, hist10, align='center', width=width10)
+        plt.savefig('e10-{0}.png'.format(ssize), bbox_inches='tight')
 
         plt.figure()
         hist100, bins100 = np.histogram(e100f, bins=50)
         width100 = 0.6*(bins100[1] - bins100[0])
         centers100 = (bins100[:-1] + bins100[1:]) / 2
         plt.bar(centers100, hist100, align='center', width=width100)
-        import ipdb;ipdb.set_trace()
+        plt.savefig('e100-{0}.png'.format(ssize), bbox_inches='tight')
+
+generate_results()

### 2014-04-06 cs250-hw/hw1.tex
c1

--- 2014-04-05 stat246/hw1.tex
@@
 \usepackage{graphicx}
+\usepackage{mathtools}
 \usepackage[cm]{fullpage}
@@
 \DeclareMathOperator{\cov}{cov}
-\DeclareMathOperator{\var}{var}
 \DeclareMathOperator{\rank}{rank}
@@
 \newcommand{\rn}{\mathbb{R}}
+\newcommand{\zn}{\mathbb{Z}}
 \nc{\cn}{\mathbb{C}}
@@
 \nc{\norm}[1]{\left\|{#1}\right\|}
+\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
+\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
 \begin{document}
-
 Name: Hall Liu
@@
 \vspace{1.5cm}
+
 \subsection*{1}
-Over such a thin shell of thickness $\ep$, we can assume the density function to be roughly constant. Then, evaluating the density function at any point $x$ with norm $r$ gives
-\[\frac{1}{(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-1}{2}x^T\sigma^{-2}Ix\right)=\frac{1}{(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-r^2}{2\sigma^2}\right)\]
-The hypervolume of the surface of the sphere is $S_dr^{d-1}$ (since it's a $d-1$-dimensional set), so the volume of the shell is approximately $\ep S_dr^{d-1}$.
-The integral of the density over the shell is approximately the volume of the shell times the value of the density, or
-\[\ep\frac{S_dr^{d-1}}{(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-r^2}{2\sigma^2}\right)\]
-To find the maximum of $f$, differentiate it by $r$ to obtain
-\[\frac{(d-1)S_dr^{d-2}}{(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-r^2}{2\sigma^2}\right)-\frac{S_dr^d}{\sigma^2(2\pi\sigma^2)^{d/2}}\exp\left(\frac{-r^2}{2\sigma^2}\right)\]
-Setting this equal to $0$ and canceling gives
-\[0=(d-1)-\frac{r^2}{\sigma^2}\implies r=\pm\sigma\sqrt{d-1}\]
-Now, to check that this is a maximum, taking the second derivative gives us a common, positive multiplier times the following:
-\[(d-1)(d-2)-\frac{r^2}{\sigma^2}(2d-1)+\frac{r^4}{\sigma^4}\]
-Plugging in the previously obtained value for the stationary point, we have
-\[d^2-3d+2-(d-1)(2d-1)+(d-1)^2=-2d+2\]
-which becomes negative for large $d$. Thus, the point is a maximum, and we can estimate it by $\sigma\sqrt{d}$ for large $d$.
 
-Plugging $r^*+\ep$ to $f$, we get 
-\[\frac{S_d(r^*+\ep)^{d-1}}{(2\pi\sigma^2)^{d/2}}\exp\left(-\frac{{r^*}^2+2r^*\ep+\ep^2}{2\sigma^2}\right)\]
-Ignoring the first-order term in $\ep$ inside the exponential, we're left with
-\[\frac{S_d(r^*+\ep)^{d-1}}{(2\pi\sigma^2)^{d/2}}\exp\left(-\frac{{r^*}^2}{2\sigma^2}\right)\exp\left(-\frac{\ep^2}{2\sigma^2}\right)\]
-As $\ep$ grows larger, the decay from the exponential term will dominate the at-most-order-$d-1$ contribution from $(r^*+\ep)^{d-1}$, so we can approximate this as 
-\[f(r^*)\exp\left(-\frac{\ep^2}{2\sigma^2}\right)\]
-Thus, the value of $f$ decays as the exponential of the square of the distance from the maximum point, which means that the mass of the density in shells away from that optimal radius is in fact very low.
-\subsection{2}
-By the AM-GM inequality, we have $\frac{a+b}{2}\leq\sqrt{ab}$. Since the minimum of $a$ and $b$ is bounded above by the arithmetic mean, we thus also have $\min(a,b)\leq\sqrt{ab}$.
+\subsection*{3}
+\ssn{b}
+I have a general policy against running unknown scripts on my computer, and since I know that I have a large part of the Python toolchain set up already, I decided to read the setup script (\verb|toolkit.sh|) and replicate the necessary parts.
 
-Now, we have that $P(\text{error})=P(\text{error}|Y=1)P(Y=1)+P(\text{error}|Y=2)P(Y=2)$. If we let $\Sigma_i$ be the set which the classifier classifies as $\h{Y}=i$, then the above is equal to 
-\[\int_{\Sigma_2}P(x|Y=1)P(Y=1)dx+\int_{\Sigma_1}P(x|Y=2)P(Y=2)dx\]
-Over $\Sigma_1$, we have that $P(x|Y=1)P(Y=1)\leq P(x|Y=2)P(Y=2)$ by the form of the Bayes classifier, and vice versa for $\Sigma_2$. Thus, the integrals above are bounded above by
-\begin{align*}
-&\frac{1}{2}\int_{\Sigma_2}P(x|Y=1)P(Y=1)+P(x|Y=2)P(Y=2)dx+ \frac{1}{2}\int_{\Sigma_1}P(x|Y=1)P(Y=1)+P(x|Y=2)P(Y=2)dx\\
-    &=\frac{1}{2}\int_{\rn^n}P(x|Y=1)P(Y=1)+P(x|Y=2)P(Y=2)dx
-\end{align*}
-if we replace the integrands with the average of the two integrands. Now, using the AM-GM inequality, we have that this is bounded above by 
-\begin{align*}
-\int_{\rn^n}\sqrt{P(x|Y=1)P(Y=1)P(x|Y=2)P(Y=2)}dx&=\sqrt{P(Y=1)P(Y=2)}\int_{\rn^n}\sqrt{P(x|Y=1)P(x|Y=2)}dx\\
-&\leq\frac{1}{2}\int_{\rn^n}\sqrt{P(x|Y=1)P(x|Y=2)}dx
-\end{align*}
-by another application of AM-GM to $\sqrt{P(Y=1)P(Y=2)}$, noting that the average of $P(Y=1)$ and $P(Y=2)$ is $\frac{1}{2}$.
-\subsection*{3}
-\ssn{a}
-The decision boundary is the point at which $P(x|Y=1)P(Y=1)=P(x|Y=2)P(Y=2)$, or $\pi_1\exp\left((x-\mu_1)^2/(2\sigma^2)\right)=\pi_2\exp\left((x-\mu_2)^2/(2\sigma^2)\right)$. Solving for $x$, we have 
-\[x^*=\frac{2\sigma\log\frac{\pi_1}{\pi_2}+\mu_2^2-\mu_1^2}{2(\mu_2-\mu_1)}\]
-\ssn{b}
-The error probability is
-\[\int_{-\infty}^{x^*}P(x|Y=2)P(Y=2)dx+\int_{x^*}^\infty P(x|Y=1)P(Y=1)dx=\pi_2\Phi\left(\frac{x^*-\mu_2}{\sigma}\right)+\pi_1-\pi_1\Phi\left(\frac{x^*-\mu_1}{\sigma}\right)\]
+The first difficulty came when trying to install the packages.
+I ran into a syntax error while installing czipfile via pip, and this turned out to be because my system defaults to Python 3.
+Deleting the virtualenv and re-running mkvirtualenv with the option \verb|--python=/usr/bin/python2.7| fixed the issue.
 
-Rewrite $x^*-\mu_2=\frac{2\sigma\log\frac{\pi_1}{\pi_2}}{2(\mu_2-\mu_1)}+\frac{\mu_2+\mu_1}{2}-\mu_2$.
-Then, dividing this by $\sigma$ produces a term constant in $\sigma$ and a term that goes to $-\infty$ as $\sigma\to0$.
-Thus, the $\pi_2$ term in the error vanishes.
-Similarly, $\frac{x^*-\mu_1}{\sigma}$ goes to $\infty$ as $\sigma\to0$, which means that the overall error term approaches $\pi_1-\pi_1=0$.
-\ssn{c}
-As $\pi_1\to0$, the log term in the decision boundary approaches $-\infty$, dragging the decision boundary with it, as all other things are fixed.
-In this case, always classifying things as class 2 would produce an error rate of $\pi_1$, which is known to be small.
-\ssn{d}
-{
-\nc{\lot}{L_{1,2}}
-\nc{\lto}{L_{2,1}}
-Let $S_1$ be the set of points in class $1$ and $S_2$ the same for class 2.
-If we predict class 1, the expected loss is $E_X(\lto P(Y=2|X)$.
-If we predict class 2, then it is $E_X(\lot P(Y=1|X))$.
-If we minimize this pointwise, we should predict class 1 when $\lto P(Y=2|X)\leq\lot P(Y=1|X)$, and class 2 otherwise.
-
-Rewriting the conditional probabilities, the prediction is then $\text{argmax}\left(\lto P(X|Y=2)P(Y=2), \lot P(X|Y=1)P(Y=1)\right)$. Setting these equal to find the boundary, we have
-\[\lto \pi_2\exp\left((x-\mu_2)^2/(2\sigma^2)\right)=\lot\pi_1\exp\left((x-\mu_1)^2/(2\sigma^2)\right)\]
-Solving for $x$, we obtain
-\[x^*=\frac{2\sigma\log\frac{\lot\pi_1}{\lto\pi_2}+\mu_2^2-\mu_1^2}{2(\mu_2-\mu_1)}\]
-Now, if we choose $\lot$ to be proportional to $\pi_1^{-1}$ and choose $\lto$ to be proportional to $\pi_2^{-1}$ (possibly with different constants), we will avoid the degeneracy problem.
-\ssn{e}
-Using these values, we have that $x^*=1$, so plugging into the expression for the error rate from (b) gives $0.5\Phi(-1)+0.5-0.5\Phi(1)=0.1587$
-}
-\subsection*{4}
-The expected loss of some decision function $\h{H}$ is $E_{X, Y}(L(\h{H}(x), Y)$. Writing out the integral gives
-\[\int_{\rn^n}\sum_{k=1}^KL(\h{H}(x), k)p(x, k) dx=\int_{\rn^n}p(x)\sum_{k=1}^KL(\h{H}(x), k)P(k|x)dx\]
-
-%If we define $R_i$ as the subset of $\rn^n$ for which $\h{H}(x)=i$, we can separate the integral as follows
-%\[\sum_{i=1}^K\int_{R_i}p(x)\sum_{k=1}^KL(i, k)P(k|x)dx\]
-%If there is some subset $A$ of $R_i$ for which there exists a $j$ such that  for all $x\in A$, then moving $A$ from $R_i$ to $R_j$ will reduce the value of the expected loss by comparison
-
-Define $R_i$ as $\{x\in\rn^n|\h{H}(x)=i\}$.
-Then, suppose that there exists some subset of nonzero measure $A$ of $R_i$ such that $\sum_{k=1}^KL(i, k)P(k|x)\ge\sum_{k=1}^KL(j, k)P(k|x)$.
-Decompose the above integral as
-\[\int_{\rn^n-A}p(x)\sum_{k=1}^KL(\h{H}(x), k)P(k|x)dx+\int_Ap(x)\sum_{k=1}^KL(i, k)P(k|x)dx\]
-Then, changing the class of $A$ from $i$ to $j$ will result in a strict decrease in the expected loss, which means that the optimal decision rule must satisfy, for each $i$, $\sum_{k=1}^KL(i, k)P(k|x)\ge\sum_{k=1}^KL(j, k)P(k|x)$ for all $j\neq i$ and almost all $x\in R_i$.
-The rule given satisfies this condition, and since there are a finite number of classes, changing the rule on a set of measure zero from the region associated with each class will not result in a change in the expected loss.
-Thus, the rule given is optimal.
-\subsection*{5}
-\ssn{a}
-We have $\cov(X_2,X_2)=v_2$, so writing $X_2=\alpha X_1+Z$ gives
-\[\cov(\alpha X_1+Z, \alpha X_1+Z)=\cov(\alpha X_1, \alpha X_1)+0+\cov(Z, Z)=\alpha^2v_1+\var(Z)=v_2\]
-In addition, using $\cov(X_1, X_2)=a$ gives us
-\[a=\cov(X_1, \alpha X_1+Z)=\alpha v_1+0\implies \alpha=\frac{a}{v_1}\]
-Substituting into the first equation gives us $\var(Z)=v_2-\frac{a^2}{v_1}$
-
-Finally, since $E(X_1)=E(X_2)=0$, we must also have $E(Z)=0$.
-\ssn{b}
-We know that $Z$ is Gaussian because we can express $Z$ as the sum of two Gaussians, $X_2-\alpha X_1$.
-Thus, it is independent from $X_1$ because their covariance is zero.
-The variance was derived above.
-\ssn{c}
-We have $E(X_2|X_1=x)=E(\alpha x+Z)=\alpha x$ and $\var(X_2|X_1=x)=\var(\alpha x+Z)=\var(Z)$. 
-\ssn{d}
-Since the samples are all independent, the cross terms in the formula for the variance of a sum disappear and we get
-\[\var(\h{a})=\frac{1}{n^2}\sum_{i=1}^n\var(X_{i1}X_{i2})=\frac{1}{n}\var(X_1X_2)\]
-Then, we have $\var(X_1X_2)=E((X_1X_2)^2)-E(X_1X_2)^2=E((X_1X_2)^2)-a^2$, since the means are zero.
-
-To obtain the first term, note that it is a cross moment and can be obtained by applying $\frac{\partial^4}{\partial x_1^2\partial x_2^2}$ to the mgf of the distribution and evaluating at zero.
-The mgf itself is
-\[\exp\left(\frac{1}{2}\openm x_1&x_2\closem\openm v_1&a\\a&v_2\closem\openm x_1\\x_2\closem\right)=\exp\left(\frac{1}{2}(x_1^2v_1+2x_1x_2a+x_2^2v_2)\right)\]
-
-Differentiating this wrt $x_1$ twice and wrt $x_2$ once gives 
-\[ax_1v_1T+x_1v_1(x_1v_1+x_2a)(x_2v_2+x_1a)T+v_1(x_2v_2+x_1a)T\]
-where $T$ is the exponential term from above. Then, taking the derivatives and then evaluating gives $v_1v_2$, so the variance of $\h{a}$ is $\frac{1}{n}(v_1v_2-a^2)$.
-\subsection*{6}
+The Python version problem came up again when creating the bin/notebook file. I modified the line to start with \verb|python2.7 /usr/bin/ipython| to force the script to use 
 \end{document}

### 2014-04-07 cs250-hw/hw1.tex
c1

@@
 \DeclareMathOperator{\cov}{cov}
+\DeclareMathOperator{\var}{var}
 \DeclareMathOperator{\rank}{rank}
+\DeclareMathOperator{\tr}{tr}
 \DeclareMathOperator{\im}{im}
@@
 \subsection*{1}
+\ssn{a}
+First, note that since $F$ is strictly increasing, we must have that $F(x)\in(0,1)$ for any $x\in\rn$ and $F^{-1}$ exists.
+Then, for any $y\in(0,1)$, we have $P(Y<y)=P(F(X)<y)=P(X<F^{-1}(y))=F(F^{-1}(y))=y$, which implies that $Y$ has a cdf of $0$ on $(-\infty, 0)$, $y$ on $[0,1]$, and $1$ on $(1, \infty)$.
+Thus, we have $Y\sim\text{Unif}(0, 1)$.
 
+Going the other way, we have $P(X<x)=P(F^{-1}(U)<x)=P(U<F(x))=F(x)$.
+
+\ssn{b}
+The joint density $f_{X,Y}(x,y)$ of $X$ and $Y$ is the product of their densities (which is 1) since they are independent. Thus, for the difference, for $z\in(0,1]$, we have
+\[P(Z<z)=P(X-Y<z)=P(X<z+Y)=\int_0^{1-z}\int_0^{z+y}1dxdy+\int_{1-z}^1\int_0^11dxdy=\int_0^{1-z}(z+y)dy+z=-\frac{z^2}{2}+z+\frac{1}{2}\]
+For $z\in[-1,0]$, we have
+\[P(Z<z)=P(X-Y<z)=P(X<z+Y)=\int_{-z}^1\int_0^{y+z}1dxdy=\int_{-z}^1(z+y)dy=\frac{z^2}{2}+z+\frac{1}{2}\]
+Differentiating produces the density $z+1$ on $[-1,0]$ and $-z+1$ on $(0,1]$.
+
+For the minimum, we want $X<z$ or $Y<z$, which is equivalent to the negation of ($X\geq z$ and $Y\geq z$), so the integral for the cdf of $Z$ is 
+\[1-\int_{1-z}^1\int_{1-z}^11dxdy=1-(1-z)^2\]
+and differentiating this produces the density $2(1-z)$.
+\ssn{c}
+The density of $X$ is $f_X(x)=\frac{1}{\sqrt{2\pi}}e^{-\frac{x^2}{2}}$. Taking the expectation of $e^X$ wrt this density gives
+\[\frac{1}{\sqrt{2\pi}}\int_{-\infty}^\infty\exp\left(-\frac{x^2}{2}+x\right)dx=\frac{1}{\sqrt{2\pi}}e^{1/2}\int_{-\infty}^\infty\exp\left(-\frac{(x-1)^2}{2}\right)=e^{1/2}\]
+The variance can be calculated by $E(Y^2)-(E(Y))^2$, and evaluating $E(Y^2)=E(e^{2X})$ gives
+\[\frac{1}{\sqrt{2\pi}}\int_{-\infty}^\infty\exp\left(-\frac{x^2}{2}+2x\right)dx=\frac{1}{\sqrt{2\pi}}e^2\int_{-\infty}^\infty\exp\left(-\frac{(x-2)^2}{2}\right)=e^2\]
+so the variance is $e^2-e$
+\ssn{d}
+$\var_Y(Y|X)=E_Y(Y^2|X)-(E_Y(Y|X))^2$, so the first term is $E_X(\var_Y(Y|X))=E_X(E_Y(Y^2|X))-E_X((E_Y(Y|X))^2)=E_Y(Y^2)-E_X((E_Y(Y|X))^2)$.
+Similarly, the second term is $E_X((E_Y(Y|X))^2)-(E_X(E_Y(Y|X)))^2=E_X((E_Y(Y|X))^2)-(E_Y(Y))^2$.
+Adding them together cancels out the $E_X((E_Y(Y|X))^2)$ term, which results in $E_Y(Y^2)-(E_Y(Y))^2=\var(Y)$.
+\subsection*{2}
+\ssn{a}
+Since $X^TX$ is nonsingular and $n>d$, this means that $X$ has rank $d$.
+The least-squares estimate $\h{\beta}$ satisfies $\h{\beta}=\text{argmin}_\beta\|X\beta-y\|_2^2$.
+The solution is given by $X^+y$, where $X^+$ is the Moore-Penrose pseudoinverse of $X$.
+Due to $X$ having full column rank, we can write $X^+=(X^TX)^{-1}X^T$, which means that $\h{\beta}=(X^TX)^{-1}X^Ty$, so the least-squares estimates are $X\h{\beta}=X(X^TX)^{-1}X^Ty=Hy$.
+\ssn{b}
+$HX=X(X^TX)^{-1}X^TX=X$
+\ssn{c}
+$H^T=(X(X^TX)^{-1}X^T)^T=X(X^TX)^{-T}X^T=X(X^TX)^{-1}X^T$
+\ssn{d}
+$HH=X(X^TX)^{-1}X^TX(X^TX)^{-1}X^T=X(X^TX)^{-1}X^T=H$
+\ssn{e}
+By (d), $H$ is a projection matrix.
+Thus, we just need to show that the image of $H$ is the column space.
+Since the image of $X$ is the column space (a subset of $\rn^n$), this means that we need to show that $(X^TX)^{-1}X^T$ has rank $d$ (or that its image is all of $\rn^d$).
+This is true because $(X^TX)^{-1}$ is invertible and $X^T$ has rank $d$ also.
+\ssn{f}
+Traces of products are invariant under cyclic permutation, so $\tr(H)=\tr(X(X^TX)^{-1}X^T)=\tr(X^TX(X^TX)^{-1})=\tr(I_d)=d$.
 \subsection*{3}

### 2014-04-10 stat246/hw1.tex
s1

@@
 Python code is attached.
+\ssn{c,d}
+Histograms for $e$: $d=100$ on the right, $d=10$ on the left. $n=500$ on top, $n=5000$ on bottom.
+
+\includegraphics[width=0.5\textwidth]{hw1_files/e10-500.png}
+\includegraphics[width=0.5\textwidth]{hw1_files/e100-500.png}
+
+\includegraphics[width=0.5\textwidth]{hw1_files/e10-5000.png}
+\includegraphics[width=0.5\textwidth]{hw1_files/e100-5000.png}
+
+The mean and stddev for $e$ are in the following table (presented as a pair mean,stddev):
+
+\begin{tabular}{c|c|c}
+    &$d=10$&$d=100$\\
+    \hline
+    $n=500$&$0.447,64.85$&$0.0302,60.67$\\
+    \hline
+    $n=5000$&$-4.137,204.75$&$0.00706,200.00$\\
+\end{tabular}
+
+Histograms for the eigenvalue estimates: $d=100$ on right, $d=10$ on left
+
+\includegraphics[width=0.5\textwidth]{hw1_files/m10-500.png}
+\includegraphics[width=0.5\textwidth]{hw1_files/m100-500.png}
+
+\includegraphics[width=0.5\textwidth]{hw1_files/m10-5000.png}
+\includegraphics[width=0.5\textwidth]{hw1_files/m100-5000.png}
+
+The table of means and standard deviations:
+
+\begin{tabular}{c|c|c}
+    &$d=10$&$d=100$\\
+    \hline
+    $n=500$&$-0.0214,0.0626$&$-0.274,0.0669$\\
+    \hline
+    $n=5000$&$-0.00140,0.0205$&$-0.0239,0.0197$\\
+\end{tabular}
+
+As expected, the standardized errors cluster around zero with an apparently normal distribution.
+The standard deviation is lower for $d=100$ than $d=10$ and higher for $n=5000$ than $n=500$.
+The eigenvalue ratios indicate that the highest eigenvalue of the inverse of the sample covariance is consistently lower than that of the highest eigenvalue of the actual.
 \end{document}

### 2014-04-10 stat246/hw1_files/p6.py

@@ -41,10 +41,15 @@ def generate_results():
         e100_data, m100_data = compute_results(100, a100, ssize)
         e10f = e10_data.flatten()
         e100f = e100_data.flatten()
+        m10p = np.log(100/(1/m10_data))
+        m100p = np.log(100/(1/m100_data))
 
-        print('10 mean: {0}. 100 mean: {1}'.format(e10f.mean(), e100f.mean()))
-        print('10 std: {0}. 100 std: {1}'.format(e10f.std(), e100f.std()))
+        print('e10 mean: {0}. e100 mean: {1}'.format(e10f.mean(), e100f.mean()))
+        print('e10 std: {0}. e100 std: {1}'.format(e10f.std(), e100f.std()))
 
+        print('m10 mean: {0}. m100 mean: {1}'.format(m10p.mean(), m100p.mean()))
+        print('m10 std: {0}. m100 std: {1}'.format(m10p.std(), m100p.std()))
+        '''
         plt.figure()
         hist10, bins10 = np.histogram(e10f, bins=50)
         width10 = 0.6*(bins10[1] - bins10[0])
@@ -58,5 +63,18 @@ def generate_results():
         centers100 = (bins100[:-1] + bins100[1:]) / 2
         plt.bar(centers100, hist100, align='center', width=width100)
         plt.savefig('e100-{0}.png'.format(ssize), bbox_inches='tight')
+        '''
 
-generate_results()
+        plt.figure()
+        hist10, bins10 = np.histogram(m10p, bins=50)
+        width10 = 0.6*(bins10[1] - bins10[0])
+        centers10 = (bins10[:-1] + bins10[1:]) / 2
+        plt.bar(centers10, hist10, align='center', width=width10)
+        plt.savefig('m10-{0}.png'.format(ssize), bbox_inches='tight')
+
+        plt.figure()
+        hist100, bins100 = np.histogram(m100p, bins=50)
+        width100 = 0.6*(bins100[1] - bins100[0])
+        centers100 = (bins100[:-1] + bins100[1:]) / 2
+        plt.bar(centers100, hist100, align='center', width=width100)
+        plt.savefig('m100-{0}.png'.format(ssize), bbox_inches='tight')

### 2014-04-10 stat246/hw1.tex
s1 done

@@
 \ssn{e}
-Using these values, we have that $x^*=1$, so plugging into the expression for the error rate from (b) gives $0.5\Phi(-1)+0.5-0.5\Phi(1)=0.1587$
+Using these values, we have that $x^*=1$, so plugging into the expression for the error rate from (b) gives $0.5\Phi(-1)+0.5-0.5\Phi(1)=0.1587$.
+
+If we estimate the two means as $\conj{x_1}$ and $\conj{x_2}$, using the expression from (a), the decision boundary would be 
+{
+    \nc{\cxi}{\conj{x_1}}
+    \nc{\cxii}{\conj{x_2}}
+    $\frac{\cxii^2-\cxi^2}{2(\cxii-\cxi)}$
+}
+
+Code for the computation of error rates from the samples is attached, using the expression from (b). A histogram follows.
+
+\includegraphics[width=0.6\textwidth]{hw1_files/p3.png}
+
 }

### 2014-04-10 stat246/hw1_files/p3.py

@@ -1,0 +1,24 @@
+from scipy.stats import norm
+import matplotlib.pyplot as plt
+import numpy as np
+
+
+def compute_error_rates():
+    error_rates = np.zeros(100)
+    for i in range(100):
+        sample1 = np.random.randn(10)
+        sample2 = np.random.randn(10) + 2
+        m1 = sample1.mean()
+        m2 = sample2.mean()
+        x = (m2*m2 - m1*m1)/(2*(m2-m1))
+        error_rates[i] = 0.5*norm.cdf(x-m2)+0.5-0.5*norm.cdf(x-m1)
+
+    
+    plt.figure()
+    hist, bins = np.histogram(error_rates, bins=50)
+    width = 0.6*(bins[1] - bins[0])
+    centers = (bins[:-1] + bins[1:]) / 2
+    plt.bar(centers, hist, align='center', width=width)
+    plt.savefig('p3.png', bbox_inches='tight')
+
+    return error_rates

### 2014-04-10 cs250-hw/hw1.tex
c1

@@
 The least-squares estimate $\h{\beta}$ satisfies $\h{\beta}=\text{argmin}_\beta\|X\beta-y\|_2^2$.
-The solution is given by $X^+y$, where $X^+$ is the Moore-Penrose pseudoinverse of $X$.
-Due to $X$ having full column rank, we can write $X^+=(X^TX)^{-1}X^T$, which means that $\h{\beta}=(X^TX)^{-1}X^Ty$, so the least-squares estimates are $X\h{\beta}=X(X^TX)^{-1}X^Ty=Hy$.
+Decompose $X$ via the SVD $X=U\Sigma V^T$.
+Then, we have
+\[\|X\beta-y\|_2^2=\norm{U\Sigma V^T\beta-y}_2^2=\norm{\Sigma \gamma-U^Ty}_2^2\]
+where $\gamma=V^T\beta$, by the unitary invariance of the $2$-norm.
+Let $z=U^Ty$.
+We know that precisely the first $d$ entries in $\Sigma $ are nonzero because $X$ is rank $d$, so the norm expression is
+\[\sum_{i=1}^d(\sigma_i\gamma_i-z_i)^2+\sum_{i=d+1}^nz_i^2\]
+To minimize this, we set $\gamma_i=z_i/\sigma_i$ for $i\in[1..d]$, and we can choose the others arbitrarily.
+Might as well make the others zero, so then we have that $\h{\beta}=V\h{\gamma}=V\Sigma^+U^Ty$, where $\Sigma^+$ is diagonal with entries $\frac{1}{\sigma_i}$ where $\sigma_i\neq0$ and zero elsewhere.
+
+Now, note that both $V\Sigma^+U^Ty$ and $(X^TX)^{-1}X^T$ satisfy the requirements for being the Moore-Penrose pseudoinverse of $X$, so they must be equal. Thus, we have $\h{\beta}=(X^TX)^{-1}X^Ty$.
 \ssn{b}

### 2014-04-10 cs250-hw/hw1.tex
c1

@@
 
-Now, note that both $V\Sigma^+U^Ty$ and $(X^TX)^{-1}X^T$ satisfy the requirements for being the Moore-Penrose pseudoinverse of $X$, so they must be equal. Thus, we have $\h{\beta}=(X^TX)^{-1}X^Ty$.
+Now, note that both $V\Sigma^+U^TX$ and $(X^TX)^{-1}X^T$ satisfy the requirements for being the Moore-Penrose pseudoinverse of $X$.
+The calculations are a bit tedious to type out and not very illuminating, so I'll omit them.
+Thus, by the uniqueness of the pseudoinverse, they must be equal.
+Thus, we have $\h{\beta}=(X^TX)^{-1}X^Ty$.
 \ssn{b}

### 2014-04-11 stat246/hw2.tex
s2

\documentclass{article}
\usepackage{geometry}
\usepackage[namelimits,sumlimits]{amsmath}
\usepackage{amssymb,amsfonts}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage[cm]{fullpage}
\newcommand{\tab}{\hspace*{5em}}
\newcommand{\conj}{\overline}
\newcommand{\dd}{\partial}
\newcommand{\ep}{\epsilon}
\newcommand{\openm}{\begin{pmatrix}}
\newcommand{\closem}{\end{pmatrix}}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\Null}{null}
\newcommand{\nc}{\newcommand}
\newcommand{\rn}{\mathbb{R}}
\nc{\cn}{\mathbb{C}}
\nc{\ssn}[1]{\subsubsection*{#1}}
\nc{\inner}[2]{\langle #1,#2\rangle}
\nc{\h}[1]{\widehat{#1}}
\nc{\tl}[1]{\widetilde{#1}}
\nc{\norm}[1]{\left\|{#1}\right\|}
\begin{document}

Name: Hall Liu

Date: \today 
\vspace{1.5cm}
\subsection*{1}
The mgf of $Y$ is $M_Y(t)=E(e^{t^TY})=E(e^{t^TAx})=E(e^{(A^Tt)^Tx})=M_X(A^Tt)$, which we can write as $\exp\left(t^TA\mu+\frac{1}{2}t^TA\Sigma A^Tt\right)$. This is the mgf of a normal RV with mean $A\mu$ and covariance $A\Sigma A^T$, so $Y$ must be also distributed as such.

\subsection*{2}
\ssn{a}
We have that $f(x,y)=f_Y(y|x)f_X(x)$, so $f(x,y)$ is jointly normal iff this product has the form of one. In (b), we show that there exists a particular definition for $Z$ such that $f_Z(z|x)=f_Y(y|x)$ and $X$ and $Z$ are jointly normal. Thus, since $f(x,z)$ and $f(x,y)$ are equivalent, $f(x,y)$ is normal.
\ssn{b}
Let $Z=AX+b+U$, $U\sim N(0,Q)$ and independent of $X$. Then, if given a value for $X$, we have $Z=Ax+b+U$, so consequently we have $f(z|x)=N(Ax+b, Q)$, the same as that of $f(y|x)$.

The joint mgf of $X$ and $Z$ can be written as 
\[E\left(\exp\left(\openm t_1^T&t_2^T\closem\openm X\\Z\closem\right)\right)=E(\exp(t_1^TX+t_2^T(AX+b+U)))=E(\exp((t_1+A^Tt_2)^TX)\exp(t_2^Tb)\exp(t_2^TU))\]
Resolving the expected value and expressing the result using the mgfs of $X$ and $U$, we have
\[\exp\left((t_1+A^Tt_2)^T\mu+t_2^Tb+\frac{1}{2}(t_1+A^Tt_2)^T\Sigma (t_1+A^Tt_2)+\frac{1}{2}t_2^TQt_2\right)\]
Now, we note that 
\[(t_1+A^Tt_2)^T\mu+t_2^Tb=\openm t_1^T&t_2^T\closem\openm\mu\\A\mu+b\closem\text{ and}\]
\[(t_1+A^Tt_2)^T\Sigma (t_1+A^Tt_2)+t_2^TQt_2=\openm t_1^T&t_2^T\closem\openm \Sigma &\Sigma A^T\\A\Sigma &A\Sigma A^T+Q\closem\openm t_1\\t_2\closem\]
Thus, if we let $t=\openm t_1\\t_2\closem$, $\mu'=\openm\mu\\A\mu+b\closem$, and $\Sigma'=\openm \Sigma &\Sigma A^T\\A\Sigma &A\Sigma A^T+Q\closem$, we have that the joint mgf of $X,Z$ is $\exp\left(t^T\mu'+\frac{1}{2}t^T\Sigma't\right)$, which means that it's multivariate normal with mean $\mu'$ and covariance $\Sigma'$.
\ssn{c}
The marginal is simply restricting the joint distribution to the relevant components. In the case of $Y$, its marginal distribution is then $N(A\mu+b, A\Sigma A^T+Q)$. For the density of $X$ conditional on $Y=y$, (2.81) and (2.82) in the book give us that it's normal with $\mu_{x|y}=\mu+\Sigma A^T(A\Sigma A^T+Q)^{-1}(y-A\mu+b)$ and $\Sigma_{x|y}=\Sigma-\Sigma A^T(A\Sigma A^T+Q)^{-1}A\Sigma$
\end{document}

### 2014-04-11 stat246/hw2.tex
s2

@@
 \DeclareMathOperator{\im}{im}
+\DeclareMathOperator{\tr}{tr}
 \DeclareMathOperator{\Span}{span}
@@
 \subsection*{1}
-The mgf of $Y$ is $M_Y(t)=E(e^{t^TY})=E(e^{t^TAx})=E(e^{(A^Tt)^Tx})=M_X(A^Tt)$, which we can write as $\exp\left(t^TA\mu+\frac{1}{2}t^TA\Sigma A^Tt\right)$.
+The mgf of $Y$ is $M_Y(t)=E(e^{t^TY})=E(e^{t^TAX})=E(e^{(A^Tt)^TX})=M_X(A^Tt)$, which we can write as $\exp\left(t^TA\mu+\frac{1}{2}t^TA\Sigma A^Tt\right)$.
 This is the mgf of a normal RV with mean $A\mu$ and covariance $A\Sigma A^T$, so $Y$ must be also distributed as such.
@@
 For the density of $X$ conditional on $Y=y$, (2.81) and (2.82) in the book give us that it's normal with $\mu_{x|y}=\mu+\Sigma A^T(A\Sigma A^T+Q)^{-1}(y-A\mu+b)$ and $\Sigma_{x|y}=\Sigma-\Sigma A^T(A\Sigma A^T+Q)^{-1}A\Sigma$
+\subsection*{3}
+The likelihood of the data is 
+\[\prod_{i=1}^N\frac{C}{|\Sigma|}\exp\left((X_i-\mu_{k_i})^T\Sigma^{-1}(X_i-\mu_{k_i})\right)\pi_{k_i}\]
+where $C$ is a constant that depends on none of the parameters.
+We have immediately that $\h{\pi}_{k_i}$ is $\frac{\#i}{N}$, where $\#i$ is the number of occurrances of class $i$ in the data.
+Taking the loglikelihood (and ignoring the constant $C$), we have
+\[\sum_{i=1}^N-\log|\Sigma|+(X_i-\mu_{k_i})^T\Sigma^{-1}(X_i-\mu_{k_i})+\log(\pi_{k_i})\]
+To minimize this expression over the $\mu_k$, write the expression with an inner sum over points with the same class:
+\[-N\log|\Sigma|+\sum_{k=1}^K\sum_{X:Y(X)=k}(X-\mu_k)^T\Sigma^{-1}(X-\mu_k)\]
+The $\pi_{k_i}$ terms have been discarded because they are not dependent on the parameters.
+Now, we can simply minimize over each class, and find that $\mu_k$ is the sample mean over all points that are in class $k$.
+
+Now, let $S_k$ be the sample covariance over class $k$, or $\sum_{i=1}^{\#k}(X_i-\conj{X})(X_i-\conj{X})^T$. Then, we can write the log-likelihood (divided out by $N$) as
+\[-\log|\Sigma|+\frac{1}{N}\sum_{i=1}^N\tr\left((X_i-\mu_{k_i})(X_i-\mu_{k_i})^T\Sigma^{-1}\right)=-\log|\Sigma|+\frac{1}{N}\sum_{k=1}^K\tr\left(\#kS_k\Sigma^{-1}\right)=-\log|\Sigma|+\tr\left(\left(\sum_{k=1}^K\frac{\#k}{N}S_k\right)\Sigma^{-1}\right)\]
+It was shown that minimizing an expression of the form $-\log|\Sigma|+\tr(A\Sigma^{-1})$ over $\Sigma$ is given by $\Sigma=A$, so we have that the MLE for the pooled covariance is $\sum_{k=1}^K\frac{\#k}{N}S_k$
 \end{document}

### 2014-04-12 stat246/hw2.tex
s2

@@
 \DeclareMathOperator{\Null}{null}
+\DeclareMathOperator{\flatv}{flat}
 \newcommand{\nc}{\newcommand}
@@
 The likelihood of the data is 
-\[\prod_{i=1}^N\frac{C}{|\Sigma|}\exp\left((X_i-\mu_{k_i})^T\Sigma^{-1}(X_i-\mu_{k_i})\right)\pi_{k_i}\]
+\[\prod_{i=1}^N\frac{C}{\sqrt{|\Sigma|}}\exp\left(-\frac{1}{2}(X_i-\mu_{k_i})^T\Sigma^{-1}(X_i-\mu_{k_i})\right)\pi_{k_i}\]
 where $C$ is a constant that depends on none of the parameters.
 We have immediately that $\h{\pi}_{k_i}$ is $\frac{\#i}{N}$, where $\#i$ is the number of occurrances of class $i$ in the data.
-Taking the loglikelihood (and ignoring the constant $C$), we have
+Taking the loglikelihood (and ignoring the constant $C$ and the factor of $\frac{1}{2}$), we have
 \[\sum_{i=1}^N-\log|\Sigma|+(X_i-\mu_{k_i})^T\Sigma^{-1}(X_i-\mu_{k_i})+\log(\pi_{k_i})\]
@@
 It was shown that minimizing an expression of the form $-\log|\Sigma|+\tr(A\Sigma^{-1})$ over $\Sigma$ is given by $\Sigma=A$, so we have that the MLE for the pooled covariance is $\sum_{k=1}^K\frac{\#k}{N}S_k$
+\subsection*{4}
+Let $g(\eta)=\exp(A(\eta))$.
+Then, we have $f(x;\eta)=\frac{h(x)}{g(\eta)}e^{\eta^TT(x)}$, and integrating this over $\rn^n$ (the space $x$ lies in) gives $1$, so we have $g(\eta)=\int_{\rn^n}h(x)e^{\eta^TT(x)}dx$.
+
+Taking the first derivative of $g$ gives $\nabla g(\eta)=e^{A(\eta)}\nabla A(\eta)$ and 
+\[\nabla g(\eta)=\int h(x)e^{\eta^TT(x)}T(x)dx\implies\nabla A(\eta)=\int h(x)e^{\eta^TT(x)-A(\eta)}T(x)=E(T(x))\]
+Using the notation $\nabla^2 g$ for the second derivative, we have that $\nabla^2g(\eta)=e^{A(\eta)}\left(\nabla^2A(\eta)+\nabla A(\eta)\nabla A(\eta)^T\right)$ and
+\[\nabla^2 g(\eta)=\int h(x)e^{\eta^TT(x)}T(x)T(x)^Tdx\]
+Combining the two expressions gives us 
+\[e^{A(\eta)}\left(\nabla^2 A(\eta)+E(T(x))E(T(x)^T)\right)=\int h(x)e^{\eta^TT(x)}T(x)T(x)^Tdx\]
+Dividing out by $e^{A(\eta)}$ and note that the RHS is equal to $E(T(x)T(x)^T)$, so subtracting off $E(T(x))E(T(x)^T)$ from both sides gives $\nabla^2A(\eta)=E(T(x)T(x)^T)-E(T(x))E(T(x)^T)$, which is the covariance of $T(x)$.
+\subsection*{5}
+The density of the multivariate normal distribution can be written as 
+\[\frac{1}{(2\pi)^{d/2}|\Sigma|}\exp\left(-\frac{1}{2}\tr\left((x-\mu)(x-\mu)^T\Sigma^{-1}\right)\right)\]
+We can let $(2\pi)^{-d/2}$ be $h(x)$ and put $|\Sigma^{-1}|$ into $g(\mu, \Sigma)$, so we are left with considering the exponential term. Expanding out the exponent, we get
+\[\frac{-1}{2}\tr\left(xx^T-\mu x^T-x\mu^T+\mu\mu^T)\Sigma^{-1}\right)=-\frac{1}{2}\left(\tr(xx^T\Sigma^{-1})-2\tr(x\mu^T\Sigma^{-1})+\tr(\mu\mu^T\Sigma^{-1})\right)\]
+Now, let $g(\mu,\Sigma)=|\Sigma^{-1}|\exp\left(-\frac{1}{2}\mu^T\Sigma^{-1}\mu\right)$, so we can now discard the last term from the above expression and try to find what $T(x)$ is. We're left with
+\[-\frac{1}{2}\tr(xx^T\Sigma^{-1})+\mu^T\Sigma^{-1}x\]
+
+Define $\flatv(A)$ for a $n\times m$ matrix $A$ to be a $nm$-long column vector that consists of the rows of $A$ laid out in sequential order.
+Then, we have that $\tr(AB)=\flatv(A)^T\flatv(B^T)$ for matrices $A,B$.
+Let $\eta=\openm\Sigma^{-1}\mu\\-\frac{1}{2}\flatv(\Sigma^{-1})\closem$ and let $T(x)$ be $\openm x\\\flatv(xx^T)\closem$.
+Then, we have that
+\[\eta^TT(x)=\openm\mu^T\Sigma^{-1}&-\frac{1}{2}\flatv(\Sigma^{-1})^T\closem\openm x\\\flatv(xx^T)\closem=\mu^T\Sigma^{-1}x-\frac{1}{2}\tr(\Sigma^{-1}xx^T)\]
+which corresponds to the terms inside the exponent that were not absorbed by $g(\mu,\Sigma)$. Thus, we have that the dimension of $T(x)$ and $\eta$ is $d+d^2$.
+
+In summary, $h(x)=(2\pi)^{-d/2}$, $g(\mu,\Sigma)=|\Sigma^{-1}|\exp\left(-\frac{1}{2}\mu^T\Sigma^{-1}\mu\right)$, $\eta=\openm\Sigma^{-1}\mu\\-\frac{1}{2}\flatv(\Sigma^{-1})\closem$, and $T(x)=\openm x\\\flatv(xx^T)\closem$.
+\subsection*{6}
+
 \end{document}

### 2014-04-12 stat246/hw2.tex
s2

@@
 \subsection*{6}
+\ssn{a}
+If we write $x^{\alpha-1}=e^{(\alpha-1)\log(x)}$, the density of the gamma distribution becomes
+\[\frac{1}{\Gamma(\alpha)\beta^\alpha}\exp\left(-\frac{x}{\beta}+(\alpha-1)\log(x)\right)=\frac{1}{\Gamma(\alpha)\beta^\alpha}\exp\left(\openm-\beta^{-1}&\alpha-1\closem\openm x\\\log(x)\closem\right)\]
+Thus, we have $\eta=\openm-\beta^{-1}\\\alpha-1\closem$, $T(x)=\openm x\\\log(x)\closem$, $g(\eta)=\frac{1}{\Gamma(\alpha)\beta^\alpha}$, and $h(x)=1$.
+
+By (2.228), the ML estimator for $\eta$ satisfies
+\[-\nabla \log g(\eta)=\frac{1}{N}\sum_{i=1}^NT(x_i)\]
+In terms of the components of $\eta$, we have $g(\eta)=\frac{(-\eta_1)^{\eta_2+1}}{\Gamma(\eta_2+1)}$.
+The negative log of this is $\log\Gamma(\eta_2+1)-(\eta_2+1)\log(-\eta_1)$, and its gradient is $\openm\psi^{(0)}(\alpha)-\log\beta\\-\frac{\alpha}{\beta}\closem$, where $\psi^{(0)}$ is the polygamma function.
+
+Then, the pair of equations for $\alpha$ and $\beta$ are 
+\begin{align*}
+    \psi^{(0)}(\alpha)-\log\beta&=\conj{x}\text{ and}\\
+    -\frac{\alpha}{\beta}&=\frac{1}{N}\sum_{i=1}^N\log x_i\\
+\end{align*}
+Due to the lack of a closed-form inverse for the polygamma function, there probably exists no closed-form solution to this system.
+\ssn{b}
+The notation got really confusing because $\beta$ is used as a scale parameter here rather than the rate parameter that it usually stands for.
+I'm just going to flip it around and let $\theta=\beta^{-1}$.
+If we hold $\alpha$ fixed, we can write the gamma-$\alpha$ distribution with parameter $\theta$ in the exponential family form as
+\[\frac{x^{\alpha-1}}{\Gamma(\alpha)}\theta^\alpha e^{\theta(-x)}\]
+with $g(\theta)=\theta^\alpha$ and $T(x)=-x$. Then, by (2.229), the conjugate prior for $\theta$ has the form
+\[f(\chi, \nu)\theta^{\alpha\nu}e^{\nu\theta\chi}\]
+where $\chi$ is a hyperparameter.
+If we set $\nu=\frac{a-1}{\alpha}$, this has the functional form of a gamma distribution with parameters $a,b$($b$ is a rate parameter), where $b=-\frac{\chi(a-1)}{\alpha}$.
+Thus, the conjugate prior of $\theta$ is a gamma distribution, which means that the conjugate prior of $\beta$ is an inverse gamma distribution.
+
+If we work in terms of $\theta$, the likelihood for the data $X_i$ is 
+\[\frac{\theta^{N\alpha}}{\Gamma(\alpha)^N}\prod_{i=1}^NX_i^{\alpha-1}e^{-X_i\theta}\]
+Multiplying this by the conjugate prior produces
+\[\left(\frac{\prod_{i=1}^NX_i^{\alpha-1}}{\Gamma(\alpha)^N}\cdot\frac{b^a}{\Gamma(a)}\right)\theta^{N\alpha+a-1}e^{-\theta\sum_{i=1}^NX_i-b\theta}\]
+Up to some normalizing constant that absorbs all the stuff in the front in the parens, this is another gamma distribution with shape $a+N\alpha$ and rate $b+\sum X_i$, so the posterior mean of $\theta$ is $\frac{a+N\alpha}{b+\sum X_i}$, which means that the posterior mean of $\beta$ is the reciprocal of this.
 

### 2014-04-13 stat246/hw2_files/p7.py
s2

@@ -1,0 +1,60 @@
+import numpy as np
+
+m1 = np.array([[0],[0]])
+s1 = np.array([[1,0],[0,0.1]])
+
+m2 = np.array([[0],[2]])
+s2 = np.array([[0.1,0],[0,1]])
+
+def estimate_from_training_sample(n_samples=5):
+    # Generate the first sample
+    # Since the covariances are diagonal, the componentwise sqrt is enough.
+    sample1 = np.random.randn(2, n_samples)
+    sample1 = np.dot(np.sqrt(s1), sample1)
+    
+    # Generate the second sample.
+    sample2 = np.random.randn(2, n_samples)
+    sample2 = np.dot(np.sqrt(s2), sample2)
+    sample2 = m2 + sample2
+
+    # Calculate estimated means by averaging along axis 1
+    est_mean1 = np.reshape(sample1.mean(1), (2,1))
+    est_mean2 = np.reshape(sample2.mean(1), (2,1))
+
+    est_cov1 = np.dot(sample1 - est_mean1, (sample1 - est_mean1).T) / n_samples
+    est_cov2 = np.dot(sample2 - est_mean2, (sample2 - est_mean2).T) / n_samples
+
+    return est_mean1, est_cov1, est_mean2, est_cov2
+
+'''
+Calculates the decision boundary between two distributions specified
+in the arguments. Returns A, b, c where the boundary is
+x^TAx+b^Tx+c=0
+'''
+def calculate_quad_boundary(mean1, cov1, mean2, cov2):
+    # Convert everything into matrices so we can do the multiplications concisely
+    mean1 = np.matrix(mean1)
+    cov1 = np.matrix(cov1)
+    mean2 = np.matrix(mean2)
+    cov2 = np.matrix(cov2)
+
+    # Logs of the determinants
+    logdet1 = np.log(np.linalg.det(cov1))
+    logdet2 = np.log(np.linalg.det(cov2))
+
+    # Precision matrices
+    prec1 = np.linalg.inv(cov1)
+    prec2 = np.linalg.inv(cov2)
+
+    # Calculate the constant term
+    c = logdet1 - logdet2 + 0.5*(mean1.T * prec1 * mean1) - 0.5*(mean2.T * prec2 * mean2)
+
+    # Calculate the linear term
+    b = mean2.T * prec2 - mean1.T * prec1
+
+    # Finally, the quadratic matrix term is the difference in the precision matrices
+    A = 0.5 * (prec1 - prec2)
+    return A, b, c
+
+'''
+

### 2014-04-13 stat246/hw2.tex
s2 done

@@
 
+\subsection*{7}
+Since the number of samples from each distribution is equal, the decision can be made by just seeing which of the likelihoods is greater.
+Doing so, we obtain the following plot of the decision boundaries, each a locus of solutions to a quadratic $x^TAx+b^Tx+c=0$.
+The black curve is the boundary obtained from using the true parameters, the red curve the non-pooled covariance, and the blue line is the linear boundary from the pooled covariance (still a quadratic, but $A=0$).
+
+\includegraphics[width=0.6\textwidth]{hw2_files/boundaries.png}
+
+The red curve from the non-pooled covariance is pretty different from the true decision boundary -- it classifies a lot of points to the side as being from the first distribution instead of the second, as the true decision boundary does.
+However, the linear decision boundary looks a lot more reasonable -- if we had to draw a line separating the two distributions (with assumed equal covariance), one with mean at $(0,2)$ and the other with mean at $(0,0)$, a straight line through $(0,1)$ would be the reasonable choice.
+Indeed, we see this in the error rates.
+The true decision boundary achieved an error rate of $0.0465$, the nonpooled covariance an error rate of $0.0955$, and the pooled covariance an error rate of $0.08$, so the simpler decision boundary actually acheived a better error rate.
+
+Also notable is the fact that the equal-covariance assumption is way off in this case.
+The two distributions are cigar-shaped, but aligned in perpendicular directions, so with enough samples, the pooled covariance should end up with two very similar eigenvalues.
+Nevertheless, with so few samples, it's still possible for it to do better than a wildly inaccurate quadratic boundary.
 \end{document}

### 2014-04-13 stat246/hw2_files/p7.py

@@ -1,4 +1,5 @@
 import numpy as np
+import matplotlib.pyplot as plt
 
 m1 = np.array([[0],[0]])
 s1 = np.array([[1,0],[0,0.1]])
@@ -6,7 +7,7 @@ s1 = np.array([[1,0],[0,0.1]])
 m2 = np.array([[0],[2]])
 s2 = np.array([[0.1,0],[0,1]])
 
-def estimate_from_training_sample(n_samples=5):
+def get_samples(n_samples=5):
     # Generate the first sample
     # Since the covariances are diagonal, the componentwise sqrt is enough.
     sample1 = np.random.randn(2, n_samples)
@@ -17,6 +18,11 @@ def estimate_from_training_sample(n_samples=5):
     sample2 = np.dot(np.sqrt(s2), sample2)
     sample2 = m2 + sample2
 
+    return sample1, sample2
+
+def get_nonpooled_estimate(sample1, sample2):
+    n_samples = sample1.shape[1]
+
     # Calculate estimated means by averaging along axis 1
     est_mean1 = np.reshape(sample1.mean(1), (2,1))
     est_mean2 = np.reshape(sample2.mean(1), (2,1))
@@ -26,12 +32,26 @@ def estimate_from_training_sample(n_samples=5):
 
     return est_mean1, est_cov1, est_mean2, est_cov2
 
+def get_pooled_estimate(sample1, sample2):
+    n_samples = sample1.shape[1]
+
+    est_mean1 = np.reshape(sample1.mean(1), (2,1))
+    est_mean2 = np.reshape(sample2.mean(1), (2,1))
+
+    est_cov1 = np.dot(sample1 - est_mean1, (sample1 - est_mean1).T) / n_samples
+    est_cov2 = np.dot(sample2 - est_mean2, (sample2 - est_mean2).T) / n_samples
+    # Since the number of samples from each class is the same, the pooled covariance is just
+    # the average of the two.
+    est_cov = 0.5 * (est_cov1 + est_cov2)
+
+    return est_mean1, est_mean2, est_cov
+
 '''
 Calculates the decision boundary between two distributions specified
 in the arguments. Returns A, b, c where the boundary is
 x^TAx+b^Tx+c=0
 '''
-def calculate_quad_boundary(mean1, cov1, mean2, cov2):
+def calculate_boundary(mean1, cov1, mean2, cov2):
[…]

### 2014-04-13 stat246/hw2_files/p7.py
s2

@@ -49,7 +49,7 @@ def get_pooled_estimate(sample1, sample2):
 '''
 Calculates the decision boundary between two distributions specified
 in the arguments. Returns A, b, c where the boundary is
-x^TAx+b^Tx+c=0
+x^TAx+b^Tx+c=0. Also works on pooled covariance -- just make cov1 == cov2
 '''
 def calculate_boundary(mean1, cov1, mean2, cov2):
     # Convert everything into matrices so we can do the multiplications concisely
@@ -78,7 +78,8 @@ def calculate_boundary(mean1, cov1, mean2, cov2):
 
 '''
 Returns a tuple (X,Y,Z) of arguments suitable to passing into the contour plot command
-Takes A, b, c as parameters (of the quadratic)
+Takes A, b, c as parameters (of the quadratic). Don't worry about this function -- it's
+mostly python/numpy magic.
 '''
 def get_contour_params(A, b, c):
     A = A.A

### 2014-04-16 cs250-hw/hw2.tex
c2

--- 2014-04-07 cs250-hw/hw1.tex
@@
 \DeclareMathOperator{\cov}{cov}
-\DeclareMathOperator{\var}{var}
 \DeclareMathOperator{\rank}{rank}
-\DeclareMathOperator{\tr}{tr}
 \DeclareMathOperator{\im}{im}
@@
 \vspace{1.5cm}
-
 \subsection*{1}
 \ssn{a}
-First, note that since $F$ is strictly increasing, we must have that $F(x)\in(0,1)$ for any $x\in\rn$ and $F^{-1}$ exists.
-Then, for any $y\in(0,1)$, we have $P(Y<y)=P(F(X)<y)=P(X<F^{-1}(y))=F(F^{-1}(y))=y$, which implies that $Y$ has a cdf of $0$ on $(-\infty, 0)$, $y$ on $[0,1]$, and $1$ on $(1, \infty)$.
-Thus, we have $Y\sim\text{Unif}(0, 1)$.
+Suppose we have a population which is uniformly distributed on the unit circle.
+Consider some optimal codebook with two clusters.
+Since the population distribution has rotational symmetry, the population clustering risk is invariant under rotations.
+Thus, we can obtain a distinct codebook which is also optimal by rotating any optimal codebook.
+Since we can approach the optimal risk by approaching any of the rotated variants of the optimal codebook, we do not necessarily have to be close to the specified optimal codebook.
+\ssn{b}
+Let $C_k$ be the optimal clustering with $k$ clusters.
+If we arbitrarily add a center $d$ to this clustering to obtain a clustering $D$, we will have $R(D)\leq R(C_k)$.
+For any point $x$ in the space we're in, we have $\min_D\|x-d_i\|^2\leq\min_{C_k}\|x-c_i\|^2$, as $C_k$ is a subset of $D$.
+Thus, when we integrate to find the expected value, the integral with $D$ will be less than or equal by comparison, which means that we can always find a clustering with $k+1$ clusters that either matches or equals the optimal with $k$ clusters.
 
-Going the other way, we have $P(X<x)=P(F^{-1}(U)<x)=P(U<F(x))=F(x)$.
-
-\ssn{b}
-The joint density $f_{X,Y}(x,y)$ of $X$ and $Y$ is the product of their densities (which is 1) since they are independent. Thus, for the difference, for $z\in(0,1]$, we have
-\[P(Z<z)=P(X-Y<z)=P(X<z+Y)=\int_0^{1-z}\int_0^{z+y}1dxdy+\int_{1-z}^1\int_0^11dxdy=\int_0^{1-z}(z+y)dy+z=-\frac{z^2}{2}+z+\frac{1}{2}\]
-For $z\in[-1,0]$, we have
-\[P(Z<z)=P(X-Y<z)=P(X<z+Y)=\int_{-z}^1\int_0^{y+z}1dxdy=\int_{-z}^1(z+y)dy=\frac{z^2}{2}+z+\frac{1}{2}\]
-Differentiating produces the density $z+1$ on $[-1,0]$ and $-z+1$ on $(0,1]$.
-
-For the minimum, we want $X<z$ or $Y<z$, which is equivalent to the negation of ($X\geq z$ and $Y\geq z$), so the integral for the cdf of $Z$ is 
-\[1-\int_{1-z}^1\int_{1-z}^11dxdy=1-(1-z)^2\]
-and differentiating this produces the density $2(1-z)$.
-\ssn{c}
-The density of $X$ is $f_X(x)=\frac{1}{\sqrt{2\pi}}e^{-\frac{x^2}{2}}$. Taking the expectation of $e^X$ wrt this density gives
-\[\frac{1}{\sqrt{2\pi}}\int_{-\infty}^\infty\exp\left(-\frac{x^2}{2}+x\right)dx=\frac{1}{\sqrt{2\pi}}e^{1/2}\int_{-\infty}^\infty\exp\left(-\frac{(x-1)^2}{2}\right)=e^{1/2}\]
-The variance can be calculated by $E(Y^2)-(E(Y))^2$, and evaluating $E(Y^2)=E(e^{2X})$ gives
-\[\frac{1}{\sqrt{2\pi}}\int_{-\infty}^\infty\exp\left(-\frac{x^2}{2}+2x\right)dx=\frac{1}{\sqrt{2\pi}}e^2\int_{-\infty}^\infty\exp\left(-\frac{(x-2)^2}{2}\right)=e^2\]
-so the variance is $e^2-e$
-\ssn{d}
-$\var_Y(Y|X)=E_Y(Y^2|X)-(E_Y(Y|X))^2$, so the first term is $E_X(\var_Y(Y|X))=E_X(E_Y(Y^2|X))-E_X((E_Y(Y|X))^2)=E_Y(Y^2)-E_X((E_Y(Y|X))^2)$.
-Similarly, the second term is $E_X((E_Y(Y|X))^2)-(E_X(E_Y(Y|X)))^2=E_X((E_Y(Y|X))^2)-(E_Y(Y))^2$.
-Adding them together cancels out the $E_X((E_Y(Y|X))^2)$ term, which results in $E_Y(Y^2)-(E_Y(Y))^2=\var(Y)$.
-\subsection*{2}
-\ssn{a}
-Since $X^TX$ is nonsingular and $n>d$, this means that $X$ has rank $d$.
-The least-squares estimate $\h{\beta}$ satisfies $\h{\beta}=\text{argmin}_\beta\|X\beta-y\|_2^2$.
-The solution is given by $X^+y$, where $X^+$ is the Moore-Penrose pseudoinverse of $X$.
-Due to $X$ having full column rank, we can write $X^+=(X^TX)^{-1}X^T$, which means that $\h{\beta}=(X^TX)^{-1}X^Ty$, so the least-squares estimates are $X\h{\beta}=X(X^TX)^{-1}X^Ty=Hy$.
-\ssn{b}
-$HX=X(X^TX)^{-1}X^TX=X$
-\ssn{c}
-$H^T=(X(X^TX)^{-1}X^T)^T=X(X^TX)^{-T}X^T=X(X^TX)^{-1}X^T$
-\ssn{d}
-$HH=X(X^TX)^{-1}X^TX(X^TX)^{-1}X^T=X(X^TX)^{-1}X^T=H$
-\ssn{e}
-By (d), $H$ is a projection matrix.
-Thus, we just need to show that the image of $H$ is the column space.
-Since the image of $X$ is the column space (a subset of $\rn^n$), this means that we need to show that $(X^TX)^{-1}X^T$ has rank $d$ (or that its image is all of $\rn^d$).
-This is true because $(X^TX)^{-1}$ is invertible and $X^T$ has rank $d$ also.
-\ssn{f}
-Traces of products are invariant under cyclic permutation, so $\tr(H)=\tr(X(X^TX)^{-1}X^T)=\tr(X^TX(X^TX)^{-1})=\tr(I_d)=d$.
-\subsection*{3}
-\ssn{b}
-I have a general policy against running unknown scripts on my computer, and since I know that I have a large part of the Python toolchain set up already, I decided to read the setup script (\verb|toolkit.sh|) and replicate the necessary parts.
-
-The first difficulty came when trying to install the packages.
-I ran into a syntax error while installing czipfile via pip, and this turned out to be because my system defaults to Python 3.
-Deleting the virtualenv and re-running mkvirtualenv with the option \verb|--python=/usr/bin/python2.7| fixed the issue.
-
-The Python version problem came up again when creating the bin/notebook file. I modified the line to start with \verb|python2.7 /usr/bin/ipython| to force the script to use 
 \end{document}

### 2014-04-16 stat246/hw2.tex
c2

@@
 \[\left(\frac{\prod_{i=1}^NX_i^{\alpha-1}}{\Gamma(\alpha)^N}\cdot\frac{b^a}{\Gamma(a)}\right)\theta^{N\alpha+a-1}e^{-\theta\sum_{i=1}^NX_i-b\theta}\]
-Up to some normalizing constant that absorbs all the stuff in the front in the parens, this is another gamma distribution with shape $a+N\alpha$ and rate $b+\sum X_i$, so the posterior mean of $\theta$ is $\frac{a+N\alpha}{b+\sum X_i}$, which means that the posterior mean of $\beta$ is the reciprocal of this.
+Up to some normalizing constant that absorbs all the stuff in the front in the parens, this is another gamma distribution with shape $a+N\alpha$ and rate $b+\sum X_i$.
+Thus, $\beta=1/\theta$ has an inverse-gamma distribution with shape $a+N\alpha$ and scale $b+\sum X_i$, which means that the posterior mean of $\beta$ is $\frac{b+\sum X_i}{a+N\alpha-1}$
 

### 2014-04-19 stat246/hw3.tex
s3

\documentclass{article}
\usepackage{geometry}
\usepackage[namelimits,sumlimits]{amsmath}
\usepackage{amssymb,amsfonts}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage[cm]{fullpage}
\newcommand{\tab}{\hspace*{5em}}
\newcommand{\conj}{\overline}
\newcommand{\dd}{\partial}
\newcommand{\ep}{\epsilon}
\newcommand{\openm}{\begin{pmatrix}}
\newcommand{\closem}{\end{pmatrix}}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\Null}{null}
\newcommand{\nc}{\newcommand}
\newcommand{\rn}{\mathbb{R}}
\nc{\cn}{\mathbb{C}}
\nc{\ssn}[1]{\subsubsection*{#1}}
\nc{\inner}[2]{\langle #1,#2\rangle}
\nc{\h}[1]{\widehat{#1}}
\nc{\tl}[1]{\widetilde{#1}}
\nc{\norm}[1]{\left\|{#1}\right\|}
\begin{document}

Name: Hall Liu

Date: \today 
\vspace{1.5cm}
\subsection*{1}
\ssn{a}
Let $Y=1,\ldots,M$ with $P(Y=m)=\pi_m$, and let $(X|Y=m)\sim N(\mu_m, \sigma_m^2)$. Then, the joint distribution of $X$ and $Y$ is the product of these,
\[f_{X,Y}(x,y)=\phi(x;\mu_y,\sigma_y^2)\pi_y\]
To get the marginal density of $X$ we sum over $Y$, which gives
\[\sum_{m=1}^M\phi(x;\mu_m,\sigma_m^2)\pi_m\]
where $\phi$ is the normal pdf, so this gives us the density for the mixture model.
\ssn{b}
The likelihood for a sample $(x_i,y_i)$ is 
\[\prod_{i=1}^N\phi(x_i;\mu_{y_i}, \sigma_{y_i}^2)\pi(y_i)=\frac{1}{(2\pi)^{N/2}}\prod_{i=1}^N\frac{1}{\sigma_{y_i}}\exp\left(\frac{(x_i-\mu_{y_i})^2}{2\sigma_{y_i}^2}\right)\pi(y_i)\]
whose logarithm is 
\[-\frac{N}{2}\log(2\pi)\sum_{i=1}^N\left(-\log(\sigma_{y_i})+\log\pi(y_i)+\frac{(x_i-\mu_{y_i})^2}{2\sigma_{y_i}^2}\right)\]
Now, separate it by the value of the $Y_i$, letting $N_m$ be the number of observations with $Y=m$. Doing this and discarding the constant factor in the front, we get
\[\sum_{m=1}^MN_m\left(\log\pi_m-\log\sigma_m\right)+\sum_{i=1}^{N_m}\frac{(x_i-\mu_m)^2}{2\sigma_m^2}\]
Differentating wrt $\pi_m$ and setting equal to $0$, we obtain $\h{\pi_m}=N_m$, which after normalizing so that their sum is $1$, becomes $\h{\pi_m}=\frac{N_m}{N}$. Then, what we obtain is
\[\sum_{m=1}^M\sum_{i=1}^{N_m}\frac{(x_i-\mu_m)^2}{2\sigma_m^2}-\log\sigma_m+\h{\pi_m}\]
Then, we just have a collection of $M$ unrelated normal log-likelihoods to maximize, so the MLE of the means and the variances are just the sample mean and sample variance within the classes.
\ssn{c}

\end{document}

### 2014-04-20 stat246/hw3.tex
s3

@@
 \ssn{c}
+The likelihood of the data $X_i$ (ignoring the constant multipliers that don't matter for maximization) is
+\[\prod_{i=1}^N\frac{\pi_1}{\sigma_1}\exp\left(-\frac{(X_i-\mu_1)^2}{2\sigma_1^2}\right)+\frac{\pi_2}{\sigma_2}\exp\left(-\frac{(X_i-\mu_2)^2}{2\sigma_2^2}\right)\]
+and taking the log gives
+\[\sum_{i=1}^N\log\left(\frac{\pi_1}{\sigma_1}\exp\left(-\frac{(X_i-\mu_1)^2}{2\sigma_1^2}\right)+\frac{\pi_2}{\sigma_2}\exp\left(-\frac{(X_i-\mu_2)^2}{2\sigma_2^2}\right)\right)\]
+If we plug in the given estimates for the parameters apart from the $\sigma$s, and assume that $\sigma_2$ is fixed, then examine the behaviour of the first term inside the log for each $i$ as $\h{\sigma}_1\to0$.
+If $i\neq1$, then the expression inside the exponential will decay to $0$ extremely quickly, dominating the $\frac{1}{\sigma_1}$ term outside the exponential.
+However, for $i=1$, the term inside the exponential is zero, which means that the first term becomes $\frac{\pi_1}{\sigma_1}$, and this will go to infinity as $\sigma_1\to0$.
+Thus, the whole log-likelihood will also diverge to infinity.
+This means that if we try to use the EM algorithm on this and one of the estimates of the means ends up falling on one of the data points exactly, it'll start to diverge because the likelihood has no global maximum.
+\ssn{d}
+For the $i\neq1$ terms, setting $\sigma_1=\sigma_2=\sigma$ and letting that go to zero in the log-likelihood produces $N-1$ terms that asymptotically behaves like $\log\left(\frac{1}{\sigma}\exp(-\sigma^{-2})\right)$, and the $i=1$ term still behaves like $\log\sigma^{-1}$.
+Adding these together results in the exponential terms decaying much faster thatn the $\sigma^{-1}$ terms, which means that the log-likelihood will go to $-\infty$ instead, avoiding the problem.
+\ssn{e}
+Continuing from (b), after estimating the $\pi_m$, we obtain 
+\[\sum_{m=1}^MN_m\left(\log\h{\pi}_m-\log\sigma_m\right)+\sum_{i=1}^{N_m}\frac{(x_i-\mu_m)^2}{2\sigma_m^2}\]
+for the log-likelihood. Differentiating wrt $\mu_1$ and setting equal to zero, we have that $\h{\mu}_m$ is the sample mean of the $X_i$ with corresponding $Y_i$ equal to $m$, for $m=1,2$.
+
+Finally, note that this is now the same as the $2$-class classification problem whose MLE we computed in problem 3 of the last homework.
+From there, we can conclude that the MLE of the covariance is the weighted average of the sample covariances, with
+\ssn{f}
+Begin with initial estimates $\mu_1^{(0)}$, $\mu_2^{(0)}$, $\sigma^{(0)}$, $\pi_1^{(0)}$, and $\pi_2^{(0)}$.
+Collectively, call these parameters $\eta$.
+In the $k$th step, let $w_{n, m}=f(Y=m|X_n;\eta^{(k-1)})$ for $m=1,2$.
+Estimate $\pi_m^{(k)}$ by taking $\frac{1}{N}\sum_{i=1}^Nw_{m, i}$.
+Then, set
+\begin{align*}
+    \mu_1^{(k)}&=\frac{\frac{1}{N}\sum_{i=1}^Nw_{1,i}X^{(i)}}{\sum_{i=1}^Nw_{1, i}}\\
+    \mu_2^{(k)}&=\frac{\frac{1}{N}\sum_{i=1}^Nw_{2,i}X^{(i)}}{\sum_{i=1}^Nw_{2, i}}\\
+    \sigma^{(k)}&=\pi_1^{(k)}\frac{
 

### 2014-04-21 stat246/hw3.tex
s2

@@
 The likelihood for a sample $(x_i,y_i)$ is 
-\[\prod_{i=1}^N\phi(x_i;\mu_{y_i}, \sigma_{y_i}^2)\pi(y_i)=\frac{1}{(2\pi)^{N/2}}\prod_{i=1}^N\frac{1}{\sigma_{y_i}}\exp\left(\frac{(x_i-\mu_{y_i})^2}{2\sigma_{y_i}^2}\right)\pi(y_i)\]
+\[\prod_{i=1}^N\phi(x_i;\mu_{y_i}, \sigma_{y_i}^2)\pi(y_i)=\frac{1}{(2\pi)^{N/2}}\prod_{i=1}^N\frac{1}{\sigma_{y_i}}\exp\left(-\frac{(x_i-\mu_{y_i})^2}{2\sigma_{y_i}^2}\right)\pi(y_i)\]
 whose logarithm is 
-\[-\frac{N}{2}\log(2\pi)\sum_{i=1}^N\left(-\log(\sigma_{y_i})+\log\pi(y_i)+\frac{(x_i-\mu_{y_i})^2}{2\sigma_{y_i}^2}\right)\]
+\[-\frac{N}{2}\log(2\pi)\sum_{i=1}^N\left(-\log(\sigma_{y_i})+\log\pi(y_i)-\frac{(x_i-\mu_{y_i})^2}{2\sigma_{y_i}^2}\right)\]
 Now, separate it by the value of the $Y_i$, letting $N_m$ be the number of observations with $Y=m$. Doing this and discarding the constant factor in the front, we get
-\[\sum_{m=1}^MN_m\left(\log\pi_m-\log\sigma_m\right)+\sum_{i=1}^{N_m}\frac{(x_i-\mu_m)^2}{2\sigma_m^2}\]
+\[\sum_{m=1}^MN_m\left(\log\pi_m-\log\sigma_m\right)-\sum_{i=1}^{N_m}\frac{(x_i-\mu_m)^2}{2\sigma_m^2}\]
 Differentating wrt $\pi_m$ and setting equal to $0$, we obtain $\h{\pi_m}=N_m$, which after normalizing so that their sum is $1$, becomes $\h{\pi_m}=\frac{N_m}{N}$. Then, what we obtain is
-\[\sum_{m=1}^M\sum_{i=1}^{N_m}\frac{(x_i-\mu_m)^2}{2\sigma_m^2}-\log\sigma_m+\h{\pi_m}\]
+\[\sum_{m=1}^M\sum_{i=1}^{N_m}-\frac{(x_i-\mu_m)^2}{2\sigma_m^2}-\log\sigma_m+\h{\pi_m}\]
 Then, we just have a collection of $M$ unrelated normal log-likelihoods to maximize, so the MLE of the means and the variances are just the sample mean and sample variance within the classes.
@@
 Continuing from (b), after estimating the $\pi_m$, we obtain 
-\[\sum_{m=1}^MN_m\left(\log\h{\pi}_m-\log\sigma_m\right)+\sum_{i=1}^{N_m}\frac{(x_i-\mu_m)^2}{2\sigma_m^2}\]
+\[\sum_{m=1}^2N_m\left(\log\h{\pi}_m-\log\sigma_m\right)-\sum_{i=1}^{N_m}\frac{(x_i-\mu_m)^2}{2\sigma_m^2}\]
 for the log-likelihood. Differentiating wrt $\mu_1$ and setting equal to zero, we have that $\h{\mu}_m$ is the sample mean of the $X_i$ with corresponding $Y_i$ equal to $m$, for $m=1,2$.
@@
 In the $k$th step, let $w_{n, m}=f(Y=m|X_n;\eta^{(k-1)})$ for $m=1,2$.
-Estimate $\pi_m^{(k)}$ by taking $\frac{1}{N}\sum_{i=1}^Nw_{m, i}$.
-Then, set
-\begin{align*}
-    \mu_1^{(k)}&=\frac{\frac{1}{N}\sum_{i=1}^Nw_{1,i}X^{(i)}}{\sum_{i=1}^Nw_{1, i}}\\
-    \mu_2^{(k)}&=\frac{\frac{1}{N}\sum_{i=1}^Nw_{2,i}X^{(i)}}{\sum_{i=1}^Nw_{2, i}}\\
-    \sigma^{(k)}&=\pi_1^{(k)}\frac{
+Then, our goal becomes to maximize $E_Y(\log L(\mu_1,\mu_2,\sigma,\pi_1,\pi_2;X,Y))$.
+
+This log-likelihood, ignoring constant multipliers in the front, is 
+\[\sum_{i=1}^N\left(-\log(\sigma)+\log\pi(y_i)-\frac{(x_i-\mu_{y_i})^2}{2\sigma^2}\right)\]
+which we can rewrite as 
+\[\sum_{i=1}^N\sum_{m=1}^2\mathbb{I}(y_i=m)\left(-\log\sigma+\log\pi_m-\frac{(x_i-\mu_m)^2}{2\sigma^2}\right)\]
+where $\mathbb{I}$ is the indicatior function. The expectation of $\mathbb{I}(y_i=m)$ given the data $X$ and the previous parameters is $w_{i,m}$, so the expectation of the log-likelihood is
+\[\sum_{i=1}^N\sum_{m=1}^2w_{i,m}\left(-\log\sigma+\log\pi_m-\frac{(x_i-\mu_m)^2}{2\sigma^2}\right)\]
+Differentiating wrt $\pi_m$ gives 
+\[\sum_{i=1}^N\frac{w_{i,m}}{\pi_m}\]
+so the new estimate for $\pi_m$ is $\sum_{i=1}^Nw_{i,m}$.
+
+Differentiating wrt $\mu_m$ gives
+\[\sum_{i=1}^N-w_{i,m}\frac{-(x_i-\mu_m)}{\sigma^2}\]
+and setting this equal to zero produces
+\[\h{\mu}_m=\frac{\sum_{i=1}^Nw_{i,m}x_i}{\sum_{i=1}^Nw_{i,m}}\]
+Finally, differentiating wrt $\sigma$ produces
+\[\sum_{i=1}^N\sum_{m=1}^2w_{i,m}\left(\frac{(x_i-\mu_m)^2}{\sigma^3}-\frac{1}{\sigma}\right)\]
+Setting this equal to $0$ produces
+\[\h{\sigma}=\frac{\sum_{i=1}^N\sum_{m=1}^2w_{i,m}(x_i-\h{\mu}_m)^2}{\sum_{i=1}^N\sum_{m=1}^2w_{i,m}}\]
+\ssn{g}
+Since this is an iterative algorithm, we could set limits on the estimate of the likelihood and the covariance, then just interrupt the algorithm when those limits are exceeded.
+If the mean starts to move too close to a single data point, the covariance will tend to go to zero because that increases the likelihood.
+Resetting the mean to some other value and putting the covariance at some reasonable value would interrupt the cycle.
+\subsection*{2}
+\ssn{a}
+If $\rho$ is the permutation defined by $(x_1,x_2,\ldots,x_n)\mapsto(x_{m+1},\ldots,x_n,x_1,\ldots,x_m)$, then the inverse of $\rho$ maps $(x_1,\ldots,x_n)$ to $(x_{n-m+1},\ldots,x_n,x_1,\ldots,x_{n-m})$.
+For the rest of this problem, let $\rho\equiv\rho^{(1)}$ be the generator of this cyclic subgroup.
+\ssn{b}
+The joint distribution of $X,Y$ is $f(x,y)=\pi_yf(\rho^{-y}(x);\theta)$. Given data points $(x_1,y_1),\ldots,(x_N,y_N)$, the likelihood is
+\[\prod_{i=1}^N\pi(y_i)f(\rho^{-y_i}(x);\theta)=\prod_{i=1}^N\pi(y_i)\prod_{j=1}^df_j(x_{\rho^{-y_i}(j)};\theta_j)\]
+where $\rho(j)$ for some $j\in[1..d]$ is the usual action of the permutations on that set. Taking the log gives
+\[\sum_{i=1}^N\left(\log\pi(y_i)+\sum_{j=1}^d\log f_j(x_{\rho^{-y_i}(j)};\theta_j)\right)\]
+\ssn{c}
+Differentiating wrt $\theta_j$ and setting equal to zero, we have
+\[\sum_{i=1}^N\frac{f_j'(x_{\rho^{-y_i}(j)};\theta_j)}{f_j(x_{\rho^{-y_i}(j)};\theta_j)}=0\]
+which we cannot simplify further without knowledge of the form of $f$.
+\ssn{d}
+This sort of model has less free parameters than a general mixture, which would help to prevent overfitting.
+Also, it might reveal some sort of natural symmetry in the data, if we find that all the $\pi_y$ are similar.
+\ssn{f}
+For a Poisson distribution, we have $f(x;\theta)=\frac{\theta^xe^{-\theta}}{x!}$, so the log-likelihood becomes
+\[\sum_{i=1}^N\left(\log\pi_{y_i}+\sum_{j=1}^d\log \frac{\theta_j^{x_{\rho^{-y_i}(j)}}e^{-\theta_j}}{x_{\rho^{-y_i}(j)}!}\right)
+=\sum_{i=1}^N\left(\log\pi_{y_i}+\sum_{j=1}^dx_{\rho^{-y_i}(j)}\log\theta_j-\theta_j-\log(x_{\rho^{-y_i}(j)}!)\right)\]
+Writing this in terms of the indicator function on $y$, it becomes
+\[\sum_{i=1}^N\sum_{m=1}^d\mathbb{I}(y_i=m)\left(\log\pi_m+\sum_{j=1}^dx_{\rho^{-m}(j)}\log\theta_j-\theta_j-\log(x_{\rho^{-m}(j)})\right)\]
+The expectation of $\mathbb{I}(y_i=m)$ is $w_{i,m}=P(Y=m|X,\theta)$, which can be computed by plugging into the joint distribution. Then, the expectation of the log-likelihood becomes
+
+\[\sum_{i=1}^N\sum_{m=1}^dw_{i,m}\left(\log\pi_m+\sum_{j=1}^dx_{\rho^{-m}(j)}\log\theta_j-\theta_j-\log(x_{\rho^{-m}(j)})\right)\]
+Differentiating wrt $\pi_m$ and setting to zero gives $\h{\pi}_m=\frac{1}{N}\sum_{i=1}^Nw_{i,m}$. Differentiating wrt $\theta_j$ produces
+\[\sum_{i=1}^N\sum_{m=1}^d\left(w_{i,m}\frac{x_{\rho^{-m}(j)}}{\theta_j}-1\right)=\theta_j\sum_{i=1}^N\sum_{m=1}^dw_{i,m}x_{\rho^{-m}(j)}-Nd\]
+so setting this to zero produces $\theta_j=\frac{Nd}{\sum_{i=1}^N\sum_{m=1}^dw_{i,m}x_{\rho^{-m}(j)}}$
+%TODO: computation
+\subsection*{3}
 

### 2014-04-26 stat246/hw3.tex
s3
s2

@@
 \usepackage{graphicx}
+\usepackage{mathrsfs}
 \usepackage[cm]{fullpage}
@@
 \nc{\tl}[1]{\widetilde{#1}}
+\nc{\kl}{\mathcal{K}}
 \nc{\norm}[1]{\left\|{#1}\right\|}
@@
 \subsection*{3}
+\ssn{a}
+We have 
+\[\kl(X\|U)=\sum_{i=1}^KP(X=k)\log\left(\frac{K^{-1}}{P(X=k)}\right)=-\sum_{i=1}^KP(X=k)\log(KP(X=k))=-\sum_{i=1}^KP(X=k)\log P(X=k)-\log K\]
+so $\log K-\kl(X\|U)=\sum_{i=1}^KP(X=k)\log P(X=k)=H(X)$.
 
+We know that $\kl(X\|U)\geq0$, with equality when $X=U$, so the upper bound for $H(X)$ is $\log K$.
+A lower bound for the entropy is $0$, since the log term will always be non-positive and the probability term will always be non-negative.
+This is a tight bound, since if we let $P(X=1)\to1$ and the other probabilities tend towards zero, since $\lim_{x\to0}\log x^x=0$ (for the $k\neq1$ terms) and $\lim_{x\to1}\log x^x=0$ (for the $k=1$ term).
+\ssn{b,c}
+\begin{align*}
+    H(X,Y)&=-\sum_{i=1}^K\sum_{j=1}^KP(i,j)\log P(i,j)\\
+          &=-\sum_{i=1}^K\sum_{j=1}^KP(i|j)P(j)\log \left(P(i|j)P(j)\right)\\
+          &=-\sum_{j=1}^KP(j)\sum_{i=1}^KP(i|j)\left(\log P(i|j) + \log P(j)\right)\\
+          &=-\sum_{j=1}^KP(j)\left(\sum_{i=1}^KP(i|j)\log P(i|j)+\log P(j)\sum_{i=1}^KP(i|j)\right)\\
+          &=H(X|Y)-\sum_{j=1}^KP(j)\log P(j)\\
+          &=H(X|Y)+H(Y)\\
+\end{align*}
+Huh. Looks like I'm doing this backwards then. Writing $H(X|Y)$ as $\sum_{i=1}^K\sum_{j=1}^KP(i,j)\log \P(i|j)$, we have 
+\begin{align*}
+    H(X|Y)-H(X)&=\sum_{i=1}^KP(i)\log P(i)-\sum_{i=1}^K\sum_{j=1}^KP(i,j)\log P(i|j)\\
+               &=\sum_{i=1}^K \left(P(i)\log P(i)-\sum_{j=1}^KP(i,j)\log P(i|j)\right)\\
+               &=\sum_{i=1}^K\sum_{j=1}^KP(i,j)\log\frac{P(i)}{P(i|j)}\\
+               &\leq\log\left(\sum_{i=1}^K\sum_{j=1}^K\frac{P(i,j)P(i)}{P(i|j)}\ \text{by Jensen's ineq}\\
+               &=\log\left(\sum_{i=1}^K\sum_{j=1}^KP(i)P(j)\right)\\
+               &=0
+\end{align*}
+so we have $H(X|Y)\leq H(X)$, with equality only when $\frac{P(i)}{P(i|j)}$ are equal for every $i,j$.
+This only happens when $P(i|j)=P(i)$ for all $i,j$ ($X$ and $Y$ independent), since otherwise we'd have some $i$ for which there exists $j_1$ and $j_2$ such that $P(i|j_1)>P(i)$ and $P(i|j_2)<P(i)$.
+Then, it follows that $H(X,Y)=H(X|Y)+H(Y)\leq H(X)+H(Y)$ with equality iff $X$ and $Y$ are independent.
+
+Intuitively, $H(X|Y)$ is the randomness of $X$ when we are given information about $Y$.
+We can't have the information make $X$ any more random than it already is, but it can be and is completely useless when $X$ and $Y$ are independent.
+Similarly, the randomness of two RVs, considered together, can't be any more than how random they each are added together, but it's the same if the two RVs don't carry information about each other.
+The inequality is a result of redundant information.
 \end{document}

### 2014-04-26 stat246/hw3.tex
s3

@@
                &=\sum_{i=1}^K\sum_{j=1}^KP(i,j)\log\frac{P(i)}{P(i|j)}\\
-               &\leq\log\left(\sum_{i=1}^K\sum_{j=1}^K\frac{P(i,j)P(i)}{P(i|j)}\ \text{by Jensen's ineq}\\
+               &\leq\log\left(\sum_{i=1}^K\sum_{j=1}^K\frac{P(i,j)P(i)}{P(i|j)}\right)\ \text{by Jensen's ineq}\\
                &=\log\left(\sum_{i=1}^K\sum_{j=1}^KP(i)P(j)\right)\\

### 2014-04-27 stat246/hw3.tex
s3

@@
 Differentiating wrt $\pi_m$ and setting to zero gives $\h{\pi}_m=\frac{1}{N}\sum_{i=1}^Nw_{i,m}$. Differentiating wrt $\theta_j$ produces
-\[\sum_{i=1}^N\sum_{m=1}^d\left(w_{i,m}\frac{x_{\rho^{-m}(j)}}{\theta_j}-1\right)=\theta_j\sum_{i=1}^N\sum_{m=1}^dw_{i,m}x_{\rho^{-m}(j)}-Nd\]
-so setting this to zero produces $\theta_j=\frac{Nd}{\sum_{i=1}^N\sum_{m=1}^dw_{i,m}x_{\rho^{-m}(j)}}$
-%TODO: computation
+\[\sum_{i=1}^N\sum_{m=1}^d\left(w_{i,m}\frac{x_{\rho^{-m}(j)}}{\theta_j}-w_{i,m}\right)=\theta_j^{-1}\sum_{i=1}^N\sum_{m=1}^dw_{i,m}x_{\rho^{-m}(j)}-N\]
+so setting this to zero produces $\theta_j=N^{-1}\sum_{i=1}^N\sum_{m=1}^dw_{i,m}x_{\rho^{-m}(j)}$, where there's an implicit $i$ subscript on the $x$
+
+Below are index plots of $\theta$ and $\pi$.
+The first row is for $N=10$, descending to $N=20$ and $N=200$, and the true values are at the bottom.
+As might be expected, some of the results differ from the true value by a cyclic permutation, since the model is invariant under such permutations.
+The $\theta$s are on the left, the $\pi$s on the right.
+
+\includegraphics[width=0.5\textwidth]{hw3_files/theta_10.png}
+\includegraphics[width=0.5\textwidth]{hw3_files/pi_10.png}
+
+\includegraphics[width=0.5\textwidth]{hw3_files/theta_20.png}
+\includegraphics[width=0.5\textwidth]{hw3_files/pi_20.png}
+
+\includegraphics[width=0.5\textwidth]{hw3_files/theta_200.png}
+\includegraphics[width=0.5\textwidth]{hw3_files/pi_200.png}
+
+\includegraphics[width=0.5\textwidth]{hw3_files/true_theta.png}
+\includegraphics[width=0.5\textwidth]{hw3_files/true_pi.png}
+
 \subsection*{3}

### 2014-04-27 stat246/hw3_files/p2.py
s4

@@ -1,0 +1,49 @@
+import numpy as np
+import scipy.stats as st
+'''
+data is a N by d matrix; each row is an observation
+'''
+def poisson_em(data, epsilon, pis=None, thetas=None):
+    N, d = data.shape
+    if pis is None:
+        pis = np.random.rand(d)
+        pis /= pis.sum()
+    if thetas is None:
+        thetas = np.random.rand(d) * 2
+
+    while True:
+        weight_matrix = compute_weights(data, thetas, pis)
+
+        new_pis = weight_matrix.sum(0) / N
+
+        new_thetas = np.zeros(d)
+        for j in range(d):
+            data_permuted = np.roll(data, -j, axis=1)
+            new_thetas[j] = (weight_matrix * data_permuted).sum() / N 
+
+        if np.linalg.norm(new_pis - pis) < epsilon and np.linalg.norm(new_thetas - thetas) < epsilon:
+            return (thetas, pis)
+
+        print np.linalg.norm(new_thetas - thetas)
+        pis = new_pis
+        thetas = new_thetas
+
+'''
+Returns the weights in the same shape as data
+'''
+def compute_weights(data, theta, pis):
+    N, d = data.shape
+    p_fn = st.poisson(theta).pmf
+
+    weights = np.zeros((N, d))
+    for m in range(d):
+        data_permuted = np.roll(data, -m, axis=1)
+        for n in range(N):
+            weights[n, m] = np.log(p_fn(data_permuted[n])).sum() + np.log(pis[m])
+
+    max_alphas = np.amax(weights, axis=1)
+    weights -= max_alphas[:, np.newaxis]
+    weights = np.exp(weights)
+    w_sums = weights.sum(1)
+    weights /= w_sums[:, np.newaxis]
+    return weights

### 2014-04-28 cs233/om.py
hw1

@@ -1,0 +1,145 @@
+import sys
+
+class General():
+    def __init__(self, index, loyalty):
+        self.id = index
+        if loyalty == 'L':
+            self.loyal = True
+        else:
+            self.loyal = False
+
+        self.messages_rcvd = []
+
+    def send_to(self, lts, message):
+        new_msg_header = message[0] + [self.id]
+        if self.loyal:
+            new_message = (new_msg_header, message[1])
+            map(lambda x: x.messages_rcvd.append(new_message), lts)
+        else:
+            new_message_1 = (new_msg_header, message[1])
+            new_message_2 = (new_msg_header, not message[1])
+            for l in lts:
+                if l.id % 2 == 1:
+                    l.messages_rcvd.append(new_message_1)
+                else:
+                    l.messages_rcvd.append(new_message_2)
+
+def run_om_instance(line):
+    components = line.split()
+    m = int(components[0])
+    if m == 0:
+        return None
+    
+    n_generals = len(components[1])
+    init_command = (components[2] == 'ATTACK')
+
+    all_generals = map(lambda x: General(*x), enumerate(components[1]))
+    
+    om_message(all_generals[0], all_generals[1:], ([], init_command), m)
+    msg_trees = map(lambda x: make_msg_tree(x.messages_rcvd, n_generals, m, 0), all_generals[1:])
+
+    for s in msg_trees:
+        i = 1
+        sys.stdout.write(s[1] + ' ')
+        q = False # nasty formatting hack
+        for s_tree in s[2]:
+            if i != s_tree[0]:
+                sys.stdout.write(' ')
+                i += 1
+                q = True
+            sys.stdout.write(s_tree[3])
+            i += 1
+
+        sys.stdout.write(' ')
+        if not q:
+            sys.stdout.write(' ')
+
+        if s[3] == 'A':
+            sys.stdout.write('ATTACK\n')
+        elif s[3] == 'R':
+            sys.stdout.write('RETREAT\n')
+        else:
+            sys.stdout.write('TIE\n')
+
+    return msg_trees 
+        
+
+'''
+Performs the recursive message sending on the list of generals.
+'''
+def om_message(cmdr, lts, message, m):
+    cmdr.send_to(lts, message)
+    
+    if m == 0:
+        return
+    
+    '''
+    Perform the recursive step here, depending on whether the current
+    commander is loyal. Either case, we know what message he sent, so
[…]

### 2014-04-28 cs233/om.py
formatting

@@ -50,7 +50,7 @@ def run_om_instance(line):
             sys.stdout.write(s_tree[3])
             i += 1
 
-        sys.stdout.write(' ')
+        sys.stdout.write('  ')
         if not q:
             sys.stdout.write(' ')
 

### 2014-04-28 cs233/hw1/run.sh
hw1

@@ -1,0 +1,2 @@
+#!/bin/bash
+python2 om.py

### 2014-05-01 stat246/hw4.tex
s4

\documentclass{article}
\usepackage{geometry}
\usepackage[namelimits,sumlimits]{amsmath}
\usepackage{amssymb,amsfonts}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage[cm]{fullpage}
\usepackage{mathrsfs}
\newcommand{\tab}{\hspace*{5em}}
\newcommand{\conj}{\overline}
\newcommand{\dd}{\partial}
\newcommand{\ep}{\epsilon}
\newcommand{\openm}{\begin{pmatrix}}
\newcommand{\closem}{\end{pmatrix}}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\Null}{null}
\newcommand{\nc}{\newcommand}
\newcommand{\rn}{\mathbb{R}}
\nc{\cn}{\mathbb{C}}
\nc{\ssn}[1]{\subsubsection*{#1}}
\nc{\inner}[2]{\langle #1,#2\rangle}
\nc{\h}[1]{\widehat{#1}}
\nc{\tl}[1]{\widetilde{#1}}
\nc{\norm}[1]{\left\|{#1}\right\|}
\begin{document}

Name: Hall Liu

Date: \today 
\vspace{1.5cm}
\subsection*{1}
\ssn{a}
The value to be minimized is the negative of the entropy, or
\[\sum_{i=1}^K p_i\log p_i\]
Our equality constraints are those imposed by the expectations and that the probabilities must sum to $1$. Define $T_0(x)=1$ for all $x$ and $\alpha_0=1$, so then we can write for $r=0..R$, we have the constraint 
\[\alpha_r=\sum_{i=1}^KT_r(i)p_i\]
which includes the sum constraint. Writing the $T$s as a vector in $\rn^K$, we can express this in the form $\alpha_r-p^TT_r=0$.

Finally, we have an inequality constraint, which is that all probabilities must be nonnegative, or $p^Te_i\geq0$ for all $i=1..K$, where $e_i$ are the standard basis vectors
\ssn{b}
The Lagrangian for the system above is 
\[\sum_{i=1}^K p_i\log p_i-\sum_{r=0}^R\lambda_r(\alpha_r-p^TT_r)-\sum_{i=1}^K\mu_ip^Te_i\]
Its gradient wrt $p$ is 
\[e+\log p+\sum_{r=0}^R\lambda_rT_r-\sum_{i=1}^K\mu_ie_i\]
where $e$ is the vector of all $1$s and the log is taken componentwise. Setting this equal to $0$ gives
\[\log p=-\sum_{r=0}^R\lambda_rT_r+\mu-e\]
Componentwise, we have $p_i=\exp\left(\mu_i-\sum_{r=0}^R\lambda_r T_r(i)-1\right)$. If we let $u(i)=\openm e_i\\Te_i\closem$ ($T$ being the matrix of values of the $T_r$), then we can write
\[p(i)=e^{-1}\exp\left(\openm \mu^T &\lambda^T\closem u(i)\right)\]
which is in the form of an exponential family, with the Lagrange multiplies $\mu_i$ and $\lambda_i$ being the parameters.
\ssn{c}
If we look at $T$ as data rather than imposed constraints, then we can view $u$ as a function of $T$. The moment constraints $\alpha_i$ then become sufficient statistics, and we can compute the Lagrange multipliers as a ML problem rather than a constrained optimization problem.
\subsection*{2}
\ssn{a}
The joint distribution of $Y$ and $Z$ is bivariate normal with mean $\openm\mu\\0\closem$ and covariance $\openm \sigma^2&0\\0&\sigma_1^2\closem$. We can obtain the joint distribution of $X=Y+Z$ and $Y$ from this by transforming via the matrix $\openm 1&0\\1&1\closem$, which results in a bivariate normal with mean $\openm\mu\\\mu\closem$ and covariance $\openm\sigma^2&\sigma^2\\\sigma^2&\sigma^2+\sigma_1^2\closem$ (this is arranged as $\openm Y\\X\closem$).
\ssn{b}
If we condition on $X=x$, then the density is univariate normal, with mean 
\[\mu+\frac{\sigma}{\sqrt{\sigma^2+\sigma_1^2}}\frac{\sigma^2}{\sigma\sqrt{\sigma^2+\sigma_1^2}}(x-\mu)=\mu+\frac{\sigma^2}{\sigma^2+\sigma_1^2}(x-\mu)\]
and variance $\left(1-\frac{\sigma^2}{\sigma^2+\sigma_1^2}\right)\sigma^2$
\ssn{c}
The marginal distribution of $X$ is normal with mean $\mu$ and variance $\sigma^2+\sigma_1^2$. Then, the ML estimate for $\mu$ is the sample mean of the observed $X^{(n)}$, and the ML estimate for $\sigma^2$ is the sample covariance minus $\sigma_1^2$.
\ssn{d}
First, we can write the joint density of $X$ and $Y$ as $f(X|Y=y)f(y)$. Since we have the identity $X=Y+Z$, $X|Y=y$ is normal with mean $y$ and variance $\sigma_1^2$. Thus, the log-likelihood of the complete data is 
\[\sum_{i=1}^N\log f(x_i|Y=y_i)+\log f(y_i)=\sum_{i=1}^N-\log(2\pi)-\log\sigma_1-\log\sigma'-\frac{(x_i-y_i)^2}{2\sigma_1^2}-\frac{(y_i-\mu')^2}{2\sigma'^2}\]
where the $\mu'$ and $\sigma'$ are yet-to-be determined. We can discard the constant terms then take the expectation wrt to $Y$ given $X$ to obtain
\[-\sum_{i=1}^N\int_\rn p(y|x_i,\mu,\sigma^2)\left(\frac{(x_i-y)^2}{2\sigma_1^2}+\frac{(y-\mu')^2}{2\sigma'^2}+\log\sigma'\right)\ dy\]
$p(y|x_i, \mu, \sigma^2)$ is a quantity calculated from the parameters of the previous iteration. As derived in (b), it is the normal density with the appropriate mean and variance (written above, not going to rewrite here). Abbreviate this function as $q_i(y)$ from now on.

To find the next value of $\mu$, we differentiate wrt $\mu'$ and set to 0 to obtain
\begin{align*}
    0&=\sum_{i=1}^N\int_\rn \frac{1}{\sigma'^2}q_i(y)(y-\mu')\ dy\\
    \mu'\sum_{i=1}^N\int_{\rn}q_i(y)\ dy&=\sum_{i=1}^N\int_\rn yq_i(y)\ dy\\
    \mu'&=\frac{1}{N}\sum_{i=1}^NE(Y|x_i,\mu,\sigma^2)\\
        &=\mu+\frac{1}{N}\frac{\sigma^2}{\sigma^2+\sigma_1^2}\sum_{i=1}^N(x_i-\mu)\\
\end{align*}

Now, differentiating wrt $\sigma'$ and setting to $0$ gives
\begin{align*}
    0&=\sum_{i=1}^N\int_\rn q_i(y)\left(\frac{(y-\mu')^2}{\sigma'^3}-\frac{1}{\sigma'}\right)\,dy\\
    N&=\frac{1}{\sigma'^2}\sum_{i=1}^N\int_\rn q_i(y)(y-\mu')^2\,dy\\
    \sigma'^2&=\frac{1}{N}\sum_{i=1}^NE_{Y|X=x_i}(Y-\mu')^2\\
\end{align*}

The expectation inside the sum expands to $\var(Y)+E(Y)^2-2\mu' E(Y)+\mu'^2$, so the final expression for $\sigma'^2$ is
\[\mu'^2+\frac{1}{N}\sum_{i=1}^N\var(Y|x_i)+E(Y|x_i)^2-2\mu'E(Y|x_i)\]
\ssn{e}
\end{document}

### 2014-05-02 stat246/hw4.tex
s4

@@
     0&=\sum_{i=1}^N\int_\rn q_i(y)\left(\frac{(y-\mu')^2}{\sigma'^3}-\frac{1}{\sigma'}\right)\,dy\\
+     &=\sum_{i=1}^N\int_\rn q_i(y)\left(\frac{(y-\mu')^2}{\sigma'^2}-1\right)\,dy\\
     N&=\frac{1}{\sigma'^2}\sum_{i=1}^N\int_\rn q_i(y)(y-\mu')^2\,dy\\
@@
 \[\mu'^2+\frac{1}{N}\sum_{i=1}^N\var(Y|x_i)+E(Y|x_i)^2-2\mu'E(Y|x_i)\]
+where the dependence on the previous $\sigma$ comes in from the conditional expectations and variance
 \ssn{e}
+If we let $\mu$ be the sample mean of the $X_i$ in the expression for $\mu'$ above, then the sum cancels out to zero and we are left with $\mu'=\mu$, which means that it's a fixed point.
+This is the only fixed point, since the equation obtained by setting $\mu'=\mu$ is linear, which means that there's only one solution.
+\ssn{f}
+If we set $\mu=\mu'=\h{\mu}_{ML}$, then our expression for $\sigma'^2$ becomes
+\[\h{\mu}^2+\left(1-\frac{\sigma^2}{\sigma^2+\sigma_1^2}\right)\sigma^2+\frac{1}{N}\sum_{i=1}^N\left(\h{\mu}+\frac{\sigma^2}{\sigma^2+\sigma_1^2}(x_i-\h{\mu})\right)^2-\frac{2\h{\mu}}{N}\sum_{i=1}^N\h{\mu}+\frac{\sigma^2}{\sigma^2+\sigma_1^2}(x_i-\h{\mu})\]
+The term with the unadorned sum of the $x_i-\h{\mu}$ sums out to zero, so canceling that and expanding the square leaves us with
+\[\left(1-\frac{\sigma^2}{\sigma^2+\sigma_1^2}\right)\sigma^2+\frac{1}{N}\sum_{i=1}^N2\h{\mu}\frac{\sigma^2}{\sigma^2+\sigma_1^2}(x_i-\h{\mu})+\frac{1}{N}\sum_{i=1}^N\left(\frac{\sigma^2}{\sigma^2+\sigma_1^2}(x_i-\h{\mu})\right)^2\]
+The middle term again disappears, and if we're assuming that $\sigma=\h{\sigma}_{ML}$, then we can rewrite the last term:
+\[\sigma^2-\frac{\sigma^4}{\sigma^2+\sigma_1^2}+\frac{\sigma^4(\sigma^2+\sigma_1^2)}{(\sigma^2+\sigma_1^2)^2}=\sigma^2\]
+so we have that the ML estimate is a fixed point.
+\subsection*{3}
+\ssn{a}
+Let $Y_1,\ldots, Y_N$ be the hidden class variables. The complete-data log-likelihood is then
+\[\sum_{i=1}^N\left(\log \pi_{y_i}+\sum_{j=1}^dx_{i,j}\log p_{j,y_i}+(1-x_{i,j})\log(1-p_{j, y_i})\right)\]
+Taking the expectation wrt $Y$ given previous parameters and the data yields
+\[\sum_{i=1}^N\sum_{m=1}^Mw_{i,m}\left(\log\pi_m+\sum_{j=1}^dx_{i,j}\log p_{j,m}+(1-x_{i,j})\log(1-p_{j,m})\right)\]
+The ML estimate of the $\pi_m$ is, as usual, $\frac{1}{N}\sum_{i=1}^Nw_{i,m}$. Differentiating wrt each $p_{j,m}$, we have
+\[\sum_{i=1}^Nw_{i,m}\left(\frac{x_{i,j}}{p_{j,m}}-\frac{1-x_{i,j}}{1-p_{j,m}}\right)\]
+Setting this equal to zero produces 
+\[p_{j,m}=\frac{\sum_{i=1}^Nw_{i,m}x_{i, j}}{\sum_{i=1}^N w_{i,m}}\]
+\ssn{b}
+We want conjugate priors for $\pi$ and $p$.
+Since the conjugate priors for the Bernoulli and the multinomial are the beta and the Dirichlet, respectively, we use $Md$ independent beta densities for the $p_{j,m}$ and a Dirichlet for all the $\pi$s.
+For simplicity's sake, initialize all the beta distributions with the same hyperparameter $\alpha$.
+Let the initial hyperparameters for the Dirichlet be $\zeta_m$.
+
+The quantity to maximize is now $l(\theta|X,Y)+\log p(\theta)$, which in this case is
+\begin{gather*}
+    \sum_{i=1}^N\sum_{m=1}^Mw_{i,m}\left(\log\pi_m+\sum_{j=1}^dx_{i,j}\log p_{j,m}+(1-x_{i,j})\log(1-p_{j,m})\right)\\
+    +\sum_{i=1}^N\sum_{m=1}^M\sum_{j=1}^d\frac{1}{N}\left((\alpha-1)\log p_{j,m}+(\alpha-1)\log(1-p_{j,m})\right)\\
+    +\frac{1}{N}\sum_{i=1}^N\sum_{m=1}^M(\zeta_m-1)\log\pi_m
+\end{gather*}
+after removing constants irrelevant to maximization. Folding the sums in gives
+\[\sum_{i=1}^N\sum_{m=1}^M\left(w_{i,m}+\frac{1}{N}(\zeta_m-1)\right)\log\pi_m+\sum_{i=1}^N\sum_{m=1}^Mw_{i,m}\sum_{j=1}^d\left(x_{i,j}+\frac{1}{N}(\alpha-1)\right)\log p_{j,m}+\left(1-x_{i,j}+\frac{1}{N}(\alpha-1)\right)\log(1-p_{j,m})\]
+Having the same form as the thing we maximized already, we find that the new equations for $\pi_m$ and $p_{j,m}$ are
+\begin{align*}
+    \pi_m&=\frac{1}{N}\sum_{i=1}^N\left(w_{i,m}+\frac{1}{N}(\zeta_m-1)\right)\\
+    p_{j,m}&=\frac{\sum_{i=1}^Nw_{i,m}\left(x_{i,j}+\frac{1}{N}(\alpha-1)\right)}{\left(\frac{2}{N}(\alpha-1)+1\right)\sum_{i=1}^Nw_{i,m}}
+\end{align*}
 \end{document}

### 2014-05-03 stat246/hw4_files/p3.py
s4

@@ -1,0 +1,29 @@
+import numpy as np
+
+'''
+Returns a 10 by N by 900 array. Digit classes on the first axis, individual cases
+on the second. N=100 for train, 1000 for test
+'''
+def load_digits(case='tr'):
+    if case == 'tr':
+        N = 100
+    else:
+        N = 1000
+    digits = np.zeros((10, N, 900), dtype='int8')
+    for i in range(10):
+        c = np.fromfile('{0}{1}.bin'.format(case, i), dtype='int8')
+        digits[i] = c.reshape(N, 900)
+
+    return digits
+
+'''
+Takes a N by 900 matrix (each row is a data vector), and trains EM on it.
+'''
+def em_map(data, M):
+    # Set the zetas to be all one and set alpha to 2
+    alpha = 2
+    zeta = 1
+
+    N = data.shape[0]
+    d = data.shape[1]
+

### 2014-05-04 stat246/hw4.tex
optimization

@@
 \begin{align*}
-    \pi_m&=\frac{1}{N}\sum_{i=1}^N\left(w_{i,m}+\frac{1}{N}(\zeta_m-1)\right)\\
+    \pi_m&=\kappa\sum_{i=1}^N\left(w_{i,m}+\frac{1}{N}(\zeta_m-1)\right)\\
     p_{j,m}&=\frac{\sum_{i=1}^Nw_{i,m}\left(x_{i,j}+\frac{1}{N}(\alpha-1)\right)}{\left(\frac{2}{N}(\alpha-1)+1\right)\sum_{i=1}^Nw_{i,m}}
 \end{align*}
+where $\kappa$ is some constant that makes the $\pi_m$ sum to 1.
+
+For computational purposes in computing the $w_{i,m}$, instead of computing
+\[\frac{P(x_i|Y=m)P(Y=m)}{\sum_{j=1}^MP(x_i|Y=j)P(Y=j)}\]
+compute instead
+\[\log (P(x_i|Y=m)P(Y=m))=\sum_{j=1}^d\left(x_{i,j}\log p_{j,m}+(1-x_{i,j})\log(1-p_{j,m})\right)+\log\pi_m\]
+for each pair $i,m$ and call them the $\alpha_{i,m}$. For each $i$, calculate $l_i=\max_m\alpha_{i,m}$, and set $w_{i,m}=\frac{\exp(\alpha_{i,m}-l_i)}{\sum_{m'}\exp(\alpha_{i,m'}-l_i)}$.
 \end{document}

### 2014-05-04 stat246/hw4_files/p3.py

@@ -20,10 +20,75 @@ def load_digits(case='tr'):
 Takes a N by 900 matrix (each row is a data vector), and trains EM on it.
 '''
 def em_map(data, M):
-    # Set the zetas to be all one and set alpha to 2
+    # Set the zetas to be all 2 and set alpha to 2
     alpha = 2
-    zeta = 1
+    zeta = 2
 
     N = data.shape[0]
     d = data.shape[1]
+    # In lieu of randomly assigning classes, we'll just assign the classes by taking
+    # the data point index mod M. Note that assigning a class to a point i is equivalent to setting
+    # w_{i,m}=1 for the assigned class m and zero for all other classes.
+    cond_probs = np.zeros((N, M), dtype='float64')
+    pis = np.zeros(M, dtype='float64')
+    bern_probs = np.zeros((d, M), dtype='float64')
+    for i in range(N):
+        cond_probs[i,i%M] = 1
 
+    pis, bern_probs = calc_params(data, cond_probs, alpha, zeta)
+
+    while True:
+        cond_probs = calc_cond_probs(data, pis, bern_probs)
+        new_pis, new_bern_probs = calc_params(data, cond_probs, alpha, zeta)
+        m1 = np.linalg.norm(new_pis - pis)
+        m2 = np.linalg.norm(new_bern_probs - bern_probs)
+        if m1 < epsilon and m2 < epsilon:
+            return new_pis, new_bern_probs
+        print m2
+        pis = new_pis
+        bern_probs = new_bern_probs
+
+'''
+Calculates parameters given the data and the W matrix, along with some priors.
+Returns the new pis and bernoulli params
+'''
+def calc_params(data, W, alpha, zeta):
+    N = data.shape[0]
+    pis = W.sum(0) + zeta - 1
+    pis /= pis.sum()
+
+    p_numer = np.dot(data.T + (alpha-1)/N, W)
+    p_denom = W.sum(0)*(2*(alpha-1)/N + 1)
+
+    bern_probs = p_numer / p_denom
+    return pis, bern_probs
+
+'''
+Calculates the matrix of conditional probabilities of given
+parameters.
+'''
+def calc_cond_probs(data, pis, bern_probs):
+    M = pis.shape[0]
+    N = data.shape[0]
+    weights = np.zeros((N,M), dtype='float64')
+
+    log_joints = np.dot(data.T, np.log(bern_probs)) + np.dot((1-data).T, np.log(1-bern_probs))
[…]

### 2014-05-04 stat246/hw4.tex
s4 done

@@
 \end{align*}
-where $\kappa$ is some constant that makes the $\pi_m$ sum to 1.
+where $\kappa$ is some constant that makes the $\pi_m$ sum to 1. In practice, all the $\zeta_m$ were set to $2$, as were all the $\alpha$s
 
@@
 for each pair $i,m$ and call them the $\alpha_{i,m}$. For each $i$, calculate $l_i=\max_m\alpha_{i,m}$, and set $w_{i,m}=\frac{\exp(\alpha_{i,m}-l_i)}{\sum_{m'}\exp(\alpha_{i,m'}-l_i)}$.
+
+Now for the results. The fitted probability vectors are shown below:
+
+\includegraphics[width=0.9\textwidth]{hw4_files/M1_princip.png}
+
+\includegraphics[width=0.9\textwidth]{hw4_files/M3_princip.png}
+
+\includegraphics[width=0.9\textwidth]{hw4_files/M5_princip.png}
+
+In classifying the test set, I went class-by-class and computed the likelihoods of each data vector given the parameters that were fitted to a certain class.
+The likelihoods were computed by exponentiating the log-likelihood, which is more numerically stable than just multiplying together the 900 Bernoulli likelihoods.
+The error rates observed were $0.2041$ for $M=1$, $0.1396$ for $M=3$, and $0.1383$ for $M=5$.
+This indicates that the digits can already be described quite well by a mixture of three components (which, looking at the images above, correspond to variations in how people draw the digits).
 \end{document}

### 2014-05-04 stat246/hw4_files/p3.py

@@ -19,10 +19,10 @@ def load_digits(case='tr'):
 '''
 Takes a N by 900 matrix (each row is a data vector), and trains EM on it.
 '''
-def em_map(data, M):
+def em_map(data, M, epsilon=1e-5):
     # Set the zetas to be all 2 and set alpha to 2
-    alpha = 2
-    zeta = 2
+    alpha = 2.0
+    zeta = 2.0
 
     N = data.shape[0]
     d = data.shape[1]
@@ -72,7 +72,7 @@ def calc_cond_probs(data, pis, bern_probs):
     N = data.shape[0]
     weights = np.zeros((N,M), dtype='float64')
 
-    log_joints = np.dot(data.T, np.log(bern_probs)) + np.dot((1-data).T, np.log(1-bern_probs))
+    log_joints = np.dot(data, np.log(bern_probs)) + np.dot((1-data), np.log(1-bern_probs))
     # fugly hack that exploits a numpy bug
     try:
         log_joints.T += np.log(pis)
@@ -92,3 +92,79 @@ def calc_cond_probs(data, pis, bern_probs):
         pass
 
     return weights
+
+'''
+Runs the model for all 10 digits, on M=1,3,5, plots the components, and returns
+the parameters.
+'''
+def train_on_digits():
+    import matplotlib.pyplot as plt
+    train_set = load_digits('tr')
+
+    pis_1 = []
+    probs_1 = []
+    plt.figure(figsize=(8,3))
+    for i in range(10):
+        plt.subplot(2, 5, i+1)
+        plt.axis('off')
+        pis, probs = em_map(train_set[i], 1)
+        plt.imshow(probs[:,0].reshape(30, 30), cmap=plt.cm.Greys)
+        pis_1.append(pis)
+        probs_1.append(probs)
+    #plt.savefig('M1_princip.png', bbox_inches='tight')
+
+    pis_3 = []
+    probs_3 = []
+    plt.figure(figsize=(8.5,3))
+    for i in range(10):
+        pis, probs = em_map(train_set[i], 3)
+        for j in range(3):
+            plt.subplot(3, 10, j*10+i+1)
+            plt.axis('off')
+            plt.imshow(probs[:,j].reshape(30, 30), cmap=plt.cm.Greys)
+        pis_3.append(pis)
+        probs_3.append(probs)
+    #plt.savefig('M3_princip.png', bbox_inches='tight')
+
+    pis_5 = []
+    probs_5 = []
[…]

### 2014-05-05 cs250-hw/hw3.tex
c3

\documentclass{article}
\usepackage{geometry}
\usepackage[namelimits,sumlimits]{amsmath}
\usepackage{amssymb,amsfonts}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage[cm]{fullpage}
\newcommand{\tab}{\hspace*{5em}}
\newcommand{\conj}{\overline}
\newcommand{\dd}{\partial}
\newcommand{\ep}{\epsilon}
\newcommand{\openm}{\begin{pmatrix}}
\newcommand{\closem}{\end{pmatrix}}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\Null}{null}
\newcommand{\nc}{\newcommand}
\newcommand{\rn}{\mathbb{R}}
\newcommand{\zn}{\mathbb{Z}}
\nc{\cn}{\mathbb{C}}
\nc{\ssn}[1]{\subsubsection*{#1}}
\nc{\inner}[2]{\langle #1,#2\rangle}
\nc{\h}[1]{\widehat{#1}}
\nc{\tl}[1]{\widetilde{#1}}
\nc{\norm}[1]{\left\|{#1}\right\|}
\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
\begin{document}
Name: Hall Liu

Date: \today 
\vspace{1.5cm}
\subsection*{2}
\ssn{a}
To derive the conditional distribution, first we need the joint distribution. Let $V$ be the number of words in the vocabulary. Let $\beta$ be a $K\times V$ matrix (each row corresponding to a topic). Let $\theta$ be a $D\times K$ matrix, each row corresponding to a document. Let $z$ and $w$ be $D\times N$ matrices, containing the index/topic of the particular word position in that particular document. If one of these matrices is written with a single subscript, it refers to that row of the matrix.

The joint distribution corresponding to the graphical description is
\[\prod_{i=1}^KP(\beta_i;\alpha)\cdot\prod_{d=1}^D\left(P(\theta_d;\eta)\prod_{n=1}^NP(z_{d,n}|\theta_d)P(w_{d,n}|z_{d,n},\beta)\right)\]
where $\alpha$ and $\eta$ are fixed parameters. If we plug in the corresponding pdfs for each parameter, we obtain the equation for the joint density:
\[\prod_{i=1}^K\left(\frac{1}{C(\eta)}\prod_{i=1}^V\beta_{ij}^{\eta-1}\right)\cdot\prod_{d=1}^D\left(\left(\frac{1}{C(\alpha)}\prod_{i=1}^K\theta_{dj}^{\eta-1}\right)\prod_{n=1}^N\theta_{d,z_{dn}}\beta_{z_{dn},w_{dn}}\right)\]
where $C(x)$ is the integral of the Dirichlet distribution without the normalizing constant.

Now, we want to integrate this over the space that all the $\theta$s lie in. Let $\Omega_\theta$ denote this space, and let $\Delta_d$ denote the simplex in which $\theta_d$ lies. We then have the integral
\begin{align*}
    &\int_{\Omega_\theta}\prod_{i=1}^K\left(\frac{1}{C(\eta)}\prod_{i=1}^V\beta_{ij}^{\eta-1}\right)\cdot\prod_{d=1}^D\left(\left(\frac{1}{C(\alpha)}\prod_{j=1}^K\theta_{dj}^{\alpha-1}\right)\prod_{n=1}^N\theta_{d,z_{dn}}\beta_{z_{dn},w_{dn}}\right)\\
    &=\prod_{i=1}^K\left(\frac{1}{C(\eta)}\prod_{i=1}^V\beta_{ij}^{\eta-1}\right)\cdot\int_{\Omega_\theta}\prod_{d=1}^D\left(\left(\frac{1}{C(\alpha)}\prod_{j=1}^K\theta_{dj}^{\alpha-1}\right)\prod_{n=1}^N\theta_{d,z_{dn}}\beta_{z_{dn},w_{dn}}\right)\\
    &=\prod_{i=1}^K\left(\frac{1}{C(\eta)}\prod_{i=1}^V\beta_{ij}^{\eta-1}\right)\cdot\prod_{d=1}^D\int_{\Delta_d}\left(\left(\frac{1}{C(\alpha)}\prod_{j=1}^K\theta_{dj}^{\alpha-1}\right)\prod_{n=1}^N\theta_{d,z_{dn}}\beta_{z_{dn},w_{dn}}\right)\\
    &=\prod_{i=1}^K\left(\frac{1}{C(\eta)}\prod_{i=1}^V\beta_{ij}^{\eta-1}\right)\cdot\prod_{d=1}^D\left(\prod_{n=1}^N\beta_{z_{dn},w_{dn}}\right)\int_{\Delta_d}\left(\left(\frac{1}{C(\alpha)}\prod_{j=1}^K\theta_{dj}^{\alpha-1}\right)\prod_{n=1}^N\theta_{d,z_{dn}}\right)\\
\end{align*}
Let $n$ be a $D\times K$ matrix, with $n_{ij}$ being the number of times topic $j$ appears in document $i$. Then, the integral above (omitting the constant multiplier out front) can be rewritten as
\[\int_{\Delta_d}\left(\frac{1}{C(\alpha)}\prod_{j=1}^K\theta_{dj}^{\alpha-1+n_{dj}}\right)\]
which, using the Dirichlet normalizing constant, can be rewritten as
\[\frac{1}{C(\alpha)}\frac{\prod_{j=1}^K\Gamma(\alpha+n_{dj})}{\Gamma(\sum_{j=1}^N\alpha+n_{dj})}\]
so we can write the marginal, excluding $\theta$, as
\[\prod_{i=1}^K\left(\frac{1}{C(\eta)}\prod_{i=1}^V\beta_{ij}^{\eta-1}\right)\cdot\prod_{d=1}^D\left(\prod_{n=1}^N\beta_{z_{dn},w_{dn}}\right)\frac{1}{C(\alpha)}\frac{\prod_{j=1}^K\Gamma(\alpha+n_{dj})}{\Gamma(\sum_{j=1}^N\alpha+n_{dj})}\]
Fix some document $d$, since we're working locally. The marginal, given this document alone, is the same as the above expression, but with the product over $d$ omitted:
\[\prod_{i=1}^K\left(\frac{1}{C(\eta)}\prod_{i=1}^V\beta_{ij}^{\eta-1}\right)\cdot\frac{1}{C(\alpha)}\frac{\prod_{j=1}^K\Gamma(\alpha+n_{dj})}{\Gamma(\sum_{j=1}^N\alpha+n_{dj})}\left(\prod_{n=1}^N\beta_{z_{dn},w_{dn}}\right)\]

Now, to find $P(z_{dn}|z_{d,-n},\beta,w_d;\alpha,\eta)$, we divide the joint likelihood of all of these by the marginal distribution of $z_{d,-n},\beta,w_d$. For the joint, we already have the likelihood in the marginal distribution derived above -- if we wish to evaluate the joint likelihood for a particular value of $z_{dn}$, we plug it in to the last term above while holding all the other $z_{d,*}$ terms constant. Finally, to find the marginal, we just have to evaluate $\sum_{k=1}^KP(z_{dn}=k,z_{d,-n},\beta,w_d;\alpha,\eta)$, and dividing the two gives an expression for the desired conditional probability.
\ssn{b}
Gibbs sampling iteratively samples from the above conditional probability to obtain samples from the overall distribution of $z_d$ given $\beta$ and $w_d$. Thus, if we obtain a large number of samples of the topics assigned to words within the document, we can estimate the proportion of each topic that exists within the document, which is what $\theta_d$ is.
\subsection*{3}
The expression inside the integral is the joint distribution of $w_d$, $z_d$, and $\theta_d$, conditional upon $\beta$ and the fixed parameter $\alpha$. The joint distribution of these variables is written above in the beginning of (2); obtaining the conditional upon $\beta$ simply involves dividing by the marginal distribution of $\beta$. Fortunately, the joint distribution contains that marginal as a multiplicative term, and the conditional is simply 
\[\cdot\prod_{d=1}^D\left(\left(\frac{1}{C(\alpha)}\prod_{i=1}^K\theta_{dj}^{\eta-1}\right)\prod_{n=1}^N\theta_{d,z_{dn}}\beta_{z_{dn},w_{dn}}\right)\]
Removing the dependence upon $d$ gives us
\[\left(\frac{1}{C(\alpha)}\prod_{i=1}^K\theta_{dj}^{\eta-1}\right)\prod_{n=1}^N\theta_{d,z_{dn}}\beta_{z_{dn},w_{dn}}\]
whose integral over $\Delta_d$ was evaluated in (2). The result is
\[\frac{1}{C(\alpha)}\frac{\prod_{j=1}^K\Gamma(\alpha+n_{dj})}{\Gamma(\sum_{j=1}^N\alpha+n_{dj})}\left(\prod_{n=1}^N\beta_{z_{dn},w_{dn}}\right)\]
\subsection*{4}
\ssn{a}
Continuing from the marginal distribution obtained in (2a), we now want to integrate out $\beta$. Let $\Omega_\beta$ indicate the space in which all the $\beta$s sit, and let $\Delta_k$ denote the simplex in which $\beta_k$ lies. Then, we have the integral
\begin{align*}
    &\int_{\Omega_\beta}\prod_{i=1}^K\left(\frac{1}{C(\eta)}\prod_{j=1}^V\beta_{ij}^{\eta-1}\right)\cdot\prod_{d=1}^D\left(\prod_{n=1}^N\beta_{z_{dn},w_{dn}}\right)\frac{1}{C(\alpha)}\frac{\prod_{j=1}^K\Gamma(\alpha+n_{dj})}{\Gamma(\sum_{j=1}^N\alpha+n_{dj})}\\
    &=\frac{1}{C(\alpha)^D}\frac{1}{C(\eta)^K}\left(\prod_{d=1}^D\frac{\prod_{j=1}^K\Gamma(\alpha+n_{dj})}{\Gamma(\sum_{j=1}^N\alpha+n_{dj})}\right)\int_{\Omega_\beta}\left(\prod_{i=1}^K\prod_{j=1}^V\beta_{ij}^{\eta-1}\right)\left(\prod_{d=1}^D\prod_{n=1}^N\beta_{z_{dn},w_{dn}}\right)\\
\end{align*}
Let $r_{ij}$ denote the number of times topic $i$ is assigned to word $j$ globally. This term is then a function of the parameter $z$. With this, we can merge the products together to obtain
\begin{align*}
    &\frac{1}{C(\alpha)^D}\frac{1}{C(\eta)^K}\left(\prod_{d=1}^D\frac{\prod_{j=1}^K\Gamma(\alpha+n_{dj})}{\Gamma(\sum_{j=1}^N\alpha+n_{dj})}\right)\int_{\Omega_\beta}\left(\prod_{i=1}^K\prod_{j=1}^V\beta_{ij}^{\eta-1+r_{ij}}\right)\\
    &=\frac{1}{C(\alpha)^D}\frac{1}{C(\eta)^K}\left(\prod_{d=1}^D\frac{\prod_{j=1}^K\Gamma(\alpha+n_{dj})}{\Gamma(\sum_{j=1}^N\alpha+n_{dj})}\right)\left(\prod_{i=1}^K\int_{\Delta_i}\prod_{j=1}^V\beta_{ij}^{\eta-1+r_{ij}}\right)\\
\end{align*}
The inner product has the form of a Dirichlet pdf, so we can evaluate the integral to obtain
\[\frac{1}{C(\alpha)^D}\frac{1}{C(\eta)^K}\left(\prod_{d=1}^D\frac{\prod_{j=1}^K\Gamma(\alpha+n_{dj})}{\Gamma(\sum_{j=1}^N\alpha+n_{dj})}\right)\left(\prod_{i=1}^K\frac{\prod_{j=1}^V\Gamma(\eta+r_{ij})}{\Gamma\left(\sum_{j=1}^V(\eta+r_{ij})\right)}\right)\]
which is the final form of the marginal distribution of $z$ conditional upon the data. The dependence of the above expression on the $z_{dn}$ and the words $w_{dn}$ comes in the form of the $r_{ij}$, which are word counts per topic.

Now that we have the distribution on all of the $z$s, we can evaluate $P(z_{nd}=k|z_{-(n,d)},w)$ in the same way as in (2a): first calculate the joint likelihood by plugging in what we want for $k$ and leaving the other $z$s as-is, then evaluate for all other values of $k$ and sum together to obtain the marginal likelihood. Taking their ratio will yield the conditional likelihood.
\ssn{b}
\end{document}

### 2014-05-05 cs250-hw/hw3.tex
c3

@@
 \ssn{b}
+This conditional likelihood gives us a way to use Gibbs sampling to sample values of $z$ with only dependence on the words and preassigned topics.
+As before, with a large number of samples of the $z$s, we can calculate an estimate for $\beta$ by averaging the within-topic word counts that each $z$ provides us with.
 \end{document}

### 2014-05-06 stat246/sheet.tex
..

\documentclass{article}
\usepackage[margin=0.2in]{geometry}
\usepackage[namelimits,sumlimits]{amsmath}
\usepackage{amssymb,amsfonts}
\usepackage{multicol}
\newcommand{\nc}{\newcommand}
\newcommand{\tab}{\hspace*{5em}}
\newcommand{\conj}{\overline}
\newcommand{\dd}{\partial}
\newcommand{\openm}{\begin{pmatrix}}
\newcommand{\closem}{\end{pmatrix}}
\nc{\pd}{\partial}
\nc{\del}{\nabla}
\nc{\ep}{\epsilon}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\cov}{cov}
\begin{document}
\small
Bayes classification rule for discrete cases: maximize $P(Y=m|X)$ over all possible values $m$. Simplifies to maximizing $P(X|Y=m)P(Y=m)$ because $P(X)$ is invariant across $m$s. Useful to think about this in terms of regions in the space that get classified into certain areas. Expected loss is integral of loss times probability over entire space. Useful to evaluate this in terms of the regions.

Multivariate normal distribution: pdf is $\frac{1}{(2\pi|\Sigma|)^{d/2}}\exp\left(\frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu)\right)$. the MGF is $\exp\left(\mu^Tt+\frac{1}{2}t^T\Sigma t\right)$. If $x\sim N(\mu,\Sigma)$, then an affine transform $c+Bx$ is distributed as $N(B\mu+c, B\Sigma B^T)$. If the mean is partitioned as $\openm \mu_1\\\mu_2\closem$ and the covariance is partitioned as $\openm\Sigma_{11}&\Sigma_{12}\\\Sigma_{21}&\Sigma_{22}\closem$, then the distribution of the first part conditional on the second is normal with mean $\mu_1+\Sigma_{12}\Sigma_{22}^{-1}(x_2-\mu_2)$ and covariance $\Sigma_{11}-\Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21}$.

ML estimates for normal RVs: use trace trick to write $x^T\Sigma^{-1} x=\tr(xx^T\Sigma^{-1})$. Proven in class that the argmin over $\Sigma$ of $-\log|\Sigma|+\tr(A\Sigma^{-1})$ is given by $\Sigma=A$. Proof of this is done by taking square root of $A$, letting $B=A^{1/2}\Sigma^{-1}A^{1/2}$, and maximizing on eigenvalues of $B$.

QDA and LDA refer to estimating decision boundaries between multivariate normals. Use a pooled covariance for LDA. Pooled covariance MLE is the weighted average of the class covariances. Prove using above.

Exponential family: distributions can be written as $p(x)=h(x)g(\eta)\exp(\eta^Tu(x))$, where $\eta$ are the parameters. The fact that this is a distribution leads to the identity $g(\eta)^{-1}=\int h(x)\exp(\eta^Tu(x))\,dx$, which is useful. This leads to $E(u(x))=-\nabla \log g(\eta)$ and $\cov(u(x))$ is the Hessian of the negative log of $g$. Prove by letting $g(\eta)=\exp(-A(\eta))$, using the integral identity, and differentiating wrt $g$ twice. Exponential family has a conjugate prior $f(\chi,\nu)g(\eta)^\nu\exp(\nu\eta^T\chi)$.

A few distributions as exponential families: $N(\mu,\Sigma)$ has $h=(2\pi)^{-d/2}$, $g=|\Sigma|^{-d/2}\exp(-\frac{1}{2}\mu^T\Sigma^{-1}\mu)$, $\eta=\openm\Sigma^{-1}\mu\\-\frac{1}{2}\text{flat}(\Sigma^{-1})\closem$, and $T(x)=\openm x\\\text{flat}(xx^T)\closem$. Binomial with parameters $n,p$ is $\binom{n}{x}\exp\left(x\log\left(\frac{p}{1-p}\right)+n\log(1-p)\right)$. Gamma distr with shape $\alpha$ and rate $\beta$ has $h(x)=1$,$g(\eta)=\frac{\Gamma(\alpha)}{\beta^\alpha}$, $\eta=\openm\alpha-1\\-\beta\closem$, and $u(x)=\openm\log x\\x\closem$. Beta distribution with parameters $\alpha,\beta$ has $h(x)=\frac{1}{x(1-x)}$, $g(\eta)=\frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha+\beta)}$, $\eta=\openm\alpha\\\beta\closem$, and $u(x)=\openm\log x\\\log(1-x)\closem$.

EM algorithm steps: First write the complete-data log likelihood. Then take the expectation over the hidden variable conditioned by the previous iteration and the data, which will give an expression involving the conditional density of $Y$ and a sum. Finally, take derivatives to get a ML estimate of this expectation and update parameters. If using Bayesian priors to do a MAP estimation, maximize the expectation of the log-likelihood plus the log of the prior. Probably should make it a conjugate prior.

Entropy of a distribution $q$ is $H(q)=-\sum_yq(y)\log q(y)$. KL-divergence is $K(q\|p)=-H(q)-E_q(\log p)=\sum_y\log\frac{q(y)}{p(y)}q(y)$. Jensen's inequality for a convex $f$ states that $f\left(\frac{\sum a_ix_i}{\sum a_i}\right)\leq\frac{\sum a_if(x_i)}{\sum a_i}$, and reversed for concave. If convexity is strict, then equality holds iff all the $x$s are equal.

Graphical models: modeled by a DAG that induces a chain of multiplied conditional probabilities. Markov blanket of a set of nodes is the union of the set of its parents, children, and nodes which share children with members of the set (not including the set itself). If $A$ is the set, $\conj{A}$ is its Markov blanket, then for any $n\in V-A-\conj{A}$, $n$ is conditionally independent from anything in $A$.

A set $C$ $d$-separates $A$ and $B$ if for any path from $A$ to $B$, there's either a T-T or H-T node in the path that's in $C$ or there's a H-H node in the path which is not in $C$ and its descendants are not in $C$/

Hidden Markov models -- insert graph below
\end{document}

### 2014-05-08 cs233/hw2/paxos.py
c2

@@ -1,0 +1,127 @@
+import sys
+
+class Msg():
+    def __init__(self, src, dst, mtype, data):
+        self.src = src
+        self.dst = dst
+        self.mtype = mtype
+        # data is going to be a dict that is accessed differently depending on mtype
+        self.data = data
+
+class Event():
+    def __init__(self, t):
+        self.t = t
+        self.fails = []
+        self.recs = []
+        self.prop = None
+        self.val = None
+
+    def run(self, net_queue, computers, prop_number):
+        # Take care of the failures and recoveries
+        for c in self.fails:
+            computers[c].failed = True
+        for c in self.recs:
+            computers[c].failed = False
+
+        # Check for proposals and run them if present
+        if self.prop is not None and self.val is not None:
+            msg = Msg(None, computers[self.prop], 'PROPOSE', {'val': self.val})
+            computers[self.prop].recv_message(msg)
+            return True
+
+        return False
+
+
+class Computer():
+    msg_dispatch = {
+            'PROPOSE': Computer.propose
+    }
+
+    def __init__(self, cid, **kwargs):
+        self.cid = cid
+        self.other_computers = kwargs.pop('computers')
+        self.mq = kwargs.pop('mq')
+        self.failed = False
+        self.role = cid[0]
+        self.misc_data = kwargs
+    
+    def recv_message(self, msg):
+
+
+
+def parse_input(infile=sys.stdin):
+    events = []
+    for line in infile:
+        if len(line) < 3:
+            continue
+        if line[0] == '#':
+            continue
+        parts = line.split()
+        if len(parts) == 3:
+            n_prop = int(parts[0])
+            n_acc = int(parts[1])
+            tmax = int(parts[2])
+            continue
+        if len(parts) == 2:
+            break
+        
+        t = int(parts[0])
+        action = parts[1]
+
+        if len(events) == 0 or t > events[-1].t:
+            e = Event(t)
+            events.append(e)
+        else:
+            e = events[-1]
+
+        if action == 'PROPOSE':
+            e.prop = 'P' + parts[2]
+            e.val = int(parts[3])
+        else:
+            action_list = e.fails if action == 'FAIL' else e.recs
+            action_list.append(parts[2][0]+parts[3])
+    
+    return n_prop, n_acc, tmax, events
+
+def extract_message(net_queue):
+    if len(net_queue) == 0:
+        return None
+
+    for m in net_queue:
[…]

### 2014-05-09 cs233/hw2/paxos.py
c2

@@ -26,7 +26,7 @@ class Event():
         # Check for proposals and run them if present
         if self.prop is not None and self.val is not None:
             msg = Msg(None, computers[self.prop], 'PROPOSE', {'val': self.val})
-            computers[self.prop].recv_message(msg)
+            computers[self.prop].recv_message(msg, self.t)
             return True
 
         return False
@@ -34,18 +34,74 @@ class Event():
 
 class Computer():
     msg_dispatch = {
-            'PROPOSE': Computer.propose
+            'PROPOSE': Computer.propose,
+            'PREPARE': Computer.prepare,
+            'PROMISE': Computer.promise,
+            'ACCEPT': Computer.accept,
+            'ACCEPTED': Computer.accepted,
+            'REJECTED': Computer.rejected
     }
 
     def __init__(self, cid, **kwargs):
         self.cid = cid
         self.other_computers = kwargs.pop('computers')
-        self.mq = kwargs.pop('mq')
+        acceptor_count = len(self.other_computers[1])
+        self.majority_count = acceptor_count / 2 + 1
+
         self.failed = False
-        self.role = cid[0]
-        self.misc_data = kwargs
+        self.highest_promised = 0
+        self.highest_accepted = None
+        self.__dict__.update(kwargs)
+
+        self.state = ('IDLE', {})
     
-    def recv_message(self, msg):
+    def recv_message(self, msg, t):
+        msg_dispatch[msg.mtype](self, msg, t)
+
+    def propose(self, msg, t):
+        prop_ctr = self.prop_ctr
+        this_prop_no = prop_ctr[0]
+        prop_ctr[0] += 1
+
+        msg_data = {'prop_no': this_prop_no}
+        for c in self.other_computers[1]:
+            new_msg = Msg(self, c, 'PREPARE', msg_data)
+            self.mq.append(new_msg)
+
+        print '{0}: {1} -> {2} PROPOSE v={3}'.format(t, '  ', computers[self.prop].cid, msg.data['val'])
+
+        self.state = ('PROMISE_WAIT', {'n_proms': 0, 'n_rejs': 0, 'prop_val': msg.data['val']})
+
+    def prepare(self, msg, t):
+        print '{0}: {1} -> {2} PREPARE n={3}'.format(t, msg.src.cid, self.cid, msg.data['prop_no'])
+        pn = msg.data['prop_no']
+        if pn < self.highest_promised:
[…]

### 2014-05-10 cs233/hw2/paxos.py
done

@@ -16,12 +16,14 @@ class Event():
         self.prop = None
         self.val = None
 
-    def run(self, net_queue, computers, prop_number):
+    def run(self, net_queue, computers):
         # Take care of the failures and recoveries
         for c in self.fails:
             computers[c].failed = True
+            print '{0: 3d}: ** {1} FAILS **'.format(self.t, c)
         for c in self.recs:
             computers[c].failed = False
+            print '{0: 3d}: ** {1} RECOVERS **'.format(self.t, c)
 
         # Check for proposals and run them if present
         if self.prop is not None and self.val is not None:
@@ -33,76 +35,136 @@ class Event():
 
 
 class Computer():
-    msg_dispatch = {
-            'PROPOSE': Computer.propose,
-            'PREPARE': Computer.prepare,
-            'PROMISE': Computer.promise,
-            'ACCEPT': Computer.accept,
-            'ACCEPTED': Computer.accepted,
-            'REJECTED': Computer.rejected
-    }
 
     def __init__(self, cid, **kwargs):
+        # Common attributes
         self.cid = cid
         self.other_computers = kwargs.pop('computers')
-        acceptor_count = len(self.other_computers[1])
-        self.majority_count = acceptor_count / 2 + 1
-
         self.failed = False
-        self.highest_promised = 0
-        self.highest_accepted = None
+
+        # Proposer attributes
+        if self.cid[0] == 'P':
+            self.majority_count = kwargs.pop('acceptor_count') / 2 + 1
+            self.state = ('IDLE', {})
+        
+        # Acceptor attributes
+        if self.cid[0] == 'A':
+            self.highest_promised = 0
+            self.highest_accepted = (0, None)
+
         self.__dict__.update(kwargs)
 
-        self.state = ('IDLE', {})
+        self.msg_dispatch = {
+                'PROPOSE': Computer.propose,
+                'PREPARE': Computer.prepare,
+                'PROMISE': Computer.promise,
+                'ACCEPT': Computer.accept,
+                'ACCEPTED': Computer.accepted,
+                'REJECTED': Computer.rejected
+        }
     
     def recv_message(self, msg, t):
-        msg_dispatch[msg.mtype](self, msg, t)
+        self.msg_dispatch[msg.mtype](self, msg, t)
 
     def propose(self, msg, t):
[…]

### 2014-05-10 cs233/hw2/run.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+python2 paxos.py

### 2014-05-10 cs233/hw2/test.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+
+for f in `ls tests | grep "\.in"`
+do
+    ./run.sh < tests/$f | diff -u --ignore-all-space `echo tests/$f | sed 's/\.in/\.out/'` -
+done

### 2014-05-12 stat246/hw5.tex
s5

\documentclass{article}
\usepackage{geometry}
\usepackage[namelimits,sumlimits]{amsmath}
\usepackage{amssymb,amsfonts}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage[cm]{fullpage}
\usepackage{mathrsfs}
\newcommand{\tab}{\hspace*{5em}}
\newcommand{\conj}{\overline}
\newcommand{\dd}{\partial}
\newcommand{\ep}{\epsilon}
\newcommand{\openm}{\begin{pmatrix}}
\newcommand{\closem}{\end{pmatrix}}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\Null}{null}
\newcommand{\nc}{\newcommand}
\newcommand{\rn}{\mathbb{R}}
\nc{\cn}{\mathbb{C}}
\nc{\ssn}[1]{\subsubsection*{#1}}
\nc{\inner}[2]{\langle #1,#2\rangle}
\nc{\h}[1]{\widehat{#1}}
\nc{\tl}[1]{\widetilde{#1}}
\nc{\norm}[1]{\left\|{#1}\right\|}
\begin{document}

Name: Hall Liu

Date: \today 
\vspace{1.5cm}

\subsection*{1}
\ssn{a}
$P(D)P(B_1)P(B_2)P(Y|D)P(X_1|Y,B_1)P(X_2|Y,B_2)$
\ssn{b}
\[\sum_DP(D)P(B_1)P(B_2)P(Y|D)P(X_1|Y,B_1)P(X_2|Y,B_2)=P(B_1)P(B_2)P(X_1|Y,B_1)P(X_2|Y,B_2)\sum_DP(Y|D)P(D)\]
    \[=P(B_1)P(B_2)P(Y)P(X_1|Y,B_1)P(X_2|Y,B_2)\]
where $P(Y)$ is the marginal of $Y$ wrt its joint distribution with $D$. $B_1$ and $Y$ are independent under this marginal since they have separate factors.
\ssn{c}
Summing over $B_1$, $B_2$ gives
\[P(Y)\left(\sum_{B_1}P(B_1)P(X_1|Y,B_1)\right)\left(\sum_{B_2}P(B_2)P(X_2|Y,B_2)\right)=P(Y)\left(\sum_{B_1}P(X_1,B_1|Y)\right)\left(\sum_{B_2}P(X_2,B_2|Y)\right)\]
\[=P(Y)P(X_1|Y)P(X_2|Y)\]
$X_1$ and $X_2$ are conditionally independent given $Y$, since the only path from $X_1$ to $X_2$ goes through and is blocked by $Y$.
\subsection*{2}
\ssn{a}
\vspace{150pt}
The $(Y_i,X_i)$ form a HMM because marginalizing over the $Z_i$ and the $U_i$ simply consists of removing the nodes, since they're all ancestor nodes. Removing them gives a graph in the HMM form.
\ssn{b}
We can write $Y_i=Y_1+\sum_{j=1}^{i-1}U_j$ and $X_i=Y_1+\sum_{j=1}^{i-1}U_j+Z_n$. Since these are both linear combinations of the mutually independent variables $Y_1,U_{1:n},Z_{1:n}$, we have that the set of variables $X_{1:n},Y_{1:n},Z_{1:n}$ and $U_{1:n}$ are produced by a linear transformation of $Y_1,U_{1:n},Z_{1:n}$, which are jointly normal. Thus all the variables are jointly normal. Further, since all the ancestors have mean 0, applying a linear transformation to that mean vector will also result in a mean vector of zero for all the variables.
\ssn{c}
We have $Y_s=Y_1+\sum_{i=1}^{s-1}U_i$ and $Y_{n+1}=Y_1+\sum_{i=1}^nU_i$. All these variables are mutually independent, so their covariance is
\[\cov(Y_s,Y_{n+1})=\var(Y_1)+\sum_{i=1}^{s-1}\var(U_i)=\sigma_1^2+(s-1)\sigma_u^2\]
Writing $X_s$ as $Y_s+Z_s$, we have $\cov(X_s, Y_{n+1})=\cov(Y_s,Y_{n+1})+\cov(Z_s, Y_{n+1})$, and the second term is zero, so the covariance is the same as $\cov(Y_s, Y_{n+1})$.

\[\var(Y_{n+1})=\cov(Y_{n}+U_n, Y_n+U_n)=\var(Y_n)+\var(U_n)=a_n+\sigma_u^2\]
\[\cov(X_{n+1},Y_{n+1})=\cov(Y_{n+1}+Z_{n+1},Y_{n+1})=\var(Y_{n+1})+0=a_n+\sigma_u^2\]
\[\var(X_{n+1})=\cov(Y_{n+1}+Z_{n+1},Y_{n+1}+Z_{n+1})=\var(Y_{n+1})+\var(Z_{n+1})=a_n+\sigma_u^2+\sigma_z^2\]
\[\cov(X_n, X_{n+1})=\cov(Y_n+Z_n,Y_{n+1}+Z_{n+1})=\cov(Y_n,Y_{n+1})=\sigma_1^2+(n-1)\sigma_u^2\]
\ssn{d}
$V_{n+1}$ can be written as a block matrix $\openm V_n&e\\e^T&\var(X_{n+1})\closem$. The $e$ vector has components $e_i=\cov(X_i,X_{n+1})=\cov(X_i,Y_{n+1}+Z_{n+1})=\sigma_1^2+(i-1)\sigma_u^2$ for $i\in[1..n]$. $b_{n+1}$ is a $n+1$-long vector, with $i$th component equal to $\cov(X_i, Y_{n+1})=\sigma_1^2+(i-1)\sigma_u^2$ for $i\neq n-1$ and $\cov(X_{n+1},Y_{n+1})=\var(Y_{n+1})=a_n+\sigma_u^2$ for the $n+1$th component. Finally, we have $a_{n+1}=a_n+\sigma_u^2$. 
\ssn{e}
Assume that the covariance $C_{n+1}$ is available -- there exists an efficient recursive algorithm to do so. Write it as the block matrix
\[\openm V_n&e&b_{n+1}^{(1:n)}\\e^T&\var(X_{n+1})&b_{n+1}^{(n+1)}\\{b_{n+1}^T}^{(1:n)}&b_{n+1}^{(n+1)}&a_{n+1}\closem\]
where $e$ is defined as above. Then, consider the joint distribution of $X_1,\ldots,X_{n+1},Y_{n+1}$. Then, the conditional mean of $\openm X_{n+1}\\Y_{n+1}\closem$ given $X_1,\ldots,X_n$ can be expressed as
\[\openm e^T\\{b_{n+1}^T}^{(1:n)}\closem V_n^{-1}\openm x_1\\\vdots\\x_n\closem\]
so the prediction for $X_{n+1}$ would be $e^TV_n^{-1}x$ and for $Y_{n+1}$ would be ${b_{n+1}^T}^{(1:n)}V_n^{-1}x$.
\end{document}

### 2014-05-18 stat246/hw6.tex
s5

\documentclass{article}
\usepackage{geometry}
\usepackage[namelimits,sumlimits]{amsmath}
\usepackage{amssymb,amsfonts}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage[cm]{fullpage}
\usepackage{mathrsfs}
\newcommand{\tab}{\hspace*{5em}}
\newcommand{\conj}{\overline}
\newcommand{\dd}{\partial}
\newcommand{\ep}{\epsilon}
\newcommand{\openm}{\begin{pmatrix}}
\newcommand{\closem}{\end{pmatrix}}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\Null}{null}
\newcommand{\nc}{\newcommand}
\newcommand{\rn}{\mathbb{R}}
\nc{\cn}{\mathbb{C}}
\nc{\ssn}[1]{\subsubsection*{#1}}
\nc{\inner}[2]{\langle #1,#2\rangle}
\nc{\h}[1]{\widehat{#1}}
\nc{\tl}[1]{\widetilde{#1}}
\nc{\norm}[1]{\left\|{#1}\right\|}
\begin{document}

Name: Hall Liu

Date: \today 
\vspace{1.5cm}
\ssn{1}
First, suppose that $C$ is empty. Then, the graph is disconnected, and $A$ and $B$ lie in different connected components, which means that no maximal cliques contain both elements from $A$ and elements from $B$. We can then factor the pdf into one part containing only variables from $A$ and not $B$ and one part only variables from $B$ and not $A$, so $A$ and $B$ are independent. If $C$ is nonempty, let $\conj{A}$ be the set of elements reachable from $A$ without going through $C$ (this excludes $C$ itself). Note that $A\subset \conj{A}$ and $B\cap\conj{A}=\emptyset$ by the assumption that $C$ separates $A$ and $B$. Further, every maximal clique which contains elements of $\conj{A}$ must not contain elements of $\conj{A}^c$ which are not in $C$ (this would violate the separation criterion) and conversely any maximal clique which contains elements of $\conj{A}^c$ must not contain any elements of $\conj{A}$ unless all the elements of $\conj{A}^c$ in that clique are also in $C$.

Thus, we have that any maximal clique must be either a subset of $\conj{A}\cup C$ or a subset of $\conj{A}^c$, which means that the probability factors into two pieces containing variables from $\conj{A}\cup C$ and $\conj{A}^c$. We can further decompose $\conj{A}^c$ into $(\conj{A}^c-C)\cup C$, the former of which contains $B$. Thus, if we integrate over the complement of $C$ and divide to get the conditional, we get a factored distribution, one piece of which contains all elements of $A$ and no elements of $B$, and the other contains all elements of $B$ and no elements of $A$, which means that we have conditional independence.
\end{document}

### 2014-05-19 stat246/hw6.tex
s6

@@
 \vspace{1.5cm}
-\ssn{1}
+\subsection*{1}
 First, suppose that $C$ is empty.
@@
 Thus, if we integrate over the complement of $C$ and divide to get the conditional, we get a factored distribution, one piece of which contains all elements of $A$ and no elements of $B$, and the other contains all elements of $B$ and no elements of $A$, which means that we have conditional independence.
+\subsection*{2}
+\ssn{a}
+We have $p(x_0,x_1,x_2)=\frac{1}{Z}e^{\alpha x_0x_1}e^{\alpha x_1x_2}$.
+Summing over $x_0$ gives $\frac{1}{Z}e^{\alpha x_1x_2}\left(e^{\alpha x_1}+1\right)$.
+Summing over $x_2$ gives $\frac{1}{Z}\left(e^{\alpha x_1}+1\right)^2$, and summing this over $x_1$ gives $\frac{1}{Z}\left(\left(e^\alpha+1\right)^2+4\right)$, so $Z=\left(e^\alpha+1\right)^2+4$.
+
+Summing the joint over $x_2$ gives $\frac{1}{Z}e^{\alpha x_0x_1}\left(e^{\alpha x_1}+1\right)$, and summing this over $x_1$ gives $\frac{1}{Z}\left(2+e^{\alpha x_0}\left(e^{\alpha}+1\right)\right)$, which is the marginal of $x_0$.
+\ssn{b}
+As $\alpha\to0$, the joint distribution becomes uniform. As $\alpha\to\infty$, the joint distribution becomes a single peak at $(1,1,1)$.
+\ssn{c}
+We have $M^2=\openm2&e^\alpha+1\\e^\alpha+1&e^{2\alpha}+1\closem$. If we sum all the elements of $M^2$, we obtain $2+2e^\alpha+2+e^{2\alpha}+1=4+\left(e^{\alpha}+1\right)^2$.
+\ssn{d}
+Define two recursions $C_0=e^\alpha+1$, $C_i=2+\sum_{j=0}^{i-2}C_{j}+e^\alpha C_{i-1}$ and $D_i=2+\sum_{j=0}^{i-2}C_j+e^{\alpha x_{i+1}}C_{i-1}$.
+The claim is that for $n\geq3$ variables, the marginal distribution of $x_i,\ldots,x_{n-1}$ is $D_{i-1}\prod_{j=i}^{n-2}e^{\alpha x_jx_{j+1}}$.
+To prove this, induct on $i$.
+For $i=1$, the marginal distribution comes from summing out $x_0$, which produces $\prod_{j=1}^{n-2}e^{\alpha x_jx_{j+1}}(e^{\alpha x_1}+1)$.
+For the inductive step, assume the marginal distribution on $x_i,\ldots,x_{n-1}$.
+Then, summing over $x_i$ produces
+\begin{align*}
+    &\sum_{x_i}\left(\prod_{j=i+1}^{n-2}e^{\alpha x_jx_{j+1}}\right)e^{\alpha x_ix_{i+1}}\left(2+\sum_{j=0}^{i-3}C_j+e^{\alpha x_{i}}C_{i-2}\right)\\
+    &=\left(\prod_{j=i+1}^{n-2}e^{\alpha x_jx_{j+1}}\right)\left(\left(2+\sum_{j=0}^{i-3}C_j+C_{i-2}\right)+e^{\alpha x_{i+1}}\left(2+\sum_{j=0}^{i-3}C_j+e^{\alpha}C_{i-2}\right)\right)\\
+    &=\left(\prod_{j=i+1}^{n-2}e^{\alpha x_jx_{j+1}}\right)\left(2+\sum_{j=0}^{i-2}C_j+e^{\alpha x_{i+1}}C_{i-1}\right)\\
+    &=D_i\left(\prod_{j=i+1}^{n-2}e^{\alpha x_jx_{j+1}}\right)\\
+\end{align*}
+which ends the proof.
+Thus, we have that the marginal distribution on $x_{n-1}$ is $D_{n-2}$, so the normalizing constant can be obtained by summing the two values of $x_{n-1}$.
+Note that the marginal on $x_0$ has the same form as that of $x_{n-1}$ due to symmetry -- the notation is less cumbersome this way.
+To show the relation to $M$, we have that
+\[M^3=\openm 2+C_0&C_1\\C_1&C_0+e^\alpha C_1-C_1+2\closem\]
+and in general
+\[M^i=\openm 2+\sum_{j=0}^{i-3}C_j&C_{i-2}\\C_{i-2}&2+\sum_{j=0}^{i-3}C_j-C_{i-2}+e^\alpha C_{i-2}\closem\]
+This identity is straightforward to show by induction but cumbersome in notation, so I'm omitting it. Now, given this, we have that the sum of the entries of $M^i$ amounts to
+\[2\left(2+\sum_{j=0}^{i-3}C_j\right)+e^\alpha C_{i-2}+C_{i-2}\]
+which is the sum of $D_{i-1}$ over the two values of $x_i$. Thus, the normalizing constant is the sum of the entries of $M^{n-1}$, which we can see by setting $i=n-1$.
+\ssn{e}
+The formula follows from the application of the definition of conditional probability, starting from the right side and collapsing leftwards.
+Using the conditional independence property, we have that $x_{i}\perp x_{i-2},\ldots,x_1|x_{i-1}$ for all $2\leq i\leq n-1$, so $p(x_{i}|x_{i-1},\ldots,x_0)=p(x_{i}|x_{i-1})$ for each term in the product.
+Since each variable depends only on the one preceding it, we have that the joint distribution is a Markov chain.
+
+The transitions are not uniform -- take the $n=3$ case.
+We have $p(x_1|x_0)=\frac{e^{\alpha x_0x_1}(e^{\alpha x_1}+1)}{2+e^{\alpha x_0}(e^\alpha+1)}$ and $p(x_2|x_1)=\frac{e^{\alpha x_1x_2}}{e^{\alpha x_1}+1}$.
+If we plug in $(1,1)$ to both of these, we get something different.
+\ssn{f}
+Integrating, we have 
+\[\int_\rn\int_\rn\int_\rn e^{\alpha x_0x_1}e^{\alpha x_1x_2}\,dx_0dx_2dx_1\]
+This integral diverges, so we cannot find a normalizing constant. If we add the individual clique functions, the joint distribution becomes
+\[\frac{1}{Z}\exp\left(-\beta x_0^2+\alpha x_0x_1-\beta x_1^2+\alpha x_1x_2-\beta x_2^2\right)\]
+This has the form of a multivariate normal distribution with mean $0$ and precision matrix 
+\[P=\openm-2\beta&\alpha&0\\\alpha&-2\beta&\alpha\\0&\alpha&-2\beta\closem\]
+which means that the normalizing constant is $\frac{1}{Z}=\frac{\sqrt{|P|}}{(2\pi)^{3/2}}$. The constraint on the value of $\beta$ is so that $P$ is a positive-definite matrix.
+\subsection*{3}
+\ssn{a}
+\vspace{200pt}
+\ssn{b}
+There are no cycles, so the largest cliques are the edges. The general form of the distribution is 
+\[\frac{1}{Z}\exp\left(\psi_{12}+\psi_{23}+\psi_{34}+\psi_{35}+\psi_{56}\right)\]
+where each $\psi_{ij}$ takes $x_i$ and $x_j$ as arguments.
+\ssn{c}
+The state space has $K^6$ elements, so that's the number of operations needed (up to a constant factor involved in evaluating the $\psi$s.
+\ssn{d}
+Sum the joint distribution (without normalization) over $x_1$, $x_6$, and $x_4$ to obtain
+\[\exp(\psi_{23}+\psi_{35})\left(\sum_{x_1}\exp(\psi_{12})\right)\left(\sum_{x_4}\exp(\psi_{34})\right)\left(\sum_{x_6}\exp(\psi_{56})\right)\]
+Define $\alpha_2(x_2)=\log\left(\sum_{x_1}\exp(\psi_{12}(x_1,x_2))\right)$, and note that we can compute all the possible values of $\alpha_2$ in $K^2$ operations and save them for later use ($K$ space usage).
+Similarly define $\alpha_3(x_3)$ and $\alpha_5(x_5)$.
+
+Then, we can rewrite the above expression as
+\[\exp(\alpha_2+\psi_{23}+\alpha_3+\psi_{35}+\alpha_5)\]
+Summing over $x_5$, we obtain
+\[\exp(\alpha_2+\psi_{23}+\alpha_3)\sum_{x_5}\exp(\psi_{35}(x_3,x_5)+\alpha_5(x_5))\]
+Define $\beta_3(x_3)$ to be the log of the sum above, so we can rewrite as $\exp(\alpha_2+\psi_{23}+\alpha_3+\beta_3)$.
+Again, the values of $\beta_3$ on all possible values of $x_3$ take $K^2$ operations to compute (so we're up to $4K^2$ now).
+
+Finally, define $\gamma_3(x_3)$ by summing over $x_2$ and repeating the above process (an additional $K^2$ operations), so that we're left with $\exp(\alpha_3+\beta_3+\gamma_3)$, which we can sum over $x_3$ to obtain the normalizing constant, for a total of $6K^2$ operations.
 \end{document}

### 2014-05-27 stat246/hw7.tex
s7

\documentclass{article}
\usepackage{geometry}
\usepackage[namelimits,sumlimits]{amsmath}
\usepackage{amssymb,amsfonts}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage[cm]{fullpage}
\usepackage{mathrsfs}
\newcommand{\tab}{\hspace*{5em}}
\newcommand{\conj}{\overline}
\newcommand{\dd}{\partial}
\newcommand{\ep}{\epsilon}
\newcommand{\openm}{\begin{pmatrix}}
\newcommand{\closem}{\end{pmatrix}}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\Null}{null}
\newcommand{\nc}{\newcommand}
\newcommand{\rn}{\mathbb{R}}
\nc{\cn}{\mathbb{C}}
\nc{\ssn}[1]{\subsubsection*{#1}}
\nc{\inner}[2]{\langle #1,#2\rangle}
\nc{\h}[1]{\widehat{#1}}
\nc{\tl}[1]{\widetilde{#1}}
\nc{\norm}[1]{\left\|{#1}\right\|}
\begin{document}

Name: Hall Liu

Date: \today 
\vspace{1.5cm}

\subsection*{1}
\ssn{a}
Suppose $j$ is a neighbor of $i$. Then, there exists some clique that contains both $i$ and $j$, so $j$ is in the union of the cliques containing $i$. Conversely, suppose $j$ is in the union of cliques containing $i$. Then, $i$ and $j$ coexist in a clique together, so they have an edge between them, which means that $j$ is a neighbor of $i$.
\ssn{b}
Denote the set of cliques which contain $X_i$ by $C(i)$ and its complement in the set of cliques by $C(-i)$. Then, we have that $P(X_i|X_{N_i})=P(X_i|X_{-i})$ (where $X_{-i}$ is all nodes except $X_i$), which can be written as 
\[\frac{\exp\left(\beta\sum_l\psi_{C_l}(X_{C_l})\right)}{\exp\left(\beta\sum_{C\in C(-i)}\psi_C(X_C)\right)\sum_{X_i}\exp\left(\beta\sum_{C\in C(i)}\psi_C(X_C)\right)}=\frac{\exp\left(\beta\sum_{C\in C(i)}\psi_{C}(X_{C})\right)}{\sum_{X_i}\exp\left(\beta\sum_{C\in C(i)}\psi_C(X_C)\right)}\]
Using this, then, we can write the log-pseudolikelihood as
\[\sum_{n=1}^N\sum_{i=1}^d\left(\beta\sum_{C\in C(i)}\psi_{C}(X_{C})-\log\left(\sum_{X_i}\exp\left(\beta\sum_{C\in C(i)}\psi_C(X_C)\right)\right)\right)\]
where the summation over $n$ operates via an implicit superscript $(n)$ over each datum $X$. Differentiating wrt $\beta$ and setting to $0$ gives 
\[0=\sum_{n=1}^N\sum_{i=1}^d\left(\sum_{C\in C(i)}\psi_{C}(X_{C})-\frac{\sum_{X_i}\left(\sum_{C\in C(i)}\psi_C(X_C)\right)\exp\left(\beta\sum_{C\in C(i)}\psi_C(X_C)\right)}{\sum_{X_i}\exp\left(\beta\sum_{C\in C(i)}\psi_C(X_C)\right)}\right)\]
It doesn't look like this can be simplified any further without knowing the forms of the cliques.
\ssn{c}

\end{document}

### 2014-05-28 stat246/hw7.tex
s7

@@
 \ssn{c}
+Let $f(X_i,X_{N_i})=\sum_{C\in C(i)}\psi_C(X_C)$. Then, we can rewrite the equation from (b) as
+\[\sum_{n=1}^N\sum_{i=1}^df(X_i, X_{N_i})=\sum_{n=1}^N\sum_{i=1}^d\frac{\sum_{x_i}\left(f(x_i,X_{N_i})\exp\left(\beta f(x_i,X_{N_i})\right)\right)}{\sum_{x_i}\exp\left(\beta f(x_i,X_{N_i})\right)}\]
+First, examine the LHS. If we look at the sum as being over the different possible configurations of $X_i$ and $X_{N_i}$ rather than being over the data points, we can rewrite it as
+\[\sum_{i=1}^d\sum_{x_{N_i}}\sum_{x_i}f(x_i, x_{N_i})\left|\{n:X_i^{(n)}=x_i,X_{N_i}^{(n)}=x_{N_i}\}\right|\]
+where the multiplier is the number of data points which have the configuration specified by $x_i$ and $x_{N_i}$. We can then express this in terms of the empirical expectation as 
+\[\sum_{i=1}^d\sum_{x_{N_i}}\left|\{n:X_{N_i}=x_{N_i}\}\right| E^{(emp)}(f(X_i,X_{N_i})|X_{N_i}=x_{N_i})\]
+
+To pare down the right side, first note that the (non-empirical) expectation is given by
+\[E(f(X_i,X_{N_i})|X_{N_i}=x_{N_i})=\sum_{x_i}f(x_i,x_{N_i})P(x_i|x_{N_i})=\sum_{x_i}f(x_i, x_{N_i})\frac{\exp\left(\beta f(x_i,x_{N_i})\right)}{\sum_{x_i'}\exp\left(\beta f(x_i',x_{N_i})\right)}\]
+Conveniently, this is the same as the inner summand of the RHS, so we can write the RHS as
+\[\sum_{i=1}^d\sum_{n=1}^NE(f(X_i^{(n)},X_{N_i}^{(n)})|X_{N_i})=\sum_{i=1}^d\sum_{x_{N_i}}E(f(X_i,X_{N_i})|X_{N_i}=x_{N_i})\left|\{n:X_{N_i}=x_{N_i}\}\right|\]
+Then, trimming away the leftover counts (which are attached to the same terms on both sides) gives us the desired equality.
+\ssn{d}
+In the same vein as (b), the log-pseudolikelihood is
+\[\sum_{n=1}^N\sum_{i=1}^d\left(\sum_{C\in C(i)}\beta_C\psi_{C}(X_{C})-\log\left(\sum_{X_i}\exp\left(\sum_{C\in C(i)}\beta_C\psi_C(X_C)\right)\right)\right)\]
+Fix a clique $C$ and differentiate wrt $\beta_C$ to obtain
+\[\sum_{n=1}^N\sum_{i\in C}\left(\psi_C(X_C)-\frac{\sum_{x_i}\psi_C(X_C)\exp\left(\sum_{C'\in C(i)}\beta_{C'}\psi_{C'}(X_{C'})\right)}{\sum_{x_i}\exp\left(\sum_{C'\in C(i)}\beta_{C'}\psi_{C'}(X_{C'})\right)}\right)=0\]
+Then, going through the same steps as in (c), we see that the result is the same, but with $f(x_i,X_{N_i})=\psi_C(X_C)$ for every clique $C$ that contains $i$.
+\subsection*{2}
 

### 2014-05-28 stat246/hw7.tex
s7

@@
 \subsection*{2}
-
+Performing the multiplication $P_nP=P_{n+1}$ to get a recurrence, we have
+\[\openm1-a_n&a_n\\b_n&1-b_n\closem\openm1-a&a\\b&1-b\closem=\openm1-(a+a_n(1-a-b))&a+a_n(1-a-b)\\b+b_n(1-a-b)&1-(b+b_n(1-a-b))\closem\]
+so the recurrence is $a_{n+1}=a+a_n(1-a-b)$ and similar for $b_{n+1}$.
+To find the steady-state solution, set $a_n=a_{n+1}=a^*$ to obtain $a^*=a+a^*(1-a-b)$ or $a^*=\frac{a}{a+b}$ (and similarly $b^*=\frac{b}{a+b}$).
+This gives us the limit of $P^n$ as
+\[\openm\frac{b}{a+b}&\frac{a}{a+b}\\\frac{b}{a+b}&\frac{a}{a+b}\closem\]
+To find closed-form formulas for $a_n$ and $b_n$, note that $a_{n+1}-a^*=(a_n-a^*)(1-a-b)$, or $a_n-a^*=(1-a-b)^{n-1}(a-a^*)$. The same goes for $b_n$.
 \end{document}

### 2014-05-28 stat246/hw7.tex
s7

@@
 To find closed-form formulas for $a_n$ and $b_n$, note that $a_{n+1}-a^*=(a_n-a^*)(1-a-b)$, or $a_n-a^*=(1-a-b)^{n-1}(a-a^*)$. The same goes for $b_n$.
+\subsection*{3}
+\ssn{a}
+\vspace{130pt}
+\ssn{b}
+To construct $Q$ (for $n\geq3$), start with a tridiagonal matrix with $\beta$ on the diagonal and $-\alpha$ on the off-diagonals, then place $-\alpha$ at the upper-right and lower-left corners.
+Then, taking $x^TQx$ results in
+\[\sum_{i=1}^nx_i(-\alpha x_{i-1}+\beta x_i-\alpha x_{i+1})\]
+which simplifies to $\sum_{i=1}^n-2\alpha x_ix_{i+1}+\beta x_i^2$ when we combine the like terms across summands. Multiplying by $-\frac{1}{2}$ gives us the term in the exponent.
+\ssn{c}
+The $l$th entry of $Qv_k$ is $\frac{\beta}{\sqrt{n}}\exp(2\pi i\frac{l}{n}k)-\frac{\alpha}{\sqrt{n}}\left(\exp(2\pi i\frac{l-1}{n}k)+\exp(2\pi i\frac{l+1}{n}k)\right)$.
+This is a real multiple of $\exp(2\pi i\frac{l}{n}k)$ since adding the two $\alpha$-multiplied vectors results in something with the same argument.
+For $k$ fixed, the multiplier is the same for every $l$, and its value is $\frac{\beta-2\alpha\cos(2\pi k/n)}{\sqrt{n}}$.
+Thus, the $v_k$ are eigenvectors.
+Further, since we know that $\beta>2\alpha$, all the eigenvalues are positive, which implies that $Q$ is strictly positive definite.
+\ssn{d}
+Since this density is invariant under cyclically permuting the variables, all that's needed is to find the distribution of $x_1$ conditioned on all the others.
+Write the covariance $Q$ as $\openm \beta&Q_{12}\\Q_{21}&Q_{22}\closem$.
+Then, if the other variables taken together equal some $n-1$-long vector $y$, the conditional mean of $x_1$ is $Q_{12}Q_{22}^{-1}y$ and the conditional variance is $\beta-Q_{12}Q_{22}^{-1}Q_{21}$.
+For all the variables, $Q_{22}$ will be a $n-1\times n-1$ tridiagonal matrix having $\beta$ on the diagonal and $-\alpha$ on the off-diagonal.
 \end{document}

### 2014-05-29 proj/broker.py
..

@@ -311,6 +311,8 @@ class Broker:
     if not hasattr(self, "devnull"):
       self.devnull = open(os.devnull, "w")
 
+
+
     proc = subprocess.Popen(args, shell=True, stdout=self.devnull, stderr=self.devnull)
     self.node_pids[command['name']] = proc
 

### 2014-05-29 proj/a.py
basic functionality

@@ -1,0 +1,71 @@
+from Queue import Queue
+import random
+import json
+import sys
+import signal
+import time
+import zmq
+from zmq.eventloop import ioloop, zmqstream
+import argparse
+ioloop.install()
+
+class Node(object):
+    def __init__(self, name, pub_endpoint, router_endpoint, peer_names):
+        self.loop = ioloop.ZMQIOLoop.current()
+        self.context = zmq.Context()
+    
+        # SUB socket for receiving messages from the broker
+        self.sub_sock = self.context.socket(zmq.SUB)
+        self.sub_sock.connect(pub_endpoint)
+        # make sure we get messages meant for us!
+        self.sub_sock.set(zmq.SUBSCRIBE, name)
+        self.sub = zmqstream.ZMQStream(self.sub_sock, self.loop)
+        self.sub.on_recv(self.msg_handler)
+    
+        # REQ socket for sending messages to the broker
+        self.req_sock = self.context.socket(zmq.REQ)
+        self.req_sock.connect(router_endpoint)
+        self.req = zmqstream.ZMQStream(self.req_sock, self.loop)
+        self.req.on_recv(self.handle_broker_message)
+
+        self.name = name
+        self.peers = peer_names
+        self.stored_data = {}
+
+    def start(self):
+        self.loop.start()
+
+    def handle_broker_message(self, frames):
+        pass
+
+    def msg_handler(self, frames):
+        msg = json.loads(frames[2])
+        if msg['type'] == 'hello':
+            self.req.send_json({'type': 'helloResponse', 'source': self.name})
+            self.req.send_json({'type': 'log', 'message': 'hello received'})
+        else:
+            self.req.send_json({'type': 'log', 'data': msg})
+        
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--pub-endpoint',
+        dest='pub_endpoint', type=str,
+        default='tcp://127.0.0.1:23310')
+    parser.add_argument('--router-endpoint',
+        dest='router_endpoint', type=str,
+        default='tcp://127.0.0.1:23311')
+    parser.add_argument('--node-name',
+        dest='node_name', type=str,
+        default='test_node')
+    parser.add_argument('--peer-names',
+        dest='peer_names', type=str,
+        default='')
+    args = parser.parse_args()
+    args.peer_names = args.peer_names.split(',')
+    
[…]

### 2014-05-29 proj/broker.py

@@ -186,7 +186,7 @@ class Broker:
     '''
     Print or log to a file the given message, along with sender.
     '''
-    # self.log_message(message)
+    self.log_message(message)
     return Message({'type': 'okay'})
 
   def make_handle_response(self, ty):
@@ -310,8 +310,6 @@ class Broker:
 
     if not hasattr(self, "devnull"):
       self.devnull = open(os.devnull, "w")
-
-
 
     proc = subprocess.Popen(args, shell=True, stdout=self.devnull, stderr=self.devnull)
     self.node_pids[command['name']] = proc

### 2014-05-29 proj/a.py
beginning raft impl

@@ -3,7 +3,7 @@ import random
 import json
 import sys
 import signal
-import time
+import datetime
 import zmq
 from zmq.eventloop import ioloop, zmqstream
 import argparse
@@ -30,22 +30,33 @@ class Node(object):
 
         self.name = name
         self.peers = peer_names
+        self.hello_sent = False
+
+
+        # Node's Raft state (possibly to be augmented by DHT later?)
+        self.raft_peers = self.peers.copy()
+        self.raft_state = 'follower'
+        self.raft_term = 0
+
         self.stored_data = {}
 
     def start(self):
         self.loop.start()
+        # Initialize the election timeout
+        self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = random.randint(150, 300)), self.leader_timeout)
 
     def handle_broker_message(self, frames):
         pass
 
     def msg_handler(self, frames):
         msg = json.loads(frames[2])
-        if msg['type'] == 'hello':
+        if msg['type'] == 'hello' and not self.hello_sent:
             self.req.send_json({'type': 'helloResponse', 'source': self.name})
+            self.hello_sent = True
             self.req.send_json({'type': 'log', 'message': 'hello received'})
-        else:
-            self.req.send_json({'type': 'log', 'data': msg})
-        
+        if msg['type'] == 'RequestVote':
+            pass
+
 if __name__ == '__main__':
     parser = argparse.ArgumentParser()
     parser.add_argument('--pub-endpoint',

### 2014-05-29 broker.py
Squashed 'proj/' changes from fe6de3b..16979bd
16979bd Documentation, fix partition
0aa6349 Finish partition command
59c28b0 Documentation
4cb7829 Add delay command
2fb2cf7 Finish tamper command
04791d3 Finish after command
dce0a09 Implement drop
7395dd3 Add parsing for first set of unreliability commands

git-subtree-dir: proj
git-subtree-split: 16979bd5ad843c043c58942420b05ae9ef337168

@@ -41,9 +41,149 @@ class Message(dict):
     msg_frames = [destination, '', b]
     socket.send_multipart(msg_frames)
 
+class MessageConditions:
+  def __init__(self, broker):
+    self.broker = broker
+    self.drop_conditions = []
+    self.delay_conditions = []
+    self.tamper_conditions = []
+    self.after_conditions = []
+
+  def add_condition(self, command):
+    if (command['command'] == 'drop'):
+      self.drop_conditions.append(command)
+    elif command['command'] == 'after':
+      self.after_conditions.append(command)
+    elif command['command'] == 'delay':
+      command['messages'] = []
+      self.delay_conditions.append(command)
+    elif command['command'] == 'tamper':
+      self.tamper_conditions.append(command)
+
+  def check_conds(self, conds, message):
+    '''
+    Generic condition checker. Returns (all_match, destinations) where
+    all_match is True iff the message as a whole matches the condition, and
+    destinations is a set containing specific recipients for which it matches.
+    '''
+    all_match = False
+    destinations = set()
+    for cond in conds:
+      if cond['count'] == 0:
+        conds.remove(cond)
+        continue
+      m = self.matches(cond, message)
+      (any_match, sender_match, destination_matches) = m
+      if any_match or sender_match:
+        tamper_all = True
+      destinations = destinations.union(destination_matches)
+      if any(m):
+        cond['count'] -= 1
+
+    return (all_match, destinations)
+
+  def check_drop_conditions(self, message):
+    '''
+    Returns (should_drop, should_receive) where should_drop is true iff the
+    message's sender has messages to be blocked, and should_receive is a list
+    of nodes that should receive the message. Updates drop conditions
+    accordingly.
+    '''
+    should_receive = set(message['destination'])
+    should_drop, should_not_receive = self.check_conds(self.drop_conditions, message)
+    should_receive.difference_update(should_not_receive)
+    return (should_drop, should_receive)
+
+  def check_tamper_conditions(self, message):
+    '''
+    Returns (tamper_all, tamper_destinations) where tamper_all is True iff
+    every message should be tampered and tamper_destinations is a set of
+    destinations that should be tampered with.
+    '''
+    return self.check_conds(self.tamper_conditions, message)
+
+  def check_after_conditions(self, message):
+    '''
+    Tallies against existing after conditions, and pushes their commands to the
+    broker's script queue for any that hits zero.
+    '''
+    for cond in self.after_conditions:
[…]

### 2014-05-29 proj/a.py
halfway through voting

@@ -7,6 +7,7 @@ import datetime
 import zmq
 from zmq.eventloop import ioloop, zmqstream
 import argparse
+from copy import copy
 ioloop.install()
 
 class Node(object):
@@ -31,12 +32,26 @@ class Node(object):
         self.name = name
         self.peers = peer_names
         self.hello_sent = False
-
+        
+        self.message_handlers = {
+                'hello': self.helloRespond,
+                'RequestVote': self.handle_reqvote
+        }
 
         # Node's Raft state (possibly to be augmented by DHT later?)
-        self.raft_peers = self.peers.copy()
+        self.raft_peers = copy(self.peers)
         self.raft_state = 'follower'
+
         self.raft_term = 0
+        self.raft_lastvote = None
+        self.raft_commitIndex = 0
+        self.raft_lastApplied = 0
+
+        self.raft_lastLogIndex = 0
+        self.raft_lastLogTerm = 0
+
+        self.raft_nextIndex = None
+        self.raft_matchIndex = None
 
         self.stored_data = {}
 
@@ -50,12 +65,34 @@ class Node(object):
 
     def msg_handler(self, frames):
         msg = json.loads(frames[2])
-        if msg['type'] == 'hello' and not self.hello_sent:
+        handler_fn = self.message_handlers.get(msg['type'], self.log_info)
+        handler_fn(msg)
+
+    '''
+    Here follows the different functions that get triggered
+    '''
+    def helloRespond(self, msg):
+        if not self.hello_sent:
             self.req.send_json({'type': 'helloResponse', 'source': self.name})
             self.hello_sent = True
             self.req.send_json({'type': 'log', 'message': 'hello received'})
-        if msg['type'] == 'RequestVote':
-            pass
+
+    def leader_timeout(self):
+        self.log_info('Server {0} hit election timeout'.format(self.name))
+        self.raft_state = 'cand'
+        self.raft_term += 1
+        reqvote_msg = {
+                'type': 'RequestVote',
+                'term': self.raft_term,
+                'lastLogIndex': self.raft_lastLogIndex,
+                'raft_lastLogTerm': self.raft_lastLogTerm
+        }
+        self.req.send_json(reqvote_msg)
+
+    def handle_reqvote(self, msg):
+
+    def log_info(self, s):
+        self.req.send_json({'type': 'log', 'info': s})
 
 if __name__ == '__main__':
     parser = argparse.ArgumentParser()

### 2014-05-30 proj/a.py
more progress on voting, about to implement becoming leader

@@ -35,7 +35,8 @@ class Node(object):
         
         self.message_handlers = {
                 'hello': self.helloRespond,
-                'RequestVote': self.handle_reqvote
+                'RequestVote': self.handle_reqvote,
+                'VoteReply': self.handle_votereply
         }
 
         # Node's Raft state (possibly to be augmented by DHT later?)
@@ -49,6 +50,7 @@ class Node(object):
 
         self.raft_lastLogIndex = 0
         self.raft_lastLogTerm = 0
+        self.raft_timeout = None
 
         self.raft_nextIndex = None
         self.raft_matchIndex = None
@@ -58,7 +60,7 @@ class Node(object):
     def start(self):
         self.loop.start()
         # Initialize the election timeout
-        self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = random.randint(150, 300)), self.leader_timeout)
+        self.raft_timeout = self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = random.randint(150, 300)), self.leader_timeout)
 
     def handle_broker_message(self, frames):
         pass
@@ -79,17 +81,85 @@ class Node(object):
 
     def leader_timeout(self):
         self.log_info('Server {0} hit election timeout'.format(self.name))
-        self.raft_state = 'cand'
+        self.raft_state = 'candidate'
         self.raft_term += 1
         reqvote_msg = {
                 'type': 'RequestVote',
+                'source': self.name,
                 'term': self.raft_term,
                 'lastLogIndex': self.raft_lastLogIndex,
-                'raft_lastLogTerm': self.raft_lastLogTerm
+                'lastLogTerm': self.raft_lastLogTerm
         }
         self.req.send_json(reqvote_msg)
+        self.raft_lastvote = self.name
+        self.raft_numVotes = 1 # Ourselves
 
     def handle_reqvote(self, msg):
+        # Check if term is larger and change own status as appropriate
+        if msg['term'] > self.raft_term:
+            self.revert_state(msg['term'])
+
+        # If term is less, simply reject
+        elif msg['term'] < self.raft_term:
+            reject_msg = {
+                    'type': 'VoteReply',
+                    'destination': msg['source'],
+                    'term': self.raft_term,
+                    'voteGranted': False
+            }
+            self.req.send_json(reject_msg)
+            self.log_info('Rejected vote req from {0} due to lagging term (self: {1}, peer: {2})'.format(msg['source'], self.raft_term, msg['term']))
+            return
+
[…]

### 2014-05-30 proj/a.py
stripped down the raft machine

@@ -8,6 +8,7 @@ import zmq
 from zmq.eventloop import ioloop, zmqstream
 import argparse
 from copy import copy
+from collections import defaultdict
 ioloop.install()
 
 class Node(object):
@@ -43,19 +44,25 @@ class Node(object):
         self.raft_peers = copy(self.peers)
         self.raft_state = 'follower'
 
+        '''
+        The format of each log entry is a dict. The 'term' value is constant among all of them, and
+        all the other entries are to be used by any derived subclasses to adjust their own state.
+        '''
+        self.raft_log = []
+
         self.raft_term = 0
         self.raft_lastvote = None
         self.raft_commitIndex = 0
         self.raft_lastApplied = 0
 
-        self.raft_lastLogIndex = 0
-        self.raft_lastLogTerm = 0
         self.raft_timeout = None
 
+        '''
+        These two variables are only relevant when self is a leader. In that case,
+        they are initialized to defaultdicts which use each server's name as a key.
+        '''
         self.raft_nextIndex = None
         self.raft_matchIndex = None
-
-        self.stored_data = {}
 
     def start(self):
         self.loop.start()
@@ -87,8 +94,8 @@ class Node(object):
                 'type': 'RequestVote',
                 'source': self.name,
                 'term': self.raft_term,
-                'lastLogIndex': self.raft_lastLogIndex,
-                'lastLogTerm': self.raft_lastLogTerm
+                'lastLogIndex': len(self.raft_log),
+                'lastLogTerm': self.raft_log[-1]['term']
         }
         self.req.send_json(reqvote_msg)
         self.raft_lastvote = self.name
@@ -114,7 +121,7 @@ class Node(object):
         # Decide whether to grant the vote and also cancel the timeout
         self.loop.remove_timeout(self.raft_timeout)
         hasNotVoted = self.raft_lastvote is None
-        cand_uptodate = self.raft_lastLogTerm <= msg['lastLogTerm'] and self.raft_lastLogIndex <= msg['raft_lastLogIndex']
+        cand_uptodate = self.raft_log[-1]['term'] <= msg['lastLogTerm'] and len(self.raft_log) <= msg['raft_lastLogIndex']
         if hasNotVoted and cand_uptodate:
             vote_msg = {
                     'type': 'VoteReply',
@@ -160,6 +167,15 @@ class Node(object):
         self.raft_state = 'follower'
         self.raft_term = term
         self.raft_lastvote = None
[…]

### 2014-05-30 proj/a.py
mostly done with leader election, need to fix timeouts'

@@ -118,8 +118,7 @@ class Node(object):
             self.log_info('Rejected vote req from {0} due to lagging term (self: {1}, peer: {2})'.format(msg['source'], self.raft_term, msg['term']))
             return
 
-        # Decide whether to grant the vote and also cancel the timeout
-        self.loop.remove_timeout(self.raft_timeout)
+        # Decide whether to grant the vote
         hasNotVoted = self.raft_lastvote is None
         cand_uptodate = self.raft_log[-1]['term'] <= msg['lastLogTerm'] and len(self.raft_log) <= msg['raft_lastLogIndex']
         if hasNotVoted and cand_uptodate:
@@ -175,7 +174,21 @@ class Node(object):
         self.raft_nextIndex = defaultdict(lambda: len(self.raft_log) + 1)
         self.raft_matchIndex = defaultdict(int)
 
-
+        # Send out a heartbeat AppendEntries to all peers
+        leader_heartbeat = {
+                'type': 'AppendEntries',
+                'destination': self.raft_peers,
+                'source': self.name,
+                'term': self.raft_term,
+                'lastLogIndex': len(self.raft_log),
+                'lastLogTerm': self.raft_log[-1]['term'],
+                'entries': [],
+                'leaderCommit': self.raft_commitIndex
+        }
+        self.req.send_json(leader_heartbeat)
+        # Remove the timeout. Leaders don't need timeouts, because they're cool.
+        if self.raft_timeout is not None:
+            self.loop.remove_timeout(self.raft_timeout)
 
     def log_info(self, s):
         self.req.send_json({'type': 'log', 'info': s})

### 2014-05-30 proj/a.py
fixed the timeout

@@ -78,7 +78,7 @@ class Node(object):
         handler_fn(msg)
 
     '''
-    Here follows the different functions that get triggered
+    Here follows the message handler functions that interface directly with the broker.
     '''
     def helloRespond(self, msg):
         if not self.hello_sent:
@@ -86,6 +86,7 @@ class Node(object):
             self.hello_sent = True
             self.req.send_json({'type': 'log', 'message': 'hello received'})
 
+    # Not strictly a handler, but it responds directly to loop events, so I'm putting it here.
     def leader_timeout(self):
         self.log_info('Server {0} hit election timeout'.format(self.name))
         self.raft_state = 'candidate'
@@ -100,12 +101,9 @@ class Node(object):
         self.req.send_json(reqvote_msg)
         self.raft_lastvote = self.name
         self.raft_numVotes = 1 # Ourselves
+        self.reset_timeout()
 
     def handle_reqvote(self, msg):
-        # Check if term is larger and change own status as appropriate
-        if msg['term'] > self.raft_term:
-            self.revert_state(msg['term'])
-
         # If term is less, simply reject
         elif msg['term'] < self.raft_term:
             reject_msg = {
@@ -117,6 +115,14 @@ class Node(object):
             self.req.send_json(reject_msg)
             self.log_info('Rejected vote req from {0} due to lagging term (self: {1}, peer: {2})'.format(msg['source'], self.raft_term, msg['term']))
             return
+
+        # If this message came from a candidate at least as up-to-date as us, no sense in
+        # going for candidacy again soon. Thus, reset the timeout.
+        self.reset_timeout()
+
+        # Check if term is larger and change own status as appropriate
+        if msg['term'] > self.raft_term:
+            self.revert_state(msg['term'])
 
         # Decide whether to grant the vote
         hasNotVoted = self.raft_lastvote is None
@@ -167,8 +173,12 @@ class Node(object):
         self.raft_term = term
         self.raft_lastvote = None
     
-    # Upon receiving a sufficient number of votes, becomes leader
+    # Upon receiving a sufficient number of votes, become leader using this function
     def become_leader(self):
+        # Remove the timeout. Leaders don't need timeouts, because they're cool.
+        if self.raft_timeout is not None:
+            self.loop.remove_timeout(self.raft_timeout)
+
         # Adjust own state.
         self.raft_state = 'leader'
[…]

### 2014-05-30 proj/a.py
receiver impl of append entries

@@ -37,7 +37,8 @@ class Node(object):
         self.message_handlers = {
                 'hello': self.helloRespond,
                 'RequestVote': self.handle_reqvote,
-                'VoteReply': self.handle_votereply
+                'VoteReply': self.handle_votereply,
+                'AppendEntry': self.handle_appendentry
         }
 
         # Node's Raft state (possibly to be augmented by DHT later?)
@@ -95,7 +96,7 @@ class Node(object):
                 'type': 'RequestVote',
                 'source': self.name,
                 'term': self.raft_term,
-                'lastLogIndex': len(self.raft_log),
+                'lastLogIndex': len(self.raft_log) - 1,
                 'lastLogTerm': self.raft_log[-1]['term']
         }
         self.req.send_json(reqvote_msg)
@@ -108,7 +109,7 @@ class Node(object):
         elif msg['term'] < self.raft_term:
             reject_msg = {
                     'type': 'VoteReply',
-                    'destination': msg['source'],
+                    'destination': [msg['source']],
                     'term': self.raft_term,
                     'voteGranted': False
             }
@@ -126,11 +127,11 @@ class Node(object):
 
         # Decide whether to grant the vote
         hasNotVoted = self.raft_lastvote is None
-        cand_uptodate = self.raft_log[-1]['term'] <= msg['lastLogTerm'] and len(self.raft_log) <= msg['raft_lastLogIndex']
+        cand_uptodate = self.raft_log[-1]['term'] <= msg['lastLogTerm'] and len(self.raft_log) - 1 <= msg['raft_lastLogIndex']
         if hasNotVoted and cand_uptodate:
             vote_msg = {
                     'type': 'VoteReply',
-                    'destination': msg['source']
+                    'destination': [msg['source']],
                     'term': self.raft_term,
                     'voteGranted': True
             }
@@ -140,7 +141,7 @@ class Node(object):
         else:
             reject_msg = {
                     'type': 'VoteReply',
-                    'destination': msg['source'],
+                    'destination': [msg['source']],
                     'term': self.raft_term,
                     'voteGranted': False
             }
@@ -160,8 +161,60 @@ class Node(object):
         # If vote granted, increment our vote count by one, then become leader if majority
         if msg['voteGranted']:
             self.raft_numVotes += 1
+            self.log_info('Vote received, {0} total'.format(self.raft_numVotes))
             if self.raft_numVotes >= len(self.raft_peers) + 1:
[…]

### 2014-05-30 stat246/hw7_files/gibbs.py
gibbs

@@ -1,0 +1,40 @@
+import numpy as np
+import scipy.linalg as sp
+
+def run_gibbs_normal(n_samp, epsilon):
+    r1 = np.array([2+epsilon, -1, 0, -1])
+    Q = sp.circulant(r1)
+    prec = Q
+    samples = np.zeros((4, n_samp), dtype='float64')
+    running_sample = np.zeros(4, dtype='float64')
+
+    # Precompute 40k std normal samples
+    std_normal_samples = np.random.randn(4*n_samp)
+    
+    # Run 10000 iterations over all 4 coordinates
+    for i in range(n_samp):
+        for j in range(4):
+            mean = np.dot(prec[j], running_sample) - prec[j, j]*running_sample[j]
+            mean /= prec[j, j]
+            std = np.sqrt(1/prec[j, j])
+            running_sample[j] = std_normal_samples[i*4+j] * std + mean
+        samples[:, i] = running_sample
+
+    return samples
+
+def est_covs(samples):
+    c1 = samples[:1000].copy()
+    cf = samples[-1000:].copy()
+    c1 -= c1.mean(0)
+    cf -= cf.mean(0)
+    samples -= samples.mean(0)
+
+    return np.dot(c1,c1.T)/999, np.dot(c2,c2.T)/999, np.dot(samples, sample.T)/9999
+
+def main(epsilon):
+    f1000_rhos = np.zeros(100)
+    l1000_rhos = np.zeros(100)
+    total_rhos = np.zeros(100)
+
+    r1 = np.array([2+epsilon, -1, 0, -1])
+    Q = sp.circulant(r1)

### 2014-05-31 stat246/hw7.tex
s7

@@
 \ssn{d}
-Since this density is invariant under cyclically permuting the variables, all that's needed is to find the distribution of $x_1$ conditioned on all the others.
-Write the covariance $Q$ as $\openm \beta&Q_{12}\\Q_{21}&Q_{22}\closem$.
-Then, if the other variables taken together equal some $n-1$-long vector $y$, the conditional mean of $x_1$ is $Q_{12}Q_{22}^{-1}y$ and the conditional variance is $\beta-Q_{12}Q_{22}^{-1}Q_{21}$.
-For all the variables, $Q_{22}$ will be a $n-1\times n-1$ tridiagonal matrix having $\beta$ on the diagonal and $-\alpha$ on the off-diagonal.
+The conditional variance of $x_i$ conditioned upon all the others is $\frac{1}{Q_{ii}}$, and the conditional mean is $-\frac{1}{Q_{ii}}Q_i^Tx$, with the $i$th coordinate being excluded from the inner product.
+\ssn{e}
+Plots of the trajectories for $\ep=0.01$:
+
+\includegraphics[width=0.9\textwidth]{hw7_files/traj_e2.png}
+
+It looks like things converged, since none of the coordinates are moving around much. On the other hand, for $\ep=0.00001$, the chain definitely has not converged to the stationary distribution.
+
+\includegraphics[width=0.9\textwidth]{hw7_files/traj_e5.png}
+
+Calculating $\rho$ from the estimated covariance matrices, at $\ep=0.00001$, the first 1000 yields $\rho=0.9693$, the last 1000 yields $\rho=1.034$, and the entire sample yields $\rho=0.971$.
+At $\ep=0.01$, the figures are $0.265$, $-0.291$, and $0.0157$ in the same order.
+\ssn{f}
+Histograms follow. $\ep=0.01$ is on the left.
+
+\includegraphics[width=0.5\textwidth]{hw7_files/e2_hist.png}
+\includegraphics[width=0.5\textwidth]{hw7_files/e5_hist.png}
+\ssn{g}
+Same order as above.
+
+\includegraphics[width=0.5\textwidth]{hw7_files/e2_direct_hist.png}
+\includegraphics[width=0.5\textwidth]{hw7_files/e5_direct_hist.png}
+
+For the $\ep=0.00001$ case, direct sampling is a lot better, but this comes at the cost of having to do a Cholesky on the precision matrix and inverting the resulting upper-triangular matrix.
+\ssn{h}
+If $x_i$ is always either $1$ or $-1$, then $-\frac{1}{2}\beta x_i^2$ is just $-\frac{\beta}{2}$, so we can write 
+\[f\propto\exp\left(\sum_{i=1}^n\alpha x_ix_{i+1}-\frac{\beta}{2}\right)=\exp\left(-\frac{\beta}{2}\right)\exp\left(\sum_{i=1}^n\alpha x_ix_{i+1}\right)\propto\exp\left(\sum_{i=1}^n\alpha x_ix_{i+1}\right)\]
+so $\beta$ is irrelevant to the distribution.
+
+The conditional is simply the joint distribution divided by the marginal over everything but $x_i$, which can be found by evaluating the joint distribution at $x_i=1$ and at $x_i=-1$.
+For $x_1$, this marginal is
+\[\exp\left(\alpha(x_2+x_2x_3+x_3x_4+x_4)\right)+\exp\left(\alpha(-x_2+x_2x_3+x_3x_4-x_4)\right)\]
+
+The actual probabilities are given in the table below ($-1$s are notated as zeros for readability):
+
+{\tiny
+    \noindent
+\begin{tabular}{r*{16}{c}}
+    0000&0001&0010&0011&0100&0101&0110&0111&1000&1001&1010&1011&1100&1101&1110&1111\\
+    $0.45$&$0.00825$&$0.00825$&$0.00825$&$0.00825$&$0.000151$&$0.00825$&$0.00825$&$0.00825$&$0.00825$&$0.000151$&$0.00825$&$0.00825$&$0.00825$&$0.00825$&$0.45$\\
+\end{tabular}
+}
+
+Most likely states are the $0000$ state and the $1111$ state.
+
+A plot of trajectories is shown below, with the components not labeled nor differentiated.
+There are 16 lines, each representing the proportion of its corresponding component.
+The $x$-axis indicates the number of samples taken.
+
+\includegraphics[width=0.6\textwidth]{hw7_files/bin_traj.png}
+
+Looks like the empirical frequencies are converging to the true probabilities.
+\ssn{i}
+The sign in front of the $\beta$ term doesn't matter, since we showed above that the value of $\beta$ doesn't even matter.
+$\Sigma$ is the same as $Q$ from part (b), except the $\alpha$s no longer carry a negative sign with them.
+This matrix is still positive definite, since if $\beta>2\alpha$, it is strictly diagonally dominant.
+
+The integral given is the expectation of $\exp(x^Ty)$ under the distribution $g$, which is the moment-generating function evaluated at $x$.
+The mgf of $g$ is $\exp\left(\frac{1}{2}x^T\Sigma x\right)$, which is proportional to $f$ from above.
+
+Given this, another possible way to sample from $f$ would be to sample first from $g$, then convert all the positive entries to $1$ and all the negative ones to $-1$.
 \end{document}

### 2014-05-31 stat246/hw7_files/gibbs.py

@@ -1,10 +1,10 @@
 import numpy as np
 import scipy.linalg as sp
+import matplotlib.pyplot as plt
 
 def run_gibbs_normal(n_samp, epsilon):
     r1 = np.array([2+epsilon, -1, 0, -1])
-    Q = sp.circulant(r1)
-    prec = Q
+    prec = sp.circulant(r1)
     samples = np.zeros((4, n_samp), dtype='float64')
     running_sample = np.zeros(4, dtype='float64')
 
@@ -15,26 +15,125 @@ def run_gibbs_normal(n_samp, epsilon):
     for i in range(n_samp):
         for j in range(4):
             mean = np.dot(prec[j], running_sample) - prec[j, j]*running_sample[j]
-            mean /= prec[j, j]
+            mean /= -prec[j, j]
             std = np.sqrt(1/prec[j, j])
             running_sample[j] = std_normal_samples[i*4+j] * std + mean
         samples[:, i] = running_sample
 
     return samples
 
-def est_covs(samples):
-    c1 = samples[:1000].copy()
-    cf = samples[-1000:].copy()
-    c1 -= c1.mean(0)
-    cf -= cf.mean(0)
-    samples -= samples.mean(0)
+def plot_trajectories(samples):
+    plt.figure(figsize=(14,10))
+    plt.subplot(221)
+    plt.plot(samples[0])
+    plt.subplot(222)
+    plt.plot(samples[1])
+    plt.subplot(223)
+    plt.plot(samples[2])
+    plt.subplot(224)
+    plt.plot(samples[3])
 
-    return np.dot(c1,c1.T)/999, np.dot(c2,c2.T)/999, np.dot(samples, sample.T)/9999
+def est_covs(samples, epsilon):
+    c1 = samples[:,:1000]
+    cf = samples[:,-1000:]
+    
+    r1 = np.array([2+epsilon, -1, 0, -1])
+    Q = sp.circulant(r1)
 
-def main(epsilon):
-    f1000_rhos = np.zeros(100)
-    l1000_rhos = np.zeros(100)
+    return 4-np.trace(np.dot(Q,np.cov(c1))), 4-np.trace(np.dot(Q, np.cov(cf))), 4-np.trace(np.dot(Q,np.cov(samples)))
+
+
+def make_gibbs_hist(epsilon):
     total_rhos = np.zeros(100)
 
     r1 = np.array([2+epsilon, -1, 0, -1])
[…]

### 2014-06-02 proj/a.py
finished receiver impl

@@ -53,7 +53,7 @@ class Node(object):
 
         self.raft_term = 0
         self.raft_lastvote = None
-        self.raft_commitIndex = 0
+        self.raft_commitIndex = -1
         self.raft_lastApplied = 0
 
         self.raft_timeout = None
@@ -172,8 +172,9 @@ class Node(object):
             revert_state(self, msg['term'])
 
         reject_msg = {
-                'term': self.raft_term,
+                'type': 'AppendEntryReply',
                 'destination': [msg['source']],
+                'term': self.raft_term,
                 'success': False
         }
 
@@ -213,8 +214,22 @@ class Node(object):
 
         self.raft_log.extend(rcvd_entries)
         self.log_info('Successfully appended {0} entries to own log'.format(len(rcvd_entries)))
-        
 
+        # Take care of updating commitIndex and actually commiting the commands that we got sent.
+        if msg['leaderCommit'] > self.raft_commitIndex:
+            new_commitIndex = min(msg['leaderCommit'], len(raft_log) - 1)
+            for entry in self.raft_log[self.raft_commitIndex + 1:new_commitIndex + 1]:
+                self.commit_entry(entry)
+            self.raft_commitIndex = new_commitIndex
+        
+        # Send a message back to the leader informing of success.
+        success_msg = {
+                'type': 'AppendEntryReply',
+                'destination': [msg['source']],
+                'term': self.raft_term,
+                'success': True
+        }
+        self.req.send_json(success_msg)
 
 
     '''
@@ -258,6 +273,10 @@ class Node(object):
 
         self.raft_timeout = self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = random.randint(150, 300)), self.leader_timeout)
 
+    # The function that applies the log entry to whatever internal state we're using.
+    # This is the function that should be overriden by any derived subclasses.
+    def commit_entry(self, log_entry):
+        pass
 
     def log_info(self, s):
         self.req.send_json({'type': 'log', 'info': s})

### 2014-06-02 proj/raft.py
figured out the ioloop timeout, implemented the heartbeat

@@ -11,7 +11,7 @@ from copy import copy
 from collections import defaultdict
 ioloop.install()
 
-class Node(object):
+class RaftNode(object):
     def __init__(self, name, pub_endpoint, router_endpoint, peer_names):
         self.loop = ioloop.ZMQIOLoop.current()
         self.context = zmq.Context()
@@ -38,12 +38,14 @@ class Node(object):
                 'hello': self.helloRespond,
                 'RequestVote': self.handle_reqvote,
                 'VoteReply': self.handle_votereply,
-                'AppendEntry': self.handle_appendentry
+                'AppendEntry': self.handle_appendentry,
+                'AppendEntryReply': self.handle_appendentry_reply
         }
 
         # Node's Raft state (possibly to be augmented by DHT later?)
         self.raft_peers = copy(self.peers)
         self.raft_state = 'follower'
+        self.raft_leader = None
 
         '''
         The format of each log entry is a dict. The 'term' value is constant among all of them, and
@@ -65,10 +67,11 @@ class Node(object):
         self.raft_nextIndex = None
         self.raft_matchIndex = None
 
-    def start(self):
-        self.loop.start()
         # Initialize the election timeout
         self.raft_timeout = self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = random.randint(150, 300)), self.leader_timeout)
+
+    def start(self):
+        self.loop.start()
 
     def handle_broker_message(self, frames):
         pass
@@ -90,6 +93,7 @@ class Node(object):
     # Not strictly a handler, but it responds directly to loop events, so I'm putting it here.
     def leader_timeout(self):
         self.log_info('Server {0} hit election timeout'.format(self.name))
+        self.raft_leader = None
         self.raft_state = 'candidate'
         self.raft_term += 1
         reqvote_msg = {
@@ -103,6 +107,23 @@ class Node(object):
         self.raft_lastvote = self.name
         self.raft_numVotes = 1 # Ourselves
         self.reset_timeout()
+
+    # A periodic (20ms) heartbeat message that gets registered for leaders
+    def leader_heartbeat(self):
+        if self.raft_state != 'leader':
+            return
+
+        leader_heartbeat = {
+                'type': 'AppendEntries',
+                'destination': self.raft_peers,
+                'source': self.name,
+                'term': self.raft_term,
+                'lastLogIndex': len(self.raft_log) - 1,
+                'lastLogTerm': self.raft_log[-1]['term'],
+                'entries': [],
[…]

### 2014-06-02 proj/raft.py
modifying heartbeat

@@ -43,7 +43,8 @@ class RaftNode(object):
         }
 
         # Node's Raft state (possibly to be augmented by DHT later?)
-        self.raft_peers = copy(self.peers)
+        self.raft_peers = copy(self.peers) 
+        self.raft_majority_num = len(self.raft_peers) / 2 # include ourself
         self.raft_state = 'follower'
         self.raft_leader = None
 
@@ -108,22 +109,27 @@ class RaftNode(object):
         self.raft_numVotes = 1 # Ourselves
         self.reset_timeout()
 
-    # A periodic (20ms) heartbeat message that gets registered for leaders
+    # A periodic heartbeat message that gets registered for leaders. 
+    # The heartbeat will consult nextIndex and send along entries as well, if
+    # there are any to be sent.
     def leader_heartbeat(self):
         if self.raft_state != 'leader':
             return
-
-        leader_heartbeat = {
-                'type': 'AppendEntries',
-                'destination': self.raft_peers,
-                'source': self.name,
-                'term': self.raft_term,
-                'lastLogIndex': len(self.raft_log) - 1,
-                'lastLogTerm': self.raft_log[-1]['term'],
-                'entries': [],
-                'leaderCommit': self.raft_commitIndex
-        }
-        self.req.send_json(leader_heartbeat)
+        
+        for peer in self.raft_peers:
+            lastLogIndex = self.raft_nextIndex[peer] - 1
+            lastLogTerm = self.raft_log[lastLogIndex]['term']
+            leader_heartbeat = {
+                    'type': 'AppendEntries',
+                    'destination': [peer],
+                    'source': self.name,
+                    'term': self.raft_term,
+                    'lastLogIndex': lastLogIndex,
+                    'lastLogTerm': lastLogTerm,
+                    'entries': self.raft_log[lastLogIndex + 1:],
+                    'leaderCommit': self.raft_commitIndex
+            }
+            self.req.send_json(leader_heartbeat)
 
     def handle_reqvote(self, msg):
         # If term is less, simply reject
@@ -251,22 +257,63 @@ class RaftNode(object):
             self.raft_commitIndex = new_commitIndex
         
         # Send a message back to the leader informing of success.
-        success_msg = {
+        # Keep all the info that the leader sent over because it's needed
+        # to keep track of our state on the leader's side.
+        success_msg = msg.copy()
+        success_msg.update({
                 'type': 'AppendEntryReply',
                 'destination': [msg['source']],
                 'term': self.raft_term,
[…]

### 2014-06-02 proj/raft.py
done with base?

@@ -68,10 +68,10 @@ class RaftNode(object):
         self.raft_nextIndex = None
         self.raft_matchIndex = None
 
-        # Initialize the election timeout
-        self.raft_timeout = self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = random.randint(150, 300)), self.leader_timeout)
 
     def start(self):
+        # Initialize the election timeout
+        self.raft_timeout = self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = random.randint(150, 300)), self.leader_timeout)
         self.loop.start()
 
     def handle_broker_message(self, frames):
@@ -133,7 +133,7 @@ class RaftNode(object):
 
     def handle_reqvote(self, msg):
         # If term is less, simply reject
-        elif msg['term'] < self.raft_term:
+        if msg['term'] < self.raft_term:
             reject_msg = {
                     'type': 'VoteReply',
                     'destination': [msg['source']],

### 2014-06-02 proj/raft_base.py
name changes

@@ -11,7 +11,7 @@ from copy import copy
 from collections import defaultdict
 ioloop.install()
 
-class RaftNode(object):
+class RaftBaseNode(object):
     def __init__(self, name, pub_endpoint, router_endpoint, peer_names):
         self.loop = ioloop.ZMQIOLoop.current()
         self.context = zmq.Context()

### 2014-06-02 proj/raftNode.py
finished the basic node, now to do the client

@@ -1,0 +1,114 @@
+from raft_base import RaftBaseNode
+
+'''
+A derived class of the base node -- this just implements a k-v store on top of Raft.
+This handles the client interaction with the custom client node (in another file)
+'''
+class RaftNode(RaftBaseNode):
+    def __init__(self, name, pub_endpoint, router_endpoint, peer_names):
+        RaftBaseNode.__init__(self, name, pub_endpoint, router_endpoint, peer_names)
+        self.data_store = {}
+
+        # Add message handlers for the client-facing get/set commands. Note that
+        # these are not raw messages from the broker -- the client process will put them
+        # into a better format. Names are in caps so the broker doesn't mess with them.
+
+        self.message_handlers.update({
+            'GET': self.handle_clientget,
+            'SET': self.handle_clientset
+        })
+
+        # A dict of GET requests
+        self.getReq_info = {}
+    def handle_clientget(self, msg):
+        # If we're not the leader, send a redirect message
+        if self.raft_state != 'leader':
+            redirect_msg = {
+                    'type': 'redirect',
+                    'destination': msg['source'],
+                    'redir_target': self.raft_leader
+            }
+            self.req.send_json(redirect_msg)
+            return
+        
+        # Send out a heartbeat to make sure we're still the leader.
+        # If everything else comes back okay, reply with the value.
+        poke_msg = {
+                'type': 'AppendEntries',
+                'destination': self.raft_peers,
+                'source': self.name,
+                'term': self.raft_term,
+                'lastLogIndex': lastLogIndex,
+                'lastLogTerm': lastLogTerm,
+                'entries': self.raft_log[lastLogIndex + 1:],
+                'leaderCommit': self.raft_commitIndex,
+                'id': msg['id']
+        }
+        self.req.send_json(poke_msg)
+
+        # Register the get request in our dict of pending ones
+        self.getReq_info[msg['id']] = {
+                'source': msg['source'],
+                'key': msg['key']
+        }
+
+    def handle_clientset(self, msg):
+        if self.raft_state != 'leader':
+            redirect_msg = {
+                    'type': 'redirect',
+                    'destination': msg['source'],
+                    'redir_target': self.raft_leader
+            }
+            self.req.send_json(redirect_msg)
+            return
+        
+        # Insert the set request into our log and immediately trigger a heartbeat
+        # which conveniently contains all unsent log entries..
+        self.raft_log.append({
+            'term': self.raft_term,
+            'action': 'set',
[…]

### 2014-06-02 proj/raft_base.py

@@ -68,6 +68,9 @@ class RaftBaseNode(object):
         self.raft_nextIndex = None
         self.raft_matchIndex = None
 
+        # This is a dict of pending GET requests. We need to wait on
+        # a majority of heartbeat messages to get sent out and come back.
+        self.pending_getReq_count = defaultdict(int)
 
     def start(self):
         # Initialize the election timeout
@@ -210,6 +213,8 @@ class RaftBaseNode(object):
             self.req.send_json(reject_msg)
             return
 
+
+
         self.reset_timeout()
         self.raft_leader = msg['source']
 
@@ -280,6 +285,12 @@ class RaftBaseNode(object):
         if self.raft_state != 'leader':
             return
         
+        # Some special handling for GET requests.
+        if msg.get('id') is not None:
+            self.pending_getReq_count[msg['id']] += 1
+            if self.pending_getReq_count[msg['id']] >= self.raft_majority_num:
+                self.respond_to_getReq(msg['id'])
+
         # If the AppendEntries was unsuccessful, decrement nextIndex for the receiver
         # and try again
         if not msg['success']:
@@ -322,6 +333,7 @@ class RaftBaseNode(object):
         self.raft_state = 'follower'
         self.raft_term = term
         self.raft_lastvote = None
+        self.pending_getReq_count = {}
         try:
             self.loop.remove_timeout(self.leader_refresh)
         except:
@@ -337,20 +349,12 @@ class RaftBaseNode(object):
         self.raft_state = 'leader'
         self.raft_nextIndex = defaultdict(lambda: len(self.raft_log) + 1)
         self.raft_matchIndex = defaultdict(int)
-
-        # Send out a heartbeat AppendEntries to all peers
-        leader_heartbeat = {
-                'type': 'AppendEntries',
-                'destination': self.raft_peers,
-                'source': self.name,
-                'term': self.raft_term,
-                'lastLogIndex': len(self.raft_log) - 1,
-                'lastLogTerm': self.raft_log[-1]['term'],
-                'entries': [],
-                'leaderCommit': self.raft_commitIndex
-        }
-        self.req.send_json(leader_heartbeat)
         self.raft_leader = self.name 
+
+        # Send out a heartbeat AppendEntries to all peers, while adding a noop.
+        self.raft_log.append({'term': self.raft_term, 'action': None})
+        self.leader_heartbeat()
+        
[…]

### 2014-06-02 proj/clientNode.py
in middle of implementing client

@@ -1,0 +1,158 @@
+from Queue import Queue
+import random
+import json
+import sys
+import signal
+import datetime
+import zmq
+from zmq.eventloop import ioloop, zmqstream
+import argparse
+from copy import copy
+from collections import defaultdict
+ioloop.install()
+
+'''
+A special node not part of Raft that's designed to interface with the broker's limited
+client capabilities. It's here because Raft has unstable leaders in the case of failures, so
+we need to be able to dynamically redirect messages. All get/set messages from the broker
+go through this node. There is only one of this node.
+'''
+class ClientNode(object):
+    def __init__(self, name, pub_endpoint, router_endpoint, peer_names):
+        self.loop = ioloop.ZMQIOLoop.current()
+        self.context = zmq.Context()
+    
+        # SUB socket for receiving messages from the broker
+        self.sub_sock = self.context.socket(zmq.SUB)
+        self.sub_sock.connect(pub_endpoint)
+        # make sure we get messages meant for us!
+        self.sub_sock.set(zmq.SUBSCRIBE, name)
+        self.sub = zmqstream.ZMQStream(self.sub_sock, self.loop)
+        self.sub.on_recv(self.msg_handler)
+    
+        # REQ socket for sending messages to the broker
+        self.req_sock = self.context.socket(zmq.REQ)
+        self.req_sock.connect(router_endpoint)
+        self.req = zmqstream.ZMQStream(self.req_sock, self.loop)
+        self.req.on_recv(self.handle_broker_message)
+
+        self.name = name
+        self.peers = peer_names
+        self.hello_sent = False
+        
+        self.message_handlers = {
+                'hello': self.helloRespond,
+                'get': self.handle_get,
+                'set': self.handle_set,
+                'redirect': self.handle_redirect,
+                'set_success': self.handle_set_success,
+                'get_success': self.handle_get_success
+        }
+
+        self.max_retries = 4
+        self.msg_timeout = datetime.timedelta(milliseconds = 300)
+        # Randomly pick a leader to interface with. If we're wrong, oh well.
+        self.raft_leader = self.peers[0] 
+
+        # Record of broker's requests associated with request id. Each entry is
+        # a dict with fields type, key, val, timeout, num_retries. Assert error if 3 retries.
+        self.broker_requests = {}
+
+        # IDs for use in Raft. Incremented once every time a request is sent off (unless
+        # it's a redirect.
+        self.id_for_raft = 0
[…]

### 2014-06-02 proj/raftNode.py

@@ -26,7 +26,8 @@ class RaftNode(RaftBaseNode):
             redirect_msg = {
                     'type': 'redirect',
                     'destination': msg['source'],
-                    'redir_target': self.raft_leader
+                    'redir_target': self.raft_leader,
+                    'id': msg['id']
             }
             self.req.send_json(redirect_msg)
             return
@@ -57,7 +58,8 @@ class RaftNode(RaftBaseNode):
             redirect_msg = {
                     'type': 'redirect',
                     'destination': msg['source'],
-                    'redir_target': self.raft_leader
+                    'redir_target': self.raft_leader,
+                    'id': msg['id']
             }
             self.req.send_json(redirect_msg)
             return
@@ -101,7 +103,7 @@ class RaftNode(RaftBaseNode):
                 'type': 'get_success',
                 'destination': req_info['source'],
                 'key': req_info['key'],
-                'value': self.data_store[req_info['key']]
+                'value': self.data_store.get(req_info['key'])
         }
         self.req.send_json(response)
         del self.getReq_info[getId]

### 2014-06-03 proj/clientNode.py
client done

@@ -45,8 +45,8 @@ class ClientNode(object):
                 'get': self.handle_get,
                 'set': self.handle_set,
                 'redirect': self.handle_redirect,
-                'set_success': self.handle_set_success,
-                'get_success': self.handle_get_success
+                'set_success': self.handle_success,
+                'get_success': self.handle_success
         }
 
         self.max_retries = 4
@@ -55,7 +55,7 @@ class ClientNode(object):
         self.raft_leader = self.peers[0] 
 
         # Record of broker's requests associated with request id. Each entry is
-        # a dict with fields type, key, val, timeout, num_retries. Assert error if 3 retries.
+        # a dict with fields type, key, val, timeout, num_retries, id. 
         self.broker_requests = {}
 
         # IDs for use in Raft. Incremented once every time a request is sent off (unless
@@ -90,7 +90,8 @@ class ClientNode(object):
                 'type': 'get',
                 'key': msg['key'],
                 'timeout': timeout,
-                'num_retries': 0
+                'num_retries': 0,
+                'id': msg['id']
         }
 
         self.raft_broker_map[self.id_for_raft] = msg['id']
@@ -111,7 +112,8 @@ class ClientNode(object):
                 'key': msg['key'],
                 'value': msg['value'],
                 'timeout': timeout,
-                'num_retries': 0
+                'num_retries': 0,
+                'id': msg['id']
         }
 
         self.raft_broker_map[self.id_for_raft] = msg['id']
@@ -132,12 +134,52 @@ class ClientNode(object):
         retry_info = self.broker_requests[self.raft_broker_map[msg['id']]]
         self.retry_message(retry_info, msg['id'])
 
+    def handle_success(self, msg):
+        success_type = msg['type'][:3]
+        req_info = self.broker_requests[self.raft_broker_map[msg['id']]]
+        self.loop.remove_timeout(req_info['timeout'])
+        del self.broker_requests[req_info['id']]
+        
+        success_msg = {
+                'type': success_type + 'Response',
+                'id': req_info['id']
+        }
+        if success_type == 'get':
+            success_msg['value'] = msg['value']
+        else:
+            success_msg['value'] = req_info['value']
+
+        self.req.send_json(success_msg)
+    '''
+    Below are utility functions and things that the timer trips.
+    '''
+
     '''
     Called as a closure by the timeout utility. Retries the message if num_retries is low.
[…]

### 2014-06-03 proj/raftNode.py

@@ -25,7 +25,7 @@ class RaftNode(RaftBaseNode):
         if self.raft_state != 'leader':
             redirect_msg = {
                     'type': 'redirect',
-                    'destination': msg['source'],
+                    'destination': [msg['source']],
                     'redir_target': self.raft_leader,
                     'id': msg['id']
             }
@@ -57,7 +57,7 @@ class RaftNode(RaftBaseNode):
         if self.raft_state != 'leader':
             redirect_msg = {
                     'type': 'redirect',
-                    'destination': msg['source'],
+                    'destination': [msg['source']],
                     'redir_target': self.raft_leader,
                     'id': msg['id']
             }
@@ -88,7 +88,8 @@ class RaftNode(RaftBaseNode):
         
         success_msg = {
                 'type': 'set_success',
-                'destination': log_entry['source'],
+                'destination': [log_entry['source']],
+                'id': log_entry['id']
         }
         self.req.send_json(success_msg)
 
@@ -101,9 +102,10 @@ class RaftNode(RaftBaseNode):
         req_info = self.getReq_info[getId]
         response = {
                 'type': 'get_success',
-                'destination': req_info['source'],
+                'destination': [req_info['source']],
                 'key': req_info['key'],
-                'value': self.data_store.get(req_info['key'])
+                'value': self.data_store.get(req_info['key']),
+                'id': getId
         }
         self.req.send_json(response)
         del self.getReq_info[getId]

### 2014-06-03 proj/main.py
ready for testing. here goes nothing

@@ -1,0 +1,34 @@
+import argparse
+from raftNode import RaftNode
+from clientNode import ClientNode
+import sys
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--pub-endpoint',
+        dest='pub_endpoint', type=str,
+        default='tcp://127.0.0.1:23310')
+    parser.add_argument('--router-endpoint',
+        dest='router_endpoint', type=str,
+        default='tcp://127.0.0.1:23311')
+    parser.add_argument('--node-name',
+        dest='node_name', type=str,
+        default='test_node')
+    parser.add_argument('--peer-names',
+        dest='peer_names', type=str,
+        default='')
+    parser.add_argument('--client',
+        dest='client', type=bool,
+        default=False)
+
+    args = parser.parse_args()
+    args.peer_names = args.peer_names.split(',')
+    
+    # Modify stdout and stderr to use a log file in /tmp
+    sys.stdout = open('/tmp/{0}_log'.format(args.node_name), 'w')
+    sys.stderr = sys.stdout
+
+    if parser.client:
+        ClientNode(args.node_name, args.pub_endpoint, args.router_endpoint, args.peer_names).start()
+    else:
+        RaftNode(args.node_name, args.pub_endpoint, args.router_endpoint, args.peer_names).start()

### 2014-06-03 proj/raft_base.py

@@ -1,8 +1,5 @@
-from Queue import Queue
 import random
 import json
-import sys
-import signal
 import datetime
 import zmq
 from zmq.eventloop import ioloop, zmqstream
@@ -376,27 +373,3 @@ class RaftBaseNode(object):
 
     def log_info(self, s):
         self.req.send_json({'type': 'log', 'info': s})
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser()
-    parser.add_argument('--pub-endpoint',
-        dest='pub_endpoint', type=str,
-        default='tcp://127.0.0.1:23310')
-    parser.add_argument('--router-endpoint',
-        dest='router_endpoint', type=str,
-        default='tcp://127.0.0.1:23311')
-    parser.add_argument('--node-name',
-        dest='node_name', type=str,
-        default='test_node')
-    parser.add_argument('--peer-names',
-        dest='peer_names', type=str,
-        default='')
-    args = parser.parse_args()
-    args.peer_names = args.peer_names.split(',')
-    
-    # Modify stdout and stderr to use a log file in /tmp
-    sys.stdout = open('/tmp/{0}_log'.format(args.node_name), 'w')
-    sys.stderr = sys.stdout
-
-    Node(args.node_name, args.pub_endpoint, args.router_endpoint, args.peer_names).start()
-

### 2014-06-03 proj/raft_base.py
ready for testing. here goes nothing

@@ -3,7 +3,6 @@ import json
 import datetime
 import zmq
 from zmq.eventloop import ioloop, zmqstream
-import argparse
 from copy import copy
 from collections import defaultdict
 ioloop.install()

### 2014-06-03 proj/main.py
fixed syntax errors... kinda surprised there was only one

@@ -4,6 +4,7 @@ from clientNode import ClientNode
 import sys
 
 if __name__ == '__main__':
+
     parser = argparse.ArgumentParser()
     parser.add_argument('--pub-endpoint',
         dest='pub_endpoint', type=str,
@@ -28,7 +29,7 @@ if __name__ == '__main__':
     sys.stdout = open('/tmp/{0}_log'.format(args.node_name), 'w')
     sys.stderr = sys.stdout
 
-    if parser.client:
+    if args.client:
         ClientNode(args.node_name, args.pub_endpoint, args.router_endpoint, args.peer_names).start()
     else:
         RaftNode(args.node_name, args.pub_endpoint, args.router_endpoint, args.peer_names).start()

### 2014-06-03 proj/raft_base.py

@@ -300,7 +300,7 @@ class RaftBaseNode(object):
                     'term': self.raft_term,
                     'lastLogIndex': lastLogIndex,
                     'lastLogTerm': lastLogTerm,
-                    'entries': self.raft_log[lastLogIndex + 1:]
+                    'entries': self.raft_log[lastLogIndex + 1:],
                     'leaderCommit': self.raft_commitIndex
             }
             self.req.send_json(retry_appendentry)

### 2014-06-03 proj/clientNode.py
timeout type incompatibility

@@ -50,7 +50,7 @@ class ClientNode(object):
         }
 
         self.max_retries = 4
-        self.msg_timeout = datetime.timedelta(milliseconds = 300)
+        self.msg_timeout = 0.3
         # Randomly pick a leader to interface with. If we're wrong, oh well.
         self.raft_leader = self.peers[0] 
 
@@ -131,6 +131,13 @@ class ClientNode(object):
     def handle_redirect(self, msg):
         # Just update who we think the leader is, then resend. If too many redirects, fail.
         self.raft_leader = msg['redir_target']
+        
+        # This can be None -- if we end up on the wrong side of a network partition
+        # or if we contact a candidate.
+        # In that case, just pick a random node and try that.
+        if self.raft_leader is None: 
+            self.raft_leader = random.choice(self.peers)
+
         retry_info = self.broker_requests[self.raft_broker_map[msg['id']]]
         self.retry_message(retry_info, msg['id'])
 

### 2014-06-03 proj/raft_base.py

@@ -70,7 +70,7 @@ class RaftBaseNode(object):
 
     def start(self):
         # Initialize the election timeout
-        self.raft_timeout = self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = random.randint(150, 300)), self.leader_timeout)
+        self.raft_timeout = self.loop.add_timeout(self.loop.time() + random.randint(150, 300)/1000., self.leader_timeout)
         self.loop.start()
 
     def handle_broker_message(self, frames):
@@ -351,7 +351,7 @@ class RaftBaseNode(object):
         self.raft_log.append({'term': self.raft_term, 'action': None})
         self.leader_heartbeat()
         
-        self.leader_refresh = self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = 20), self.leader_heartbeat)
+        self.leader_refresh = self.loop.add_timeout(self.loop.time() + 0.02, self.leader_heartbeat)
         self.log_info('Election won, becoming leader')
     
     # Reset the timeout to another 150-300ms interval from now.
@@ -359,7 +359,7 @@ class RaftBaseNode(object):
         if self.raft_timeout is not None:
             self.loop.remove_timeout(self.raft_timeout)
 
-        self.raft_timeout = self.loop.add_timeout(self.loop.time() + datetime.timedelta(milliseconds = random.randint(150, 300)), self.leader_timeout)
+        self.raft_timeout = self.loop.add_timeout(self.loop.time() + random.randint(150, 300)/1000., self.leader_timeout)
 
     # The function that applies the log entry to whatever internal state we're using.
     # This is the function that should be overriden by any derived subclasses.

### 2014-06-03 proj/raft_base.py
added dummy entry and fixed key mistake

@@ -48,11 +48,11 @@ class RaftBaseNode(object):
         The format of each log entry is a dict. The 'term' value is constant among all of them, and
         all the other entries are to be used by any derived subclasses to adjust their own state.
         '''
-        self.raft_log = []
+        self.raft_log = [{'term': -1, 'action': None}] # Start with a dummy entry in front so indices aren't confused.
 
         self.raft_term = 0
         self.raft_lastvote = None
-        self.raft_commitIndex = -1
+        self.raft_commitIndex = 0
         self.raft_lastApplied = 0
 
         self.raft_timeout = None
@@ -98,6 +98,7 @@ class RaftBaseNode(object):
         self.raft_term += 1
         reqvote_msg = {
                 'type': 'RequestVote',
+                'destination': self.raft_peers,
                 'source': self.name,
                 'term': self.raft_term,
                 'lastLogIndex': len(self.raft_log) - 1,
@@ -153,7 +154,7 @@ class RaftBaseNode(object):
 
         # Decide whether to grant the vote
         hasNotVoted = self.raft_lastvote is None
-        cand_uptodate = self.raft_log[-1]['term'] <= msg['lastLogTerm'] and len(self.raft_log) - 1 <= msg['raft_lastLogIndex']
+        cand_uptodate = self.raft_log[-1]['term'] <= msg['lastLogTerm'] and len(self.raft_log) - 1 <= msg['lastLogIndex']
         if hasNotVoted and cand_uptodate:
             vote_msg = {
                     'type': 'VoteReply',

### 2014-06-03 proj/raft_base.py
other typos

@@ -34,7 +34,7 @@ class RaftBaseNode(object):
                 'hello': self.helloRespond,
                 'RequestVote': self.handle_reqvote,
                 'VoteReply': self.handle_votereply,
-                'AppendEntry': self.handle_appendentry,
+                'AppendEntries': self.handle_appendentry,
                 'AppendEntryReply': self.handle_appendentry_reply
         }
 
@@ -78,6 +78,10 @@ class RaftBaseNode(object):
 
     def msg_handler(self, frames):
         msg = json.loads(frames[2])
+        # If we haven't sent the hello yet, don't respond to anything because
+        # it fucks up the broker
+        if not self.hello_sent and msg['type'] != 'hello':
+            return
         handler_fn = self.message_handlers.get(msg['type'], self.log_info)
         handler_fn(msg)
 
@@ -113,6 +117,7 @@ class RaftBaseNode(object):
     # The heartbeat will consult nextIndex and send along entries as well, if
     # there are any to be sent.
     def leader_heartbeat(self):
+        print 'leader heartbeat from {0}'.format(self.name)
         if self.raft_state != 'leader':
             return
         
@@ -172,7 +177,7 @@ class RaftBaseNode(object):
                     'term': self.raft_term,
                     'voteGranted': False
             }
-            self.req.send_json(vote_msg)
+            self.req.send_json(reject_msg)
             log_msg = 'Rejected vote req from {0} due to hasNotVoted={1} and uptodate={2}'.format(msg['source'], hasNotVoted, cand_uptodate)
             self.log_info(log_msg)
         
@@ -189,14 +194,14 @@ class RaftBaseNode(object):
         if msg['voteGranted']:
             self.raft_numVotes += 1
             self.log_info('Vote received, {0} total'.format(self.raft_numVotes))
-            if self.raft_numVotes >= len(self.raft_peers) + 1:
+            if self.raft_numVotes >= len(self.raft_peers)/2 + 1:
                 self.become_leader()
 
     def handle_appendentry(self, msg):
         # Check terms before doing anything else. Note that if the terms are equal and self is
         # in follower mode, revert_state does absolutely nothing.
         if self.raft_term <= msg['term']:
-            revert_state(self, msg['term'])
+            self.revert_state(msg['term'])
 
         reject_msg = {
                 'type': 'AppendEntryReply',
@@ -218,7 +223,7 @@ class RaftBaseNode(object):
         # Check log for the lastLogIndex and see if it matches up with the one from the message
[…]

### 2014-06-03 proj/raft_base.py
everything appears to be working... where's the client?

@@ -136,6 +136,8 @@ class RaftBaseNode(object):
             }
             self.req.send_json(leader_heartbeat)
 
+        self.leader_refresh = self.loop.add_timeout(self.loop.time() + 0.02, self.leader_heartbeat)
+
     def handle_reqvote(self, msg):
         # If term is less, simply reject
         if msg['term'] < self.raft_term:
@@ -231,7 +233,7 @@ class RaftBaseNode(object):
         supposed_lastLogIndex = msg['lastLogIndex']
         prev_entry_term = self.raft_log[supposed_lastLogIndex]['term']
         if prev_entry_term != msg['lastLogTerm']:
-            self.log_info('Term mismatch at index {0}'.format(supposed_lastLogIndex))
+            self.log_info('Term mismatch at index {0}. Have {1}, rcvd {2}'.format(supposed_lastLogIndex, prev_entry_term, msg['lastLogTerm']))
             self.raft_log = self.raft_log[:supposed_lastLogIndex]
             self.req.send_json(reject_msg)
             return
@@ -255,12 +257,14 @@ class RaftBaseNode(object):
 
         self.raft_log.extend(rcvd_entries)
         self.log_info('Successfully appended {0} entries to own log'.format(len(rcvd_entries)))
+        self.log_info(str(rcvd_entries[0]))
 
         # Take care of updating commitIndex and actually commiting the commands that we got sent.
         if msg['leaderCommit'] > self.raft_commitIndex:
-            new_commitIndex = min(msg['leaderCommit'], len(raft_log) - 1)
+            new_commitIndex = min(msg['leaderCommit'], len(self.raft_log) - 1)
             for entry in self.raft_log[self.raft_commitIndex + 1:new_commitIndex + 1]:
                 self.commit_entry(entry)
+            self.log_info('Committed log entries {0}-{1} inclusive at term {2}'.format(self.raft_commitIndex+1, new_commitIndex, self.raft_term))
             self.raft_commitIndex = new_commitIndex
         
         # Send a message back to the leader informing of success.
@@ -296,7 +300,7 @@ class RaftBaseNode(object):
         if not msg['success']:
             self.raft_nextIndex[msg['source']] -= 1
             lastLogIndex = self.raft_nextIndex[msg['source']] - 1
-            lastLogTerm = self.raft_log[lastLogIndex]
+            lastLogTerm = self.raft_log[lastLogIndex]['term']
             retry_appendentry = {
                     'type': 'AppendEntries',
[…]

### 2014-06-03 proj/main.py
got the client working..

@@ -18,9 +18,7 @@ if __name__ == '__main__':
     parser.add_argument('--peer-names',
         dest='peer_names', type=str,
         default='')
-    parser.add_argument('--client',
-        dest='client', type=bool,
-        default=False)
+    parser.add_argument('--client', action='store_true')
 
     args = parser.parse_args()
     args.peer_names = args.peer_names.split(',')

### 2014-06-03 proj/raftNode.py
basic get/set works!

@@ -20,6 +20,7 @@ class RaftNode(RaftBaseNode):
 
         # A dict of GET requests
         self.getReq_info = {}
+
     def handle_clientget(self, msg):
         # If we're not the leader, send a redirect message
         if self.raft_state != 'leader':
@@ -34,18 +35,21 @@ class RaftNode(RaftBaseNode):
         
         # Send out a heartbeat to make sure we're still the leader.
         # If everything else comes back okay, reply with the value.
-        poke_msg = {
-                'type': 'AppendEntries',
-                'destination': self.raft_peers,
-                'source': self.name,
-                'term': self.raft_term,
-                'lastLogIndex': lastLogIndex,
-                'lastLogTerm': lastLogTerm,
-                'entries': self.raft_log[lastLogIndex + 1:],
-                'leaderCommit': self.raft_commitIndex,
-                'id': msg['id']
-        }
-        self.req.send_json(poke_msg)
+        for peer in self.raft_peers:
+            lastLogIndex = self.raft_nextIndex[peer] - 1
+            lastLogTerm = self.raft_log[lastLogIndex]['term']
+            poke_msg = {
+                    'type': 'AppendEntries',
+                    'destination': [peer],
+                    'source': self.name,
+                    'term': self.raft_term,
+                    'lastLogIndex': lastLogIndex,
+                    'lastLogTerm': lastLogTerm,
+                    'entries': self.raft_log[lastLogIndex + 1:],
+                    'leaderCommit': self.raft_commitIndex,
+                    'id': msg['id']
+            }
+            self.req.send_json(poke_msg)
 
         # Register the get request in our dict of pending ones
         self.getReq_info[msg['id']] = {
@@ -108,6 +112,7 @@ class RaftNode(RaftBaseNode):
                 'id': getId
         }
         self.req.send_json(response)
+        print 'GET successful'
         del self.getReq_info[getId]
 
     '''

### 2014-06-03 proj/raft_base.py

@@ -217,8 +217,6 @@ class RaftBaseNode(object):
             self.req.send_json(reject_msg)
             return
 
-
-
         self.reset_timeout()
         self.raft_leader = msg['source']
 
@@ -243,29 +241,27 @@ class RaftBaseNode(object):
         # mismatch, and stick the rest of the entries from the message onto the
         # end of the remnants.
         rcvd_entries = msg['entries']
-        if len(rcvd_entries) == 0:
-            return
+        if len(rcvd_entries) > 0:
+            for logIndex in range(supposed_lastLogIndex + 1, len(self.raft_log)):
+                newEntry_index = logIndex - supposed_lastLogIndex + 1
+                if newEntry_index >= len(rcvd_entries):
+                    break
+                if self.raft_log[logIndex]['term'] != rcvd_entries[newEntry_index]['term']:
+                    self.raft_log = self.raft_log[:logIndex]
+                    break
+                rcvd_entries.pop(0) # inefficient, but these usually aren't that big...
 
-        for logIndex in range(supposed_lastLogIndex + 1, len(self.raft_log)):
-            newEntry_index = logIndex - supposed_lastLogIndex + 1
-            if newEntry_index >= len(rcvd_entries):
-                break
-            if self.raft_log[logIndex]['term'] != rcvd_entries[newEntry_index]['term']:
-                self.raft_log = self.raft_log[:logIndex]
-                break
-            rcvd_entries.pop(0) # inefficient, but these usually aren't that big...
+            self.raft_log.extend(rcvd_entries)
+            self.log_info('Successfully appended {0} entries to own log'.format(len(rcvd_entries)))
+            self.log_info(str(rcvd_entries[0]))
 
-        self.raft_log.extend(rcvd_entries)
-        self.log_info('Successfully appended {0} entries to own log'.format(len(rcvd_entries)))
-        self.log_info(str(rcvd_entries[0]))
-
-        # Take care of updating commitIndex and actually commiting the commands that we got sent.
-        if msg['leaderCommit'] > self.raft_commitIndex:
-            new_commitIndex = min(msg['leaderCommit'], len(self.raft_log) - 1)
-            for entry in self.raft_log[self.raft_commitIndex + 1:new_commitIndex + 1]:
-                self.commit_entry(entry)
-            self.log_info('Committed log entries {0}-{1} inclusive at term {2}'.format(self.raft_commitIndex+1, new_commitIndex, self.raft_term))
-            self.raft_commitIndex = new_commitIndex
[…]

### 2014-06-03 proj/clientNode.py
multiple commits work now

@@ -144,9 +144,12 @@ class ClientNode(object):
     def handle_success(self, msg):
         success_type = msg['type'][:3]
         req_info = self.broker_requests[self.raft_broker_map[msg['id']]]
+        # If we've already sent one success message, don't bother doing another.
+        if req_info.get('successful'):
+            return
+
         self.loop.remove_timeout(req_info['timeout'])
-        del self.broker_requests[req_info['id']]
-        
+
         success_msg = {
                 'type': success_type + 'Response',
                 'id': req_info['id']
@@ -157,6 +160,8 @@ class ClientNode(object):
             success_msg['value'] = req_info['value']
 
         self.req.send_json(success_msg)
+        print 'Success in request {0}'.format(req_info['id'])
+        req_info['successful'] = True
     '''
     Below are utility functions and things that the timer trips.
     '''

### 2014-06-03 proj/raftNode.py

@@ -112,8 +112,6 @@ class RaftNode(RaftBaseNode):
                 'id': getId
         }
         self.req.send_json(response)
-        print 'GET successful'
-        del self.getReq_info[getId]
 
     '''
     For good measure, override revert_state so we can clear getReq_info.

### 2014-06-03 proj/clientNode.py
got leader dropout to work

@@ -138,6 +138,7 @@ class ClientNode(object):
         if self.raft_leader is None: 
             self.raft_leader = random.choice(self.peers)
 
+        self.log_info('Redirected to {0}'.format(self.raft_leader))
         retry_info = self.broker_requests[self.raft_broker_map[msg['id']]]
         self.retry_message(retry_info, msg['id'])
 
@@ -174,8 +175,10 @@ class ClientNode(object):
         if retry_info is None:
             return
 
-        self.log_info('Request id {0} timed out. Retrying ({1} of 4)'.format(brokerId, retry_info['num_retries']))
+        # Change the point-of-contact because we don't know what made it time out
+        self.raft_leader = random.choice(self.peers) 
         self.retry_message(retry_info, self.id_for_raft)
+        self.log_info('Request id {0} timed out. Retrying {1}th iteration to {2}'.format(brokerId, retry_info['num_retries'], self.raft_leader))
 
         # Update the mapping in self.raft_broker_map and set a new timeout
         self.raft_broker_map[self.id_for_raft] = brokerId

### 2014-06-03 proj/main.py

@@ -19,6 +19,7 @@ if __name__ == '__main__':
         dest='peer_names', type=str,
         default='')
     parser.add_argument('--client', action='store_true')
+    parser.add_argument('--likely_leader', action='store_true')
 
     args = parser.parse_args()
     args.peer_names = args.peer_names.split(',')
@@ -30,4 +31,4 @@ if __name__ == '__main__':
     if args.client:
         ClientNode(args.node_name, args.pub_endpoint, args.router_endpoint, args.peer_names).start()
     else:
-        RaftNode(args.node_name, args.pub_endpoint, args.router_endpoint, args.peer_names).start()
+        RaftNode(args.node_name, args.pub_endpoint, args.router_endpoint, args.peer_names, likely_leader = args.likely_leader).start()

### 2014-06-03 proj/raftNode.py

@@ -5,8 +5,8 @@ A derived class of the base node -- this just implements a k-v store on top of R
 This handles the client interaction with the custom client node (in another file)
 '''
 class RaftNode(RaftBaseNode):
-    def __init__(self, name, pub_endpoint, router_endpoint, peer_names):
-        RaftBaseNode.__init__(self, name, pub_endpoint, router_endpoint, peer_names)
+    def __init__(self, name, pub_endpoint, router_endpoint, peer_names, **kwargs):
+        RaftBaseNode.__init__(self, name, pub_endpoint, router_endpoint, peer_names, **kwargs)
         self.data_store = {}
 
         # Add message handlers for the client-facing get/set commands. Note that

### 2014-06-03 proj/raft_base.py

@@ -8,9 +8,10 @@ from collections import defaultdict
 ioloop.install()
 
 class RaftBaseNode(object):
-    def __init__(self, name, pub_endpoint, router_endpoint, peer_names):
+    def __init__(self, name, pub_endpoint, router_endpoint, peer_names, **kwargs):
         self.loop = ioloop.ZMQIOLoop.current()
         self.context = zmq.Context()
+        self.kwargs = kwargs
     
         # SUB socket for receiving messages from the broker
         self.sub_sock = self.context.socket(zmq.SUB)
@@ -94,6 +95,9 @@ class RaftBaseNode(object):
             self.hello_sent = True
             self.req.send_json({'type': 'log', 'message': 'hello received'})
 
+            if self.kwargs.get('likely_leader'):
+                self.leader_timeout() # This guarantees that this node will start the election first
+
     # Not strictly a handler, but it responds directly to loop events, so I'm putting it here.
     def leader_timeout(self):
         self.log_info('Server {0} hit election timeout'.format(self.name))
@@ -155,6 +159,7 @@ class RaftBaseNode(object):
         # going for candidacy again soon. Thus, reset the timeout.
         self.reset_timeout()
 
+        self.log_info('Our term={0}, their term={1}'.format(self.raft_term, msg['term']))
         # Check if term is larger and change own status as appropriate
         if msg['term'] > self.raft_term:
             self.revert_state(msg['term'])
@@ -334,7 +339,7 @@ class RaftBaseNode(object):
         if term > self.raft_term:
             self.raft_lastvote = None
         self.raft_term = term
-        self.pending_getReq_count = {}
+        self.pending_getReq_count = defaultdict(int)
         try:
             self.loop.remove_timeout(self.leader_refresh)
         except:

### 2014-06-03 stat246/hw7.tex
s7

@@
 \[\sum_{i=1}^d\sum_{n=1}^NE(f(X_i^{(n)},X_{N_i}^{(n)})|X_{N_i})=\sum_{i=1}^d\sum_{x_{N_i}}E(f(X_i,X_{N_i})|X_{N_i}=x_{N_i})\left|\{n:X_{N_i}=x_{N_i}\}\right|\]
-Then, trimming away the leftover counts (which are attached to the same terms on both sides) gives us the desired equality.
+Thus we have the desired equality (after the Monday correction).
 \ssn{d}

### 2014-06-04 proj/paper.tex
paper progress

\documentclass[psamsfonts]{amsart}

\usepackage{amssymb,amsfonts}
\usepackage[all,arc]{xy}
\usepackage{enumerate}
\usepackage{hyperref}

\makeatletter
\makeatother

\bibliographystyle{plain}
\title{A distributed key-value store using Raft}

\author{Hall Liu}

\begin{document}

\maketitle
\section{Summary of the protocol}
This implementation of a distributed, fault-tolerant key-value store is based on the Raft algorithm by Ongaro and Ousterhout \footnote{Diego Ongaro, John Ousterhout; \href{https://ramcloud.stanford.edu/wiki/download/attachments/11370504/raft.pdf}{In search of an Understandable Consensus Algorithm}. 2014}. The Raft algorithm is an algorithm which implements a fault-tolerant replicated state machine, where in this case the state to be replicated is the dictionary containing the keys and values that are to be stored. In this section, I will give a brief summary of the fundamentals of the algorithm.

Each node participating in a Raft cluster has some internal state which is distinct from the state of the key-value store. The former will be referred to as the \emph{Raft state}, and the latter will be referred to as the \emph{replicated state}. 
\subsection{Roles of the nodes}
Each node in a Raft cluster may take one of three roles: {\it leader}, {\it follower}, or {\it candidate}. The leader is the node responsible for coordinating communication between the nodes and client. In normal operation, the leader receives all messages coming from outside the cluster, and it is the recipient of all messages originating with other nodes in the cluster. All other nodes are followers, which keep a copy of the leader's replicated state and update it in accordance with the leader's instructions. The candidate state is invoked when communication with the leader is disrupted for some reason -- it is the transitional state that a follower goes into in order to become the new leader of the cluster.
\subsection{Terms and Leader election}
The leader of a Raft cluster is elected for the duration of a {\it term}, which is a part of the Raft state that the algorithm aims to keep consistent across all nodes. The term is a monotonically increasing variable which is included in each message that is passed from one node to another. Upon receiving any message, a node will verify that its term matches the term of the sender. If its own term is lower, it will update its term to the term of the sender, fall back to a follower state, and wait for instructions from any emerging leaders. If its own term is higher, the message will typically be disregarded, except in the case where the leader receives an outdated reply from a follower.

During normal operation, in the absence of faults or significant delays, a term will last indefinitely. However, in the event of a fault, each follower has an internal timeout which triggers after $t$ seconds without contact from the leader node, where $t$ is drawn randomly between $150$ms and $300$ms. Upon reaching this timeout, a follower will increment its term number and send a vote request to all its peers. All nodes carry who they voted for in any term, and it will grant a vote upon request if the requester is at least up-to-date as it is. Then, once a candidate receives votes from a majority of its peers (including itself -- a candidate immediately casts a vote for itself upon becoming a candidate), it will become a leader which has a log at least as up-to-date as a majority of the cluster.
\subsection{The Raft log and AppendEntries messages}
Once elected into office, the leader communicates with all other nodes in the cluster through \emph{AppendEntries} messages (these are RPCs in the original paper, but this implementation uses message-passing for communication). These messages are the heart of the algorithm -- they carry information about the leader's replicated state and also quell the followers' instinct to rebel.

The leader is the recipient of all requests from clients to modify the replicated state of the cluster. To that purpose, it encodes all such requests as a sequence of commands that may be applied to the replicated state, and stores these commands in sequence in the \emph{Raft log}. The Raft log is the mechanism by which followers keep their replicated state consistent -- if two logs agree, then the state obtained by applying the logs to the replicated state must also agree.

AppendEntries messages, then, are requests from the leader to followers to append entries to their copy of the Raft log. In order for the log to be kept consistent, each entry in the log is recorded with the term during which it was generated by the leader. Each AppendEntries message contains the index and the term of the entry immediately prior (in the leader's log) to the first entry enclosed with the message. If the follower's log contains an entry at that index with that term, it will overwrite its own log past that point with the entries contained in the message. If not, it will respond with an error message.

AppendEntries also carry information about when to commit Raft log entries to a node's copy of the replicated state. Since committed changes are assumed to be irreversible, the leader will only instruct followers to commit an entry after it knows that a majority of the followers have the entry in their log in the proper index. This way, even if faults happen, any newly elected leader must also have this entry in their log, as otherwise they would not receive votes from a majority due to being out-of-date. The commit information is carried as an integer in each AppendEntries message -- the leader sends the index of the last entry that it has committed into its replicated state.

Finally, AppendEntries messages also serve to keep the leader in power. Whenever a node receives an AppendEntries message with a term at least as high as its own, it will revert back to follower status and reset its timeout. Thus, if a leader sends out AppendEntries messages (possibly with zero actual entries attached) with greater frequency than once every $150$ms, no other nodes will attempt to become leader.
\section{Description of the implementation}
\subsection{Structure and requests}
The actual implementation is done in Python building on the provided example node as a base. There are two distinct components to this implementation -- the client node and the server node. The server node code implements the Raft state and its transitions and stores the replicated state of the cluster. The client node serves as a middleman between the cluster and the broker due to the broker's limited abilities to address nodes. A typical run with the broker scripts will start the client node

When a 
\end{document}

### 2014-06-04 proj/clientNode.py

@@ -60,7 +60,8 @@ class ClientNode(object):
 
         # IDs for use in Raft. Incremented once every time a request is sent off (unless
         # it's a redirect.
-        self.id_for_raft = 0
+        self._id_for_raft = 0
+
         # Contains a mapping of raft ids to broker ids. 
         self.raft_broker_map = {}
 
@@ -103,7 +104,7 @@ class ClientNode(object):
                 'id': self.id_for_raft
         }
         self.req.send_json(get_msg)
-        self.id_for_raft += 1
+        self._id_for_raft += 1
 
     def handle_set(self, msg):
         timeout = self.loop.add_timeout(self.loop.time() + self.msg_timeout, lambda: self.timeout_fn(msg['id']))
@@ -126,7 +127,7 @@ class ClientNode(object):
                 'id': self.id_for_raft
         }
         self.req.send_json(set_msg)
-        self.id_for_raft += 1
+        self._id_for_raft += 1
 
     def handle_redirect(self, msg):
         # Just update who we think the leader is, then resend. If too many redirects, fail.
@@ -182,7 +183,7 @@ class ClientNode(object):
 
         # Update the mapping in self.raft_broker_map and set a new timeout
         self.raft_broker_map[self.id_for_raft] = brokerId
-        self.id_for_raft += 1
+        self._id_for_raft += 1
         retry_info['timeout'] = self.loop.add_timeout(self.loop.time() + self.msg_timeout, lambda: self.timeout_fn(brokerId))
 
     # Despite best attempts, unable to get a good response back from Raft. Send an error back.
@@ -211,6 +212,13 @@ class ClientNode(object):
                 'id': retry_id
         }
         self.req.send_json(retry_msg)
+
+    '''
+    Generate unique id to send into Raft
+    '''
+    @property
+    def id_for_raft(self):
+        return self.name + str(self._id_for_raft)
 
     def log_info(self, s):
         self.req.send_json({'type': 'log', 'info': s})

### 2014-06-04 proj/paper.tex
paper progress

@@
 The client node serves as a middleman between the cluster and the broker due to the broker's limited abilities to address nodes.
-A typical run with the broker scripts will start the client node
+A typical run with the broker scripts will start the client nodes first, then the server nodes.
 
-When a 
+\subsection{GET requests}
+When the broker issues a get request, it should always be directed towards one of the clients.
+Upon receiving this request, the client node associates the info in the request to the id provided by the broker so that it can formulate a response back to the broker after the Raft cluster responds.
+It then generates a new id which is unique with respect to all other clients and requests by prepending its name to a counter incremented every time an id is generated.
+The client node will then associate this new id with the broker's id for the request, then send off a \verb|GET| message to who it thinks the Raft leader is.
+
+The client node initially assumes that the first entry in its list of Raft nodes is the leader.
+However, if it sends a \verb|GET| message to a node which is not the leader, that node will reply with a \verb|redirect| message providing who it thinks that the current leader is.
+The client will then record this as the new leader and retry the request.
+If the cluster is currently transitioning between leaders, the response to the \verb|redirect| may be null.
+In this case, the client retries with a randomly chosen node.
+
+Due to the nature of log replication, the leader will always have the most up-to-date replicated state if it is still the sole leader and has not been deposed.
+Thus, there is no need for the leader to append an entry into its log -- it can just respond using its copy of the replicated state after verifying that it has not been deposed.
+To do this, after receiving the \verb|GET| request from the client, the cluster leader sends out a possibly empty AppendEntries message tagged with the client's request id to all followers in order to solicit a response.
+It then associates the request id to client request, then exits to process other messages.
+The leader will keep track of the responses associated with that id, and once it receives a majority affirming that the term has not ended, it will reply back to the client with the appropriate data indicating success, including the client's request id that was sent with the initial request.
+When the client receives this message, it will retrieve the broker's request using the id in the reply, then formulate a reply back to the broker indicating success and carrying the value.
+
+However, if a fault occurs during this process which leaves the leader unable to contact a majority of nodes, no response may be generated.
+To handle this case, the client nodes associate a $300$ms timeout with each request, and will retry the request if this timeout expires.
+The value of the timeout was chosen to be small while still being long enough to wait out leadership changes.
+After some number of retries (I picked $5$ to allow for two cycles of timeout-redirects), the client will return an error to the broker.
+\subsection{SET requests}
+The beginning and end of a \verb|SET| request look much like those of \verb|GET| requests, wherein the broker submits a request to the client, the client repackages it for the cluster, and the response follows the same chain backwards.
+The middle part of a \verb|SET| request differs, as setting a value requires a change to be committed into the replicated state.
+To do this, the leader appends an entry into its log containing the key, the value, the request id, and the client's name.
+It then runs through the Raft consensus algorithm until the entry is committed.
+In this implementation, committing a \verb|SET| request entry on any node involves writing it to the replicated state, then messaging the client recorded in the entry to let it know that the request has been completed.
+This results in all nodes sending the same success message to the client, which improves the odds that the client will see one of them in the event of a network partition.
+However, it also means that the client has to deal with filtering out duplicates.
+It does this by using the unique request id, which is included in all messages sent from and to the client.
 \end{document}

### 2014-06-04 proj/paper.tex
more paper

@@
 However, if a fault occurs during this process which leaves the leader unable to contact a majority of nodes, no response may be generated.
-To handle this case, the client nodes associate a $300$ms timeout with each request, and will retry the request if this timeout expires.
+To handle this case, the client nodes associate a $600$ms timeout with each request, and will retry the request if this timeout expires.
 The value of the timeout was chosen to be small while still being long enough to wait out leadership changes.
@@
 It does this by using the unique request id, which is included in all messages sent from and to the client.
+
+The cluster may fail to respond to \verb|SET| requests as with \verb|GET| requests.
+However, in the case of \verb|SET|, retrying may lead to duplicate entries being committed into the replicated state.
+This is an acceptable outcome -- the last \verb|SET| entry will effectively override any earlier ones, and since the last \verb|SET| entry must be committed before the client reports success back to the broker, we can simply choose the commit of the last \verb|SET| entry to be the linearization point of the request from the broker.
+\subsection{Fault tolerance}
+Due to the majority-based voting of the Raft internals, this implementation is able to handle the failure of a minority of its nodes while being able to return to a state where it available.
+With respect to consistency, the original paper by Ongaro and Ousterhout gives a list of consistency properties in Figure 3 which have been shown to hold at all times.
+However, this only implies that this implementation is only eventually consistent.
+Consider a run where the leader has committed some key (and therefore sent notification to the client).
+If the leader fails between now and when it next contacts its followers, none of its followers will have committed the entry, so when the next leader comes to power, it will not have committed the entry either.
+If the client issues a \verb|GET| request at this point, the leader may not have determined that the entry was safe to commit before responding to the \verb|GET|, so the client may receive stale data.
+However, as the algorithm runs, the entry will eventually be committed due to the Leader Completeness property, guaranteeing eventual consistency.
+Unfortunately, since election timeouts are randomized and the broker can't perform commands conditioned on the contents of messages, I can't come up with a script that demonstrates this happening.
+
+With respect to availability, it's easy to see how Raft may become unavailable under particular circumstances.
+If luckily timed transient node failures keep hitting the leaders just after they are elected, Raft will be unavailable.
+In addition, dueling candidates during elections can also impact availability.
+However, this is a rare case, so in most circumstances with isolated single failures, availability will only be impacted for a fraction of a second as leader election takes place.
+Thus, if we view this implementation in the CAP framework, it sits somewhere in the middle.
+It guarantees eventual consistency (but probably fully consistent without the lucky timing described above), and if we view error messages as the equivalent of not completing in a finite time, it has availability in the absence of really luckily timed faults.
+\section{Script descriptions}
+First, before I start describing the scripts, there is a small caveat.
+Since the elections are randomized to some extent, there can be a pretty long setup period in the beginning where there are no leaders.
+If this happens and if the first thing in the broker script is a set, then everything later on goes awry as well.
+Thus, everything I'm describing here assumes speedy elections, which happen most of the time but not always.
+\subsection{Leader failure}
+This script is named \verb|leader_failure.chi|, and it illustrates the scenario where the leader experiences a temporary network partition that isolates it from the rest of the network.
+The desired action in this case is for the four remaining followers to elect a new leader and keep serving requests, and upon rejoining, the former leader will realize that it is out-of-date, revert to a follower, and catch up on its log.
+
+The logging output of a sample run is attached as \verb|leader_failure.out|.
+At line 76, the client first receives a \verb|SET| request from the broker and forwards it to \verb|A1|, who happens to be the leader.
+It then broadcasts AppendEntries messages, and \verb|A2| and \verb|A3| seem to have gotten them first (lines 80, 84).
+Having received affirmation from a majority, \verb|A1| goes ahead and commits (line 129, the ordering is kinda messed up), and the broker receives a response (127).
+Next, the broker issues a \verb|GET|, which \verb|A1| responds to immediately after the next heartbeat (129, 136).
+The broker then creates a partition, and $151$ms later, \verb|A3| times out.
+In the subsequent lines, \verb|A2| and \verb|A4| grant votes to \verb|A3|, while \verb|A5| fails to grant a vote due to an incompatibility in its log (probably from starting so late).
+In any case, \verb|A3| receives a majority and becomes leader (165).
+Meanwhile, the client has timed out contacting \verb|A1| for the \verb|GET|, as \verb|A1| is no longer able to receive heartbeat responses (195).
+Thus, it picks a random node to try (\verb|A2|) and gets redirected to \verb|A3| (200).
+\verb|A3| is then able to respond to the \verb|GET| request as leader, and similarly with the subsequent \verb|SET|.
+
+Later, at line 235, the network partition is resolved, and \verb|A1| is able to contact other nodes again.
+Immediately, it sends out a heartbeat and is rejected by \verb|A5| because its term is lagging (237).
+\verb|A1| then receives AppendEntries from the current leader \verb|A3|, rejects them a few times because it is too out-of-date, and finally succeeds in appending 3 entries (241, 249, 253).
+Meanwhile, \verb|A3| handles the \verb|GET| request from the client in routine fashion, without consulting with the out-of-date \verb|A1|.
 \end{document}

### 2014-06-04 proj/clientNode.py

@@ -50,7 +50,7 @@ class ClientNode(object):
         }
 
         self.max_retries = 4
-        self.msg_timeout = 0.3
+        self.msg_timeout = 0.6
         # Randomly pick a leader to interface with. If we're wrong, oh well.
         self.raft_leader = self.peers[0] 
 

### 2014-06-04 proj/raftNode.py

@@ -31,6 +31,7 @@ class RaftNode(RaftBaseNode):
                     'id': msg['id']
             }
             self.req.send_json(redirect_msg)
+            self.log_info('Redirected client to {0}'.format(self.raft_leader))
             return
         
         # Send out a heartbeat to make sure we're still the leader.
@@ -56,6 +57,7 @@ class RaftNode(RaftBaseNode):
                 'source': msg['source'],
                 'key': msg['key']
         }
+        self.log_info('Received client GET for key {0}, id {1}'.format(msg['key'], msg['id']))
 
     def handle_clientset(self, msg):
         if self.raft_state != 'leader':
@@ -66,6 +68,7 @@ class RaftNode(RaftBaseNode):
                     'id': msg['id']
             }
             self.req.send_json(redirect_msg)
+            self.log_info('Redirected client to {0}'.format(self.raft_leader))
             return
         
         # Insert the set request into our log and immediately trigger a heartbeat
@@ -78,6 +81,7 @@ class RaftNode(RaftBaseNode):
             'source': msg['source'],
             'id': msg['id']
         })
+        self.log_info('Received client SET for key {0} to value {1}, id {2}'.format(msg['key'], msg['value'], msg['id']))
         self.leader_heartbeat()
 
     def commit_entry(self, log_entry):

### 2014-06-04 proj/raft_base.py

@@ -121,7 +121,6 @@ class RaftBaseNode(object):
     # The heartbeat will consult nextIndex and send along entries as well, if
     # there are any to be sent.
     def leader_heartbeat(self):
-        print 'leader heartbeat from {0}'.format(self.name)
         if self.raft_state != 'leader':
             return
         
@@ -159,9 +158,9 @@ class RaftBaseNode(object):
         # going for candidacy again soon. Thus, reset the timeout.
         self.reset_timeout()
 
-        self.log_info('Our term={0}, their term={1}'.format(self.raft_term, msg['term']))
         # Check if term is larger and change own status as appropriate
         if msg['term'] > self.raft_term:
+            self.log_info('reqvote: Becoming follower because {0} has term {1} and we have term {2}'.format(msg['source'], msg['term'], self.raft_term))
             self.revert_state(msg['term'])
 
         # Decide whether to grant the vote
@@ -191,6 +190,7 @@ class RaftBaseNode(object):
     def handle_votereply(self, msg):
         # If term leads ours, fall back to follower and exit
         if msg['term'] > self.raft_term:
+            self.log_info('votereply: Becoming follower because someone has term {0} and we have term {1}'.format(msg['term'], self.raft_term))
             self.revert_state(msg['term'])
             return
         # If term is lagging or no longer a candidate, just ignore the message.
@@ -208,6 +208,8 @@ class RaftBaseNode(object):
         # Check terms before doing anything else. Note that if the terms are equal and self is
         # in follower mode, revert_state does absolutely nothing.
         if self.raft_term <= msg['term']:
+            if self.raft_state != 'follower':
+                self.log_info('AppendEntries: Becoming follower because {0} has term {1} and we have term {2}'.format(msg['source'], msg['term'], self.raft_term))
             self.revert_state(msg['term'])
 
         reject_msg = {
@@ -258,7 +260,6 @@ class RaftBaseNode(object):
 
             self.raft_log.extend(rcvd_entries)
             self.log_info('Successfully appended {0} entries to own log'.format(len(rcvd_entries)))
-            self.log_info(str(rcvd_entries[0]))
 
             # Take care of updating commitIndex and actually commiting the commands that we got sent.
             if msg['leaderCommit'] > self.raft_commitIndex:
@@ -284,6 +285,7 @@ class RaftBaseNode(object):
     def handle_appendentry_reply(self, msg):
         # Check terms (like always...)
[…]

### 2014-06-05 stat246/sheet.tex
stat sheet

@@
 \DeclareMathOperator{\cov}{cov}
+\nc{\h}[1]{\widehat{#1}}
 \begin{document}
@@
 Then take the expectation over the hidden variable conditioned by the previous iteration and the data, which will give an expression involving the conditional density of $Y$ and a sum.
+Write it in terms of indicator functions.
 Finally, take derivatives to get a ML estimate of this expectation and update parameters.
@@
 Probably should make it a conjugate prior.
+
+Walkthrough of EM on Bernoulli mixture: suppose there are $M$ classes, each class has Bernoulli probability $p_i$, each class with prob $\pi_i$.
+The complete distribution for $Y=m$ is $\pi_mp_m^x(1-p_m)^{1-x}$.
+Complete data log-likelihood of data points $1..n$ is $\sum_{i=1}^n\pi_{Y_i}p_{Y_i}^{X_i}(1-p_{Y_i})^{1-X_i}$.
+Writing it with indicator functions gives $\sum_{i=1}^N\sum_{j=1}^MI(Y_i=j)\pi_jp_j^{X_i}(1-p_j)^{1-X_i}$.
 
@@
 
-Hidden Markov models -- insert graph below
+Hidden Markov models: Hidden vars have $L$ states, observed ones have $R$ states.
+Let $A$ be the transition matrix for the hidden vars, and $B$ be the transition from hidden to observed (so $A_{ij}$ is $P(Y_{n+1}=j|Y_n=i)$ and same with $B$).
+Marginal on the observed variables involves summing over all possible combinations of the hidden ones, which is exponential.
+Let $\pi$ be the initial distribution of the $Y$s.
+
+Recursive formula: let $\alpha_t(l)=P(X_{1:t}=r_{1:t}, Y_t=l)$.
+Then $\alpha_1(l)=P(X_1=r_1,Y_1=l)=\pi_lB_{lr_1}$.
+We have the recursion $\alpha_{t+1}(l)=\sum_{l'=1}^LA_{l'l}B_{l'r_{t+1}}\alpha_t(l')$.
+Total computational effort is $TL^2$ for the marginal (finding $\alpha_T$ is as good as finding the marginal).
+Similar with the backwards recursion -- $\beta_t(l)=P(X_{t+1:T}=r_{t+1}:T|Y_t=l)$ with the identity $\beta_t(l)=\sum_{l'=1}^L\beta_{t+1}(l')B_{l'r_{t+1}}A_{ll'}$.
+
+MLE for fully observed is $\h{A}_{ll'}=\frac{\sum_tI(Y_t=l,Y_{t+1}=l')}{\sum_tI(Y_t=l)}$ and something similar for $\h{B}$.
+For $Y$s unobserved, replace $I(Y_t=l,Y_{t+1}=l')$ with $P(Y_t=l,Y_{t+1}=l'|\Theta,X)$ and all other indicator functions the same way.
+We have $P(Y_t=l,Y_{t+1}=l'|X)=\frac{P(Y_t=l,Y_{t+1}=l',X)}{P(X)}$.
+The denom can be computed from the forwards/backwards algorithm.
+The numerator is
+\begin{align*}
+    P(X_{1:t},Y_t=l,X_{t+1:T},Y_{t+1}=l')&=P(X_{t+1:T},Y_{t+1}=l'|X_{1:t},Y_t=l)P(X_{1:t},Y_t=l)\\
+                                         &=P(X_{t+1:T},Y_{t+1}=l'|Y_t=l)\alpha_t(l)\text{ by $d$-separation}\\
+                                         &=P(X_{t+1:T}|Y_{t+1}=l',Y_t=l)P(Y_{t+1}=l'|Y_t=l)\alpha_t(l)\\
+                                         &=P(X_{t+1:T}|Y_{t+1}=l')A_{ll'}\alpha_t(l)\text{ by $d$-separation again}\\
+                                         &=P(X_{t+1}|Y_{t+1}=l')P(X_{t+2:T}|Y_{t+1}=l')A_{ll'}\alpha_t(l)\text{ by $d$-sep, cond indep}\\
+                                         &=B_{l',r_{t+1}}\beta_{t+1}(l')A_{ll'}\alpha_t(l)\\
+\end{align*}
+To do prediction of the $Y_{1:T}$, compute the argmax of $P(Y_{1:T}|X_{1:T})$ given the already-estimated parameters.
+Write out the probability of a sequence $l_{1:T}$ and observe that it can be written as a sum $\sum_{t=1}^{T-1}\Psi_t(l_t,l_{t+1})$, then we can do dynamic programming.
+
+Distributions Markov wrt undirected graphs: let $C$ be maximal cliques of some graph.
+A distribution that's Markov has density $\frac{1}{Z}\exp\left(\sum_C\psi_C(X_C)\right)$ where $\psi_C$ is an arbitrary function and $X_C$ is the set of variables corresponding to the clique.
+Another defn is that if a set separates two other sets in the graph, then the two sets are conditionally independent given the separating set.
+Proof of one-way equivalence is:
+
+First, suppose that $C$ is empty.
+Then, the graph is disconnected, and $A$ and $B$ lie in different connected components, which means that no maximal cliques contain both elements from $A$ and elements from $B$.
+We can then factor the pdf into one part containing only variables from $A$ and not $B$ and one part only variables from $B$ and not $A$, so $A$ and $B$ are independent.
+If $C$ is nonempty, let $\conj{A}$ be the set of elements reachable from $A$ without going through $C$ (this excludes $C$ itself).
+Note that $A\subset \conj{A}$ and $B\cap\conj{A}=\emptyset$ by the assumption that $C$ separates $A$ and $B$.
+Further, every maximal clique which contains elements of $\conj{A}$ must not contain elements of $\conj{A}^c$ which are not in $C$ (this would violate the separation criterion) and conversely any maximal clique which contains elements of $\conj{A}^c$ must not contain any elements of $\conj{A}$ unless all the elements of $\conj{A}^c$ in that clique are also in $C$.
+
+Thus, we have that any maximal clique must be either a subset of $\conj{A}\cup C$ or a subset of $\conj{A}^c$, which means that the probability factors into two pieces containing variables from $\conj{A}\cup C$ and $\conj{A}^c$.
+We can further decompose $\conj{A}^c$ into $(\conj{A}^c-C)\cup C$, the former of which contains $B$.
+Thus, if we integrate over the complement of $C$ and divide to get the conditional, we get a factored distribution, one piece of which contains all elements of $A$ and no elements of $B$, and the other contains all elements of $B$ and no elements of $A$, which means that we have conditional independence.
+
+Rejection sampling: let $q$ be a normalized density, $f$ be unnormalized. Pick some $k$ such that $kq\geq f$. Sample $z$ from $q$ then sample $u$ from uniform $(0,kq(z))$. Keep $z$ if $u\leq f(z)$.
+
+Importance sampling: estimates expectations.
+Have densities $f,g$ and some function $h$.
+$\int h(x)f(x)dx=\int h(x)\frac{f(x)}{g(x)}g(x)dx=E_g\left(\frac{h(x)f(x)}{g(x)}\right)$.
+If $h=1$, this becomes $E_g\left(\frac{f(x)}{g(x)}\right)$ and the integral devolves to the normalizing constant.
+
+MCMC sampling ideas: generate a Markov chain whose stationary distro approaches that of the distro we want.
+Need irreducibility of the chain (usual defn).
+Irreducibility implies a unique stationary distro $\pi$.
+Detailed balance implies that a distro is the stationary one: $\pi_iP_{ij}=\pi_jP_{ji}$.
+Analog for continuous state is some $K(x_1,x_2)=P(X_{i+1}=x_2|X_i=x_1)$ and $P(X_i=x,X_{i+1}\in B)=\int_B K(x,y)dy$.
+Strict positivity of $K$ implies irreducibility.
+Detailed balance is $\pi(x)K(x,y)=\pi(y)K(y,x)$.
+
+Metropolis-Hastings: let $q_x(y)$ be some proposal density.
+Begin at state $x$.
+Draw $y$ from $q$, let $r(x,y)=\min\left(\frac{f(y)q_y(x)}{f(x)q_x(y)},1\right)$.
+Accept $y$ with probability $r(x,y)$.
+This defines a MC that satisfies detailed balance for the normalized $f$.
+Let $\pi$ be the normalized version of $f$.
+Suppose $f(y)q_y(x)<f(x)q_x(y)$.
+Then prob that $x\to y$ is $r(x,y)q_x(y)=\frac{\pi(yq_y(x)}{\pi(x)}$.
+This is $K(x,y)$, so $f(x)K(x,y)=f(y)q_y(x)$.
+Going through with the other side we have $f(y)K(y,x)=f(y)q_y(x)$ too, so done.
+
+Gibbs sampling is a particular case of Metropolis-Hastings where we sample coordinate-wise $x_j|x_{-j}$.
+It's a special case of M-H where we use $q_x(y)=0$ if $x$ and $y$ differ in more than 1 coord and otherwise $q_x(y)=p(y_k|x_{-k})$ where $k$ is the coord where they differ.
+Note $r(x,y)=\frac{p(y)p(x_k|y_{-k})}{p(x)p(y_k|x_{-k})}$.
+Note we have $p(x)=p(x_k,x_{-k})$ and same for $y$, so numerator is $p(y_k|y_{-k})p(y_{-k})p(x_k|y_{-k})$ and denominator is $p(x_k|x_{-k})p(x_{-k})p(y_k|x_{-k})$, and these end up equal because $x_{-k}=y_{-k}$.
 \end{document}

### 2014-06-05 proj/paper.tex
paper section 3

@@
 To handle this case, the client nodes associate a $600$ms timeout with each request, and will retry the request if this timeout expires.
+If the timeout expires, the client will pick a node at random to contact, because the assumption is that the timeout is due to the leader having gone down in some fashion.
 The value of the timeout was chosen to be small while still being long enough to wait out leadership changes.
@@
 Meanwhile, \verb|A3| handles the \verb|GET| request from the client in routine fashion, without consulting with the out-of-date \verb|A1|.
+
+This script demonstrates the failstop tolerance of the data store, as well as its ability to handle rejoining nodes gracefully.
+Since a network partition of one node is equivalent to the failstop of that node from the perspective of the rest of the system, it produces the same effect as halting the other node and starting it again.
+\subsection{Partitions}
+This script is named \verb|leader_partition.chi|, and it runs the scenario where the data store gets partitioned into a group of 3 and a group of 2, where the leader ends up in the partition with two nodes.
+The desired action by the cluster is for the group of three nodes to recognize that they form a majority, elect a leader among them, and keep going, while the group with two nodes remains ineffective.
+After the partition is resolved, the two nodes in the smaller group should recognize that they are out-of-date and revert to being followers while updating themselves.
+
+The logging output of a sample run is attached as \verb|partition.out|.
+The beginning goes much as the previous script, with \verb|A1| being elected leader and processing a \verb|SET| and a \verb|GET| request.
+At line 105, the network partition is created, and $180$ms later, \verb|A2| times out and begins elections in the group of three.
+It wins the election at line 125 and becomes the leader of the group of three.
+Meanwhile, the client has submitted the next \verb|GET| request to \verb|A1| (101), but since it is unable to contact a majority to verify that it is still leader, it is unable to respond.
+Thus, the client times out, tries to contact \verb|A5| next, and is redirected to \verb|A2| (139, 144).
+\verb|A2| responds to this immediately (149), and the client proceeds with a \verb|SET| of \verb|k1| to $30$, which runs normally and succeeds (151, 174).
+
+Finally, the partition is resolved at line 175, and the client simultaneously issues a \verb|SET| (177).
+Next, we see \verb|A1| and \verb|A3| realize that they are behind on terms (181, 185, 189) and revert to being followers of \verb|A2|.
+At the same time, \verb|A2| handles the \verb|SET| request from the client and distributes the new value of \verb|k2| to all followers, including \verb|A1| and \verb|A3|, and they both catch up at log index $4$.
+The \verb|SET| proceeds, and after \verb|A1| and \verb|A3| are caught up, the cluster reverts to normal operation.
+
+This script demonstrates the partition tolerance of the data store.
+More specifically, it retains consistency while experiencing a network partition, while suffering brief loss of availability during the leader changeover (with potentially unbounded duration of the changeover process).
+\section{Issues and lessons learned}
 \end{document}

### 2014-06-05 proj/paper.tex
paper done

@@
 \section{Issues and lessons learned}
+\subsection{Implementation}
+The biggest issue that I had here was figuring out exactly how the event-driven model of the Tornado IOLoop works and how to integrate timeouts with it.
+Most networked programs I've written before were threaded and operated on the socket level, which means that I had control over how timeouts and message-driven events interact.
+The main concern I had was race conditions caused by timeouts triggering in the middle of a function execution.
+Fortunately, as a read over the Tornado source code revealed, the triggered timeouts were simply placed on a queue for later execution, so it turned out to be a non-issue.
+
+In addition, the design of the broker did not allow for very effective debugging.
+Since all the output got redirected to \verb|/dev/null|, I was unable to track down Python exceptions that got thrown when I messed up with indexing.
+I eventually resolved this problem by telling the interpreter to use a log file in \verb|/tmp| as stdout and stderr in the code for initializing the nodes.
+Beyond these two issues, the implementation process was fairly straightforward.
+Python and ZMQ make it very easy to abstract away the gritty details of a network, and that reduced the workload considerably compared to if I were doing this in C.
+\subsection{Design}
+The main issue in design was the limited ability of the broker to accept and send requests to arbitrary nodes.
+The design of Raft calls for a centralized leader, but who this leader is is completely arbitrary and left up to chance.
+I was able to essentially force a node to be the leader by starting it first then making it time out very early, but if I wanted that node to fail, I'd have no idea who the next leader was, so I'd be unable to hard-code that into the broker.
+The solution, of course, was to write a client node to sit between the broker and the cluster to handle all the redirects, and this seems to replicate real applications in that an extra layer of abstraction between the actual user and your server can be a very useful thing.
+
+Initially, when I was considering ideas for this project, I decided upon a Scatter-style\footnote{Glendenning et al., \href{http://homes.cs.washington.edu/~arvind/papers/scatter.pdf}{Scalable consistency in Scatter}.
+2011} combination of Raft and Chord, where the DHT overlay would use the Chord topology, and each ``node'' in the DHT would be a Raft cluster of 5 nodes.
+This turned out to be unfeasible due to time constraints, but I did learn a few things from thinking about how to integrate the two.
+First, Scatter's approach to getting the Raft/Paxos clusters to agree turns out to be widely applicable -- with the Chord overlay, the only modification needed is to use 2PC to make the changes consistent with neighboring clusters, then broadcast the change to everyone in the finger table so it can propagate around the ring faster.
+In addition, the issue of making sure the changes maintain consistency in the face of failures is simplified by the fact that replication is taken care of within the Raft/Paxos clusters, so different DHT nodes with copies of the same key won't need to worry about using the DHT protocols to keep themselves consistent.
+Given what I've implemented already, this probably won't be too much additional work -- making a new subclass of \verb|RaftBaseNode| with the rather simple Chord DHT protocols built in would be sufficient.
 \end{document}
